### 2023-01-22 template.qmd
basic things

---
title: Juliacon-proceedings Template
format:
  juliacon-proceedings-pdf:
    keep-tex: true  
  juliacon-proceedings-html: default
author:
  - name: Sarah Malloc
    affiliations:
      - name: An Organization
        department: The Department
        address: 1 Main St
        city: Boston
        country: USA
        postal-code: 02210-1022
      - A second affilication
    orcid: 0000-0000-0000-0000
    email: sm@example.org
    url: https://example.org/
  - name: Eliza Dealloc
    affiliations:
      - Another Affiliation
abstract: |
  This document is a template demonstrating the Juliacon-proceedings format.
keywords: [template, demo]
bibliography: bibliography.bib  
---

## Introduction {#sec-intro}

*TODO* Create a template that demonstrates the appearance, formatting, layout, and functionality of your format. Learn more about journal formats at <https://quarto.org/docs/journals/>.

### 2023-01-22 template.tex
basic things

% JuliaCon proceedings template
\documentclass{juliacon}
\begin{document}
\hypertarget{sec-intro}{%
\subsection{Introduction}\label{sec-intro}}

\emph{TODO} Create a template that demonstrates the appearance,
formatting, layout, and functionality of your format. Learn more about
journal formats at \url{https://quarto.org/docs/journals/}.
\end{document}

### 2023-01-23 template.qmd
sorted out the cls issue

@@
 ---
-
-## Introduction {#sec-intro}
+# Introduction {#sec-intro}
 
 *TODO* Create a template that demonstrates the appearance, formatting, layout, and functionality of your format. Learn more about journal formats at <https://quarto.org/docs/journals/>.
+
+## Proof-of-Concept
+
+This is a proof-of-concept for how we could use Quarto for JuliaCon proceedings.
+For current submissions, please ignore this repo and follow the official instructions [here](https://github.com/JuliaCon/JuliaConSubmission.jl)

### 2023-11-14 template.tex
sorted out the cls issue
header is now being inserte

@@
 \documentclass{juliacon}
+\setcounter{page}{1}
+% **************GENERATED FILE, DO NOT EDIT**************
+
+\def\@journalName{Proceedings of JuliaCon}
+\def\@volume{1}
+\def\@issue{1}
+\def\@year{2022}% **************GENERATED FILE, DO NOT EDIT**************
+
+\title{My JuliaCon proceeding}
+
+\author[1]{1st author}
+\author[1, 2]{2nd author}
+\author[2]{3rd author}
+\affil[1]{University}
+\affil[2]{National Lab}
+
+\keywords{Julia, Optimization, Game theory, Compiler}
+
+\hypersetup{
+pdftitle = {My JuliaCon proceeding},
+pdfsubject = {JuliaCon 2022 Proceedings},
+pdfauthor = {1st author, 2nd author, 3rd author},
+pdfkeywords = {Julia, Optimization, Game theory, Compiler},
+}
 \begin{document}
-\hypertarget{sec-intro}{%
-\subsection{Introduction}\label{sec-intro}}
+\maketitle
+\section{Introduction}\label{sec-intro}
 
@@
 journal formats at \url{https://quarto.org/docs/journals/}.
+
+\subsection{Proof-of-Concept}\label{proof-of-concept}
+
+This is a proof-of-concept for how we could use Quarto for JuliaCon
+proceedings. For current submissions, please ignore this repo and follow
+the official instructions
+\href{https://github.com/JuliaCon/JuliaConSubmission.jl}{here}
 \end{document}

### 2023-11-14 template.tex
uh

--- 2023-01-22 template.tex
@@
-% JuliaCon proceedings template
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+
 \documentclass{juliacon}
+
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\usepackage{orcidlink}
+\definecolor{mypink}{RGB}{219, 48, 122}
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Juliacon-proceedings Template},
+  pdfauthor={Sarah Malloc; Eliza Dealloc},
+  pdfkeywords={template, demo},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
+
+\title{Juliacon-proceedings Template}
+\author{
+Sarah Malloc~\orcidlink{0000-0000-0000-0000}\\An
+Organization\\\href{mailto:sm@example.org}{sm@example.org}\and 
+Eliza Dealloc\\Another Affiliation\\}
+\date{}
 \begin{document}
-\hypertarget{sec-intro}{%
-\subsection{Introduction}\label{sec-intro}}
+\maketitle
+\begin{abstract}
+This document is a template demonstrating the Juliacon-proceedings
+format.
+\end{abstract}
+\renewcommand*\contentsname{Table of contents}
+{
+\hypersetup{linkcolor=}
+\setcounter{tocdepth}{3}
+\tableofcontents
+}
+\section{Introduction}\label{sec-intro}
 
@@
 journal formats at \url{https://quarto.org/docs/journals/}.
+
+\subsection{Proof-of-Concept}\label{proof-of-concept}
+
+This is a proof-of-concept for how we could use Quarto for JuliaCon
+proceedings. For current submissions, please ignore this repo and follow
+the official instructions
+\href{https://github.com/JuliaCon/JuliaConSubmission.jl}{here}
+
 \end{document}

### 2023-11-14 template.qmd
damn this
slowly slowly

--- 2023-01-22 template.qmd
@@
 author:
-  - name: Sarah Malloc
+  - name: Patrick Altmeyer
     affiliations:
-      - name: An Organization
-        department: The Department
-        address: 1 Main St
-        city: Boston
-        country: USA
-        postal-code: 02210-1022
-      - A second affilication
+      - Delft University of Technology
     orcid: 0000-0000-0000-0000
-    email: sm@example.org
+    email: p.altmeyer@tudelft.nl
     url: https://example.org/
-  - name: Eliza Dealloc
-    affiliations:
-      - Another Affiliation
 abstract: |
@@
 
-## Introduction {#sec-intro}
+# Introduction {#sec-intro}
 
 *TODO* Create a template that demonstrates the appearance, formatting, layout, and functionality of your format. Learn more about journal formats at <https://quarto.org/docs/journals/>.
+
+## Proof-of-Concept
+
+This is a proof-of-concept for how we could use Quarto for JuliaCon proceedings.
+For current submissions, please ignore this repo and follow the official instructions [here](https://github.com/JuliaCon/JuliaConSubmission.jl).
+
+## Code
+
+```{.julia}
+println("Hello, world!")
+```

### 2023-11-14 template.tex
slowly slowly
damn this

@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 \fi
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{241,243,245}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
+\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
   pdftitle={Juliacon-proceedings Template},
-  pdfauthor={Sarah Malloc; Eliza Dealloc},
+  pdfauthor={Patrick Altmeyer},
   pdfkeywords={template, demo},
@@
 \author{
-Sarah Malloc~\orcidlink{0000-0000-0000-0000}\\An
-Organization\\\href{mailto:sm@example.org}{sm@example.org}\and 
-Eliza Dealloc\\Another Affiliation\\}
+Patrick Altmeyer~\orcidlink{0000-0000-0000-0000}\\Delft University of
+Technology\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}}
 \date{}
@@
 \maketitle
+
+% Abstract
 \begin{abstract}
@@
 \end{abstract}
-\renewcommand*\contentsname{Table of contents}
-{
-\hypersetup{linkcolor=}
-\setcounter{tocdepth}{3}
-\tableofcontents
-}
+
+% Keywords
+\begin{keywords}
+template, demo
+\end{keywords}
 \section{Introduction}\label{sec-intro}
@@
 the official instructions
-\href{https://github.com/JuliaCon/JuliaConSubmission.jl}{here}
+\href{https://github.com/JuliaCon/JuliaConSubmission.jl}{here}.
+
+\subsection{Code}\label{code}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{println}\NormalTok{(}\StringTok{"Hello, world!"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 

### 2023-11-14 template.qmd
getting there

@@
 ---
-title: Juliacon-proceedings Template
+title: Juliacon Proceedings in Quarto
 format:
@@
       - Delft University of Technology
-    orcid: 0000-0000-0000-0000
+    orcid: 0000-0003-4726-8613
     email: p.altmeyer@tudelft.nl
-    url: https://example.org/
+    url: https://www.paltmeyer.com/
 abstract: |
-  This document is a template demonstrating the Juliacon-proceedings format.
-keywords: [template, demo]
+  This document is a Quarto template demonstrating the `juliacon-proceedings` format.
+year: 2022
+keywords: [Template, Demo, Quarto, JuliaCon]
 bibliography: bibliography.bib  

### 2023-11-14 template.tex
getting there

@@
 \hypersetup{
-  pdftitle={Juliacon-proceedings Template},
+  pdftitle={Juliacon Proceedings in Quarto},
   pdfauthor={Patrick Altmeyer},
-  pdfkeywords={template, demo},
+  pdfkeywords={Template, Demo, Quarto, JuliaCon},
   colorlinks=true,
@@
 
-\title{Juliacon-proceedings Template}
+\title{Juliacon Proceedings in Quarto}
 \author{
-Patrick Altmeyer~\orcidlink{0000-0000-0000-0000}\\Delft University of
+Patrick Altmeyer~\orcidlink{0000-0003-4726-8613}\\Delft University of
 Technology\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}}
@@
 \begin{abstract}
-This document is a template demonstrating the Juliacon-proceedings
-format.
+This document is a Quarto template demonstrating the
+\texttt{juliacon-proceedings} format.
 \end{abstract}
@@
 % Keywords
-\begin{keywords}
-template, demo
-\end{keywords}
+\JCONkeywords{Template, Demo, Quarto, JuliaCon}
+
+% Hypersetup
+\hypersetup{
+    pdftitle = {Juliacon Proceedings in Quarto},
+    pdfsubject = {JuliaCon 2022 Proceedings},
+    pdfauthor = {Patrick Altmeyer},
+    pdfkeywords = {Template, Demo, Quarto, JuliaCon},
+}
+
+\setcounter{page}{1}
+
 \section{Introduction}\label{sec-intro}

### 2023-11-14 template.qmd
basically done

@@
     url: https://www.paltmeyer.com/
+  - name: Cynthia C. S. Liem
+    affiliations:
+      - Delft University of Technology
 abstract: |

### 2023-11-14 template.tex
basically done

@@
   pdftitle={Juliacon Proceedings in Quarto},
-  pdfauthor={Patrick Altmeyer},
+  pdfauthor={Patrick Altmeyer; Cynthia C. S. Liem},
   pdfkeywords={Template, Demo, Quarto, JuliaCon},
@@
 \title{Juliacon Proceedings in Quarto}
-\author{
-Patrick Altmeyer~\orcidlink{0000-0003-4726-8613}\\Delft University of
-Technology\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}}
+
+\author[1]{Patrick Altmeyer}
+\author[1]{Cynthia C. S. Liem}
+\affil[1]{Delft University of Technology}
 \date{}
@@
     pdfsubject = {JuliaCon 2022 Proceedings},
-    pdfauthor = {Patrick Altmeyer},
+    pdfauthor = {Patrick Altmeyer, Cynthia C. S. Liem},
     pdfkeywords = {Template, Demo, Quarto, JuliaCon},

### 2023-11-14 template.qmd
code: fall-back works as always, but would be nice to have support for executable code chunks

@@
 
-## Code
+## Executable Code
 
-```{.julia}
-println("Hello, world!")
+Ideally, we would be able to tab into Quarto's existing support for executable, cross-referenceable code chunks.
+
+> To create cross-referenceable code listings from executable code blocks, use `lst-label` and `lst-cap`. 
+> --- [Quarto Docs](https://quarto.org/docs/prerelease/1.4/crossref.html#cross-referenceable-listings-of-executable-code-blocks)
+
+For example, the following code block will be labeled as `lst-1` and calling `@lst-1` will render as @lst-1.
+This ensures that things stay up-to-date and potential errors in the code are identified early: as you render your document, the code will be executed and the output will be inserted into the document.
+
+```{julia}
+#| lst-label: lst-1
+#| lst-cap: A listing caption
+#| eval: false
+#| echo: true
+
+using Plots
+
+x = -3.0:0.01:3.0
+y = rand(length(x))
+plot(x, y)
 ```
+
+Unfortunately, I don't know how to use the special environment that is already defined for Julia in this context.
+
+## Code Listings
+
+The special environment that is already defined for Julia code can still be used as before.
+
+\begin{verbatim}
+\begin{lstlisting}[
+    language = Julia, 
+    numbers=left, 
+    label={lst:exmplg}, 
+    caption={Example Code Block.}
+]
+using Plots
+
+x = -3.0:0.01:3.0
+y = rand(length(x))
+plot(x, y)
+\end{lstlisting}
+\end{verbatim}
+\begin{lstlisting}[
+    language = Julia, 
+    numbers=left, 
+    label={lst:exmplg}, 
+    caption={Example Code Block.}
+]
+using Plots
+
+x = -3.0:0.01:3.0
+y = rand(length(x))
+plot(x, y)
+\end{lstlisting}

### 2023-11-14 template.tex
code: fall-back works as always, but would be nice to have support for executable code chunks

@@
 \fi
-
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{241,243,245}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
-\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 
-\subsection{Code}\label{code}
+\subsection{Executable Code}\label{executable-code}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{println}\NormalTok{(}\StringTok{"Hello, world!"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+Ideally, we would be able to tab into Quarto's existing support for
+executable, cross-referenceable code chunks.
+
+\begin{quote}
+To create cross-referenceable code listings from executable code blocks,
+use \texttt{lst-label} and \texttt{lst-cap}. ---
+\href{https://quarto.org/docs/prerelease/1.4/crossref.html\#cross-referenceable-listings-of-executable-code-blocks}{Quarto
+Docs}
+\end{quote}
+
+For example, the following code block will be labeled as \texttt{lst-1}
+and calling \texttt{@lst-1} will render as Listing~\ref{lst-1}. This
+ensures that things stay up-to-date and potential errors in the code are
+identified early: as you render your document, the code will be executed
+and the output will be inserted into the document.
+
+\begin{codelisting}
+
+\caption{\label{lst-1}A listing caption}
+
+\centering{
+
+\begin{verbatim}
+using Plots
+
+x = -3.0:0.01:3.0
+y = rand(length(x))
+plot(x, y)
+\end{verbatim}
+
+}
+
+\end{codelisting}%
+
+Unfortunately, I don't know how to use the special environment that is
+already defined for Julia in this context.
+
+\subsection{Code Listings}\label{code-listings}
+
+The special environment that is already defined for Julia code can still
+be used as before.
+
+\begin{verbatim}
+\begin{lstlisting}[
+    language = Julia, 
+    numbers=left, 
+    label={lst:exmplg}, 
+    caption={Example Code Block.}
+]
+using Plots
+
+x = -3.0:0.01:3.0
+y = rand(length(x))
+plot(x, y)
+\end{lstlisting}
+\end{verbatim}
+\begin{lstlisting}[
+    language = Julia, 
+    numbers=left, 
+    label={lst:exmplg}, 
+    caption={Example Code Block.}
+]
+using Plots
+
+x = -3.0:0.01:3.0
+y = rand(length(x))
+plot(x, y)
+\end{lstlisting}
 

### 2023-11-14 template.qmd

@@
     url: https://www.paltmeyer.com/
-  - name: Cynthia C. S. Liem
-    affiliations:
-      - Delft University of Technology
 abstract: |
   This document is a Quarto template demonstrating the `juliacon-proceedings` format.
-year: 2022
+year: 2023
 keywords: [Template, Demo, Quarto, JuliaCon]
@@
 
-*TODO* Create a template that demonstrates the appearance, formatting, layout, and functionality of your format. Learn more about journal formats at <https://quarto.org/docs/journals/>.
-
-## Proof-of-Concept
-
+This is a template for writing a JuliaCon Proceedings article in Quarto.
 This is a proof-of-concept for how we could use Quarto for JuliaCon proceedings.
@@
 
-## Executable Code
+## What is Quarto?
+
+[Quarto](https://quarto.org/) makes it easy to write reproducible documents that can be rendered to PDF, HTML, Word and more.
+It is based on Markdown, which is easy to learn and write.
+It also supports LaTeX, which is useful for more advanced formatting.
+As this extension demonstrates, Quarto is also very flexible and can be extended with custom templates and styles.
+
+## Why Quarto?
+
+By embracing Quarto, JuliaCon Proceedings can set an example for how to write reproducible documents.
+We would not only make it easier for authors to write their submissions but also open the door for more advanced features such as interactive figures and executable code blocks in HTML documents.
+
+## About this template
+
+This template is based on the existing [JuliaCon Proceedings LaTeX template](https://github.com/JuliaCon/JuliaConSubmission.jl).
+
+## How to use this template
+
+> The remainder of this document repeats relevant elements from the [JuliaCon Proceedings LaTeX template].
+
+# The JuliaCon Article Class {#sec-documentclass}
+
+The `juliacon` class file preserves the standard LATEX{} interface such that any document that can be produced using the standard LATEX{} article class can also be produced with the class file.
+
+It is likely that the make up will change after file submission.
+For this reason, we ask you to ignore details such as slightly long lines, page stretching, or figures falling out of synchronization, as these details can be dealt with at a later stage.
+
+Use should be made of symbolic references (\verb|\ref| ) in order to protect against late changes of order, etc.
+
+# Executable Code
 
@@
 
-## Code Listings
+# Code Listings
 

### 2023-11-14 template.tex

@@
   pdftitle={Juliacon Proceedings in Quarto},
-  pdfauthor={Patrick Altmeyer; Cynthia C. S. Liem},
+  pdfauthor={Patrick Altmeyer},
   pdfkeywords={Template, Demo, Quarto, JuliaCon},
@@
 \author[1]{Patrick Altmeyer}
-\author[1]{Cynthia C. S. Liem}
 \affil[1]{Delft University of Technology}
@@
     pdftitle = {Juliacon Proceedings in Quarto},
-    pdfsubject = {JuliaCon 2022 Proceedings},
-    pdfauthor = {Patrick Altmeyer, Cynthia C. S. Liem},
+    pdfsubject = {JuliaCon 2023 Proceedings},
+    pdfauthor = {Patrick Altmeyer},
     pdfkeywords = {Template, Demo, Quarto, JuliaCon},
@@
 
-\emph{TODO} Create a template that demonstrates the appearance,
-formatting, layout, and functionality of your format. Learn more about
-journal formats at \url{https://quarto.org/docs/journals/}.
-
-\subsection{Proof-of-Concept}\label{proof-of-concept}
-
+This is a template for writing a JuliaCon Proceedings article in Quarto.
 This is a proof-of-concept for how we could use Quarto for JuliaCon
@@
 
-\subsection{Executable Code}\label{executable-code}
+\subsection{What is Quarto?}\label{what-is-quarto}
+
+\href{https://quarto.org/}{Quarto} makes it easy to write reproducible
+documents that can be rendered to PDF, HTML, Word and more. It is based
+on Markdown, which is easy to learn and write. It also supports LaTeX,
+which is useful for more advanced formatting. As this extension
+demonstrates, Quarto is also very flexible and can be extended with
+custom templates and styles.
+
+\subsection{Why Quarto?}\label{why-quarto}
+
+By embracing Quarto, JuliaCon Proceedings can set an example for how to
+write reproducible documents. We would not only make it easier for
+authors to write their submissions but also open the door for more
+advanced features such as interactive figures and executable code blocks
+in HTML documents.
+
+\subsection{About this template}\label{about-this-template}
+
+This template is based on the existing
+\href{https://github.com/JuliaCon/JuliaConSubmission.jl}{JuliaCon
+Proceedings LaTeX template}.
+
+\subsection{How to use this template}\label{how-to-use-this-template}
+
+\begin{quote}
+The remainder of this document repeats relevant elements from the
+{[}JuliaCon Proceedings LaTeX template{]}.
+\end{quote}
+
+\section{The JuliaCon Article Class}\label{sec-documentclass}
+
+The \texttt{juliacon} class file preserves the standard LATEX\{\}
+interface such that any document that can be produced using the standard
+LATEX\{\} article class can also be produced with the class file.
+
+It is likely that the make up will change after file submission. For
+this reason, we ask you to ignore details such as slightly long lines,
+page stretching, or figures falling out of synchronization, as these
+details can be dealt with at a later stage.
+
+Use should be made of symbolic references (\verb|\ref| ) in order to
+protect against late changes of order, etc.
+
+\section{Executable Code}\label{executable-code}
 
@@
 
-\subsection{Code Listings}\label{code-listings}
+\section{Code Listings}\label{code-listings}
 

### 2023-11-15 template.tex
got rid of number at the end of abstract
uf

@@
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 \begin{abstract}
-This document is a Quarto template demonstrating the
-\texttt{juliacon-proceedings} format.
+
+This document is a template demonstrating the
+\texttt{juliacon-proceedings} format. It is a standard Quarto document
+composed of Markdown, LaTeX and code chunks. It is rendered to PDF and
+HTML. The PDF is rendered using the \texttt{juliacon-proceedings-pdf}
+format, which is based on the \texttt{juliacon-proceedings} LaTeX class.
+The HTML is rendered using the \texttt{juliacon-proceedings-html}
+format, which is based on the \texttt{juliacon-proceedings} HTML
+template.
 \end{abstract}
@@
     pdftitle = {Juliacon Proceedings in Quarto},
-    pdfsubject = {JuliaCon 2023 Proceedings},
+    pdfsubject = {JuliaCon \@year Proceedings},
     pdfauthor = {Patrick Altmeyer},
@@
 
+\setstretch{1}
 \section{Introduction}\label{sec-intro}
@@
 x = -3.0:0.01:3.0
-y = rand(length(x))
-plot(x, y)
+y = rand(length(x));
 \end{verbatim}

### 2023-11-15 template.qmd

@@
 abstract: |
-  This document is a Quarto template demonstrating the `juliacon-proceedings` format.
-year: 2023
+  This document is a template demonstrating the `juliacon-proceedings` format.
+It is a standard Quarto document composed of Markdown, LaTeX and code chunks.
+It is rendered to PDF and HTML.
+The PDF is rendered using the `juliacon-proceedings-pdf` format, which is based on the `juliacon-proceedings` LaTeX class.
+The HTML is rendered using the `juliacon-proceedings-html` format, which is based on the `juliacon-proceedings` HTML template.
 keywords: [Template, Demo, Quarto, JuliaCon]
 bibliography: bibliography.bib  
+jupyter: julia-1.9
 ---
+
+```{julia}
+#| echo: false
+#| output: false
+
+using Pkg; Pkg.activate(".")
+```
 
@@
 #| lst-cap: A listing caption
-#| eval: false
 #| echo: true
@@
 x = -3.0:0.01:3.0
-y = rand(length(x))
-plot(x, y)
+y = rand(length(x));
 ```

### 2023-11-15 examples/CounterfactualExplanations.jl/paper.qmd
uf

---
format:
  juliacon-proceedings-pdf: default
bibliography: ref.bib
execute:
  eval: false
  echo: false
  output: false
jupyter: julia-1.9
---

```{julia}
using Pkg; Pkg.activate("paper")
using CounterfactualExplanations
using CounterfactualExplanations.Data
using CounterfactualExplanations.DataPreprocessing: unpack_data
using CounterfactualExplanations.Evaluation
using CounterfactualExplanations.Generators
using CounterfactualExplanations.Models
using CounterfactualExplanations.Objectives
using Chain: @chain
using CSV
using Flux
using GraphRecipes
using Images
using MLDatasets
using MLDatasets: convert2image
using MLUtils
using Plots
using Random
using Serialization
using Tidier
Random.seed!(42)
Plots.theme(:wong)
default(size=(500, 375))
www_path = "paper/www"
output_path = "paper/output"
```

# Introduction {#sec-intro}

Machine Learning models like Deep Neural Networks have become so complex and opaque over recent years that they are generally considered black-box systems. This lack of transparency exacerbates several other problems typically associated with these models: they tend to be unstable \cite{goodfellow2014explaining}, encode existing biases \cite{buolamwini2018gender} and learn representations that are surprising or even counter-intuitive from a human perspective \cite{sturm2014simple}. Nonetheless, they often form the basis for data-driven decision-making systems in real-world applications.

As others have pointed out, this scenario gives rise to an undesirable principal-agent problem involving a group of principals---i.e. human stakeholders---that fail to understand the behaviour of their agent---i.e. the black-box system \cite{borch2022machine}. The group of principals may include programmers, product managers and other decision-makers who develop and operate the system as well as those individuals ultimately subject to the decisions made by the system. In practice, decisions made by black-box systems are typically left unchallenged since the group of principals cannot scrutinize them:

> “You cannot appeal to (algorithms). They do not listen. Nor do they bend.” \cite{oneil2016weapons}

In light of all this, a quickly growing body of literature on Explainable Artificial Intelligence (XAI) has emerged. Counterfactual Explanations fall into this broad category. They can help human stakeholders make sense of the systems they develop, use or endure: they explain how inputs into a system need to change for it to produce different decisions. Explainability benefits internal as well as external quality assurance. Explanations that involve plausible and actionable changes can be used for Algorithmic Recourse (AR): they offer the group of principals a way to not only understand their agent's behaviour but also adjust or react to it.

The availability of open-source software to explain black-box models through counterfactuals is still limited. Through the work presented here, we aim to close that gap and thereby contribute to broader community efforts towards XAI. We envision this package to one day be the go-to place for Counterfactual Explanations in Julia. Thanks to Julia's unique support for interoperability with foreign programming languages we believe that this library may also benefit the broader machine learning and data science community.

Our package provides a simple and intuitive interface to generate CE for many standard classification models trained in Julia, as well as in Python and R. It comes with detailed documentation involving various illustrative example datasets, models and counterfactual generators for binary and multi-class prediction tasks. A carefully designed package architecture allows for a seamless extension of the package functionality through custom generators and models.

The remainder of this article is structured as follows: @sec-related presents related work on XAI as well as a brief overview of the methodological framework underlying CE. @sec-arch introduces the Julia package and its high-level architecture. @sec-use presents several basic and advanced usage examples. In @sec-custom we demonstrate how the package functionality can be customized and extended. To illustrate its practical usability, we explore examples involving real-world data in @sec-emp. Finally, we also discuss the current limitations of our package, as well as its future outlook in @sec-outlook. @sec-conclude concludes.

# Background and related work {#sec-related}

In this section, we first briefly introduce the broad field of Explainable AI, before narrowing it down to Counterfactual Explanations. We introduce the methodological framework and finally point to existing open-source software.

## Literature on Explainable AI

The field of XAI is still relatively young and made up of a variety of subdomains, definitions, concepts and taxonomies. Covering all of these is beyond the scope of this article, so we will focus only on high-level concepts. The following literature surveys provide more detail: Arrieta et al. (2020) provide a broad overview of XAI \cite{arrieta2020explainable}; Fan et al. (2020) focus on explainability in the context of deep learning \cite{fan2020interpretability}; and finally, Karimi et al. (2020) \cite{karimi2020survey} and Verma et al. (2020) \cite{verma2020counterfactual} offer detailed reviews of the literature on Counterfactual Explanations and Algorithmic Recourse (see also \cite{molnar2020interpretable} and \cite{varshney2022trustworthy}). Miller (2019) explicitly discusses the concept of explainability from the perspective of a social scientist \cite{miller2019explanation}.

The first broad distinction we want to make here is between **Interpretable** and **Explainable** AI. These terms are often used interchangeably, but this can lead to confusion. We find the distinction made in \cite{rudin2019stop} useful: Interpretable AI involves models that are inherently interpretable and transparent such as general additive models (GAM), decision trees and rule-based models; Explainable AI involves models that are not inherently interpretable but require additional tools to be explainable to humans. Examples of the latter include Ensembles, Support Vector Machines and Deep Neural Networks. Some would argue that we best avoid the second category of models altogether and instead focus solely on interpretable AI \cite{rudin2019stop}. While we agree that initial efforts should always be geared towards interpretable models, avoiding black boxes altogether would entail missed opportunities and anyway is probably not very realistic at this point. For that reason, we expect the need for XAI to persist in the medium term. Explainable AI can further be broadly divided into **global** and **local** explainability: the former is concerned with explaining the average behaviour of a model, while the latter involves explanations for individual predictions \cite{molnar2020interpretable}. Tools for global explainability include partial dependence plots (PDP), which involve the computation of marginal effects through Monte Carlo, and global surrogates. A surrogate model is an interpretable model that is trained to explain the predictions of a black-box model.

Counterfactual Explanations fall into the category of local methods: they explain how individual predictions change in response to individual feature perturbations. Among the most popular alternatives to Counterfactual Explanations are local surrogate explainers including Local Interpretable Model-agnostic Explanations (LIME) and Shapley additive explanations (SHAP). Since explanations produced by LIME and SHAP typically involve simple feature importance plots, they arguably rely on reasonably interpretable features at the very least. Contrary to Counterfactual Explanations, for example, it is not obvious how to apply LIME and SHAP to high-dimensional image data. Nonetheless, local surrogate explainers are among the most widely used XAI tools today, potentially because they are easy to interpret and implemented in popular programming languages. Proponents of surrogate explainers also commonly mention that there is a straightforward way to assess their reliability: a surrogate model that generates predictions in line with those produced by the black-box model is said to have high **fidelity** and therefore considered reliable. As intuitive as this notion may be, it also points to an obvious shortfall of surrogate explainers: even a high-fidelity surrogate model that produces the same predictions as the black-box model 99 per cent of the time is useless and potentially misleading for every 1 out of 100 individual predictions.

A recent study has shown that even experienced data scientists tend to put too much trust in explanations produced by LIME and SHAP \cite{kaur2020interpreting}. Another recent work has shown that both methods can be easily fooled: they depend on random input perturbations, a property that can be abused by adverse agents to essentially whitewash strongly biased black-box models \cite{slack2020fooling}. In related work, the same authors find that while gradient-based Counterfactual Explanations can also be manipulated, there is a straightforward way to protect against this in practice \cite{slack2021counterfactual}. In the context of quality assessment, it is also worth noting that---contrary to surrogate explainers---CE always achieve full fidelity by construction: counterfactuals are searched with respect to the black-box classifier, not some proxy for it. That being said, CE should also be used with care and research around them is still in its early stages.

## A framework for Counterfactual Explanations {#sec-method}

Counterfactual search involves feature perturbations: we are interested in understanding how we need to change individual attributes in order to change the model output to a desired value or label \cite{molnar2020interpretable}. Typically the underlying methodology is presented in the context of binary classification: $M: \mathcal{X} \mapsto \mathcal{Y}$ where $\mathcal{X}\subset\mathbb{R}^D$ and $\mathcal{Y}=\{0,1\}$. Further, let $t=1$ be the target class and let $x$ denote the factual feature vector of some individual sample outside of the target class, so $y=M(x)=0$. We follow this convention here, though it should be noted that the ideas presented here also carry over to multi-class problems and regression \cite{molnar2020interpretable}.

The counterfactual search objective originally proposed by Wachter et al. (2017) \cite{wachter2017counterfactual} is as follows

$$
\min_{x^\prime \in \mathcal{X}} h(x^\prime) \ \ \ \mbox{s. t.} \ \ \ M(x^\prime) = t
$$ {#eq-obj}

where $h(\cdot)$ quantifies how complex or costly it is to go from the factual $x$ to the counterfactual $x^\prime$. To simplify things we can restate this constrained objective as the following unconstrained and differentiable problem:

$$
x^\prime = \arg \min_{x^\prime}  \ell(M(x^\prime),t) + \lambda h(x^\prime)
$$ {#eq-solution}

Here $\ell$ denotes some loss function targeting the deviation between the target label and the predicted label and $\lambda$ governs the strength of the complexity penalty. Provided we have gradient access for the black-box model $M$ the solution to this problem can be found through gradient descent. This generic framework lays the foundation for most state-of-the-art approaches to counterfactual search and is also used as the baseline approach in our package. The hyperparameter $\lambda$ is typically tuned through grid search or in some sense pre-determined by the nature of the problem. Conventional choices for $\ell$ include margin-based losses like cross-entropy loss and hinge loss. It is worth pointing out that the loss function is typically computed with respect to logits rather than predicted probabilities, a convention that we have chosen to follow.^[Implementations of loss functions with respect to logits are often numerically more stable. For example, the `logitbinarycrossentropy(ŷ, y)` implementation in `Flux.Losses` (used here) is more stable than the mathematically equivalent `binarycrossentropy(ŷ, y)`.]

Numerous extensions to this simple approach have been developed since CE were first proposed in 2017 (see \cite{verma2020counterfactual} and \cite{karimi2020survey} for surveys). The various approaches largely differ in that they use different flavours of search objective defined in @eq-solution. Different penalties are often used to address many of the desirable properties of effective CE that have been set out. These desiderata include: **proximity** --- the distance between factual and counterfactual features should be small \cite{wachter2017counterfactual}; **actionability** --- the proposed recourse should be actionable (\cite{ustun2019actionable}, \cite{poyiadzi2020face}); **plausibility** --- the counterfactual explanation should be plausible to a human (\cite{joshi2019realistic}, \cite{schut2021generating}); **sparsity** --- the counterfactual explanation should involve as few individual feature changes as possible \cite{schut2021generating}; **robustness** --- the counterfactual explanation should be robust to domain and model shifts \cite{upadhyay2021robust}; **diversity** --- ideally multiple diverse counterfactuals should be provided \cite{mothilal2020explaining}; and **causality** --- counterfactuals should respect the structural causal model underlying the data generating process (\cite{karimi2020algorithmic},\cite{karimi2021algorithmic}).

Beyond gradient-based counterfactual search, which has been the main focus in our development so far, various methodologies have been proposed that can handle non-differentiable models like decision trees. We have implemented some of these approaches and will discuss them further in @sec-gen.

## Existing software

To the best of our knowledge, the package introduced here provides the first implementation of Counterfactual Explanations in Julia and therefore represents a novel contribution to the community. As for other programming languages, we are only aware of one other unifying framework: the Python library [CARLA](https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest) \cite{pawelczyk2021carla}.^[While we were writing this paper, the `R` package `counterfactuals` was released \cite{dandl2023counterfactuals}. The developers seem to also envision a unifying framework, but the project appears to still be in its early stages.] In addition to that, there exists open-source code for some specific approaches to CE that have been proposed in recent years. The approach-specific implementations that we have been able to find are generally well-documented, but exclusively in Python. For example, a PyTorch implementation of a greedy generator for Bayesian models proposed in \cite{schut2021generating} has been released. As another example, the popular [InterpretML](https://github.com/interpretml) library includes an implementation of a diverse counterfactual generator \cite{mothilal2020explaining}.

Generally speaking, software development in the space of XAI has largely focused on various global methods and surrogate explainers: implementations of PDP, LIME and SHAP are available for both Python (e.g. [`lime`](https://github.com/marcotcr/lime), [`shap`](https://github.com/slundberg/shap)) and R (e.g. [`lime`](https://cran.r-project.org/web/packages/lime/index.html), [`iml`](https://cran.r-project.org/web/packages/lime/index.html), [`shapper`](https://modeloriented.github.io/shapper/), [`fastshap`](https://github.com/bgreenwell/fastshap)). In the Julia space, there exist two packages related to XAI: firstly, [`ShapML.jl`](https://github.com/nredell/ShapML.jl), which provides a fast implementation of SHAP; and, secondly, [`ExplainableAI.jl`](https://github.com/adrhill/ExplainableAI.jl), which enables users to easily visualise gradients and activation maps for `Flux.jl` models. We also should not fail to mention the comprehensive [Interpretable AI](https://docs.interpretable.ai/stable/IAIBase/data/) infrastructure, which focuses exclusively on interpretable models.

Arguably the current availability of tools for explaining black-box models in Julia is limited, but it appears that the community is invested in changing that. The team behind `MLJ.jl`, for example, recruited contributors for a project about both Interpretable and Explainable AI in 2022.^[For details, see the Google Summer of Code 2022 project proposal: [https://julialang.org/jsoc/gsoc/MLJ/#interpretable_machine_learning_in_julia](https://julialang.org/jsoc/gsoc/MLJ/#interpretable_machine_learning_in_julia).] With our work on Counterfactual Explanations we hope to contribute to these efforts. We think that because of its unique transparency the Julia language naturally lends itself towards building Trustworthy AI systems.

# Introducing: `CounterfactualExplanations.jl` {#sec-arch}

@fig-arch provides an overview of the package architecture. It is built around two core modules that are designed to be as extensible as possible through dispatch: 1) `Models` is concerned with making any arbitrary model compatible with the package; 2) `Generators` is used to implement counterfactual search algorithms. The core function of the package---`generate_counterfactual`---uses an instance of type `<:AbstractFittedModel` produced by the `Models` module and an instance of type `<:AbstractGenerator` produced by the `Generators` module. Relating this to the methodology outlined in @sec-method, the former instance corresponds to the model $M$, while the latter defines the rules for the counterfactual search (@eq-solution).

![High-level schematic overview of package architecture. Modules are shown in red, structs in green and functions in purple.](www/pkg_architecture.png){#fig-arch width=3.33333in height=2.38095in}

## Models

The package currently offers native support for models built and trained in [Flux](https://fluxml.ai/) \cite{innes2018flux} as well as a small subset of models made available through [MLJ](https://alan-turing-institute.github.io/MLJ.jl/dev/) \cite{blaom2020mlj}. While in general it is assumed that users resort to this package to explain their pre-trained models, we provide a simple API call to train the following [models](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/):

\begin{itemize}
\item Linear Classifier (Logistic Regression and Multinomial Logit)
\item Multi-Layer Perceptron (Deep Neural Network)
\item Deep Ensemble \cite{lakshminarayanan2016simple}
\item Decision Tree, Random Forest, Gradient Boosted Trees
\end{itemize}

As we demonstrate below, it is straightforward to extend the package through custom models. Support for `torch` models trained in Python or R is also available.^[We are currently relying on `PythonCall.jl` and `RCall.jl` and this functionality is still somewhat brittle. Since this is more of an edge case, we may move this feature into its own package in the future.]

## Generators {#sec-gen}

A large and growing number of counterfactual [generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/) have already been implemented in our package (Table \ref{tab-gen}). At a high level, we distinguish generators in terms of their compatible model types, their default search space, and their composability. All "gradient-based" generators are compatible with differentiable models, e.g. `Flux` and `torch`, while "tree-based" generators are only applicable to models that involve decision trees. Concerning the search space, it is possible to search counterfactuals in a lower-dimensional latent embedding of the feature space that implicitly encodes the data-generating process (DGP). To learn the latent embedding, existing work has typically relied on generative models or existing causal knowledge (\cite{joshi2019realistic}, \cite{karimi2021algorithmic}). While this notion is compatible with all of our gradient-based generators, only some generators search a latent space by default. Finally, composability implies that the given generator can be blended with any other composable generator, which we discuss in @sec-gen-comp.

Beyond these broad technical distinctions, generators largely differ in terms of how they address the various desiderata mentioned above: *ClapROAR* aims to preserve the classifier, i.e. to generate counterfactuals that are robust to endogenous model shifts \cite{altmeyer2023endogenous}; *CLUE* searches plausible counterfactuals in the latent embedding of a generative model by explicitly minimising predictive entropy \cite{antoran2020getting}; *DiCE* is designed to generate multiple, maximally diverse counterfactuals \cite{mothilal2020explaining}; *FeatureTweak* leverages the internals of decision trees to search counterfactuals on a feature-by-feature basis, finding the counterfactual that tweaks the features in the least costly way \cite{tolomei2017interpretable}; *Gravitational* aims to generate plausible and robust counterfactuals by minimising the distance to observed samples in the target class \cite{altmeyer2023endogenous}; *Greedy* aims to generate plausible counterfactuals by implicitly minimising predictive uncertainty of Bayesian classifiers \cite{schut2021generating}; *GrowingSpheres* is model-agnostic, relying solely on identifying nearest neighbours of counterfactuals in the target class by gradually increasing the search radius and then moving counterfactuals in that direction \cite{laugel2017inversea}; *PROBE* generates probabilistically robust counterfactuals \cite{pawelczyk2022probabilistically}; *REVISE* addresses the need for plausibility by searching counterfactuals in the latent embedding of a Variational Autoencoder (VAE) \cite{joshi2019realistic}; *Wachter* is the baseline approach that only penalises the distance to the original sample \cite{wachter2017counterfactual}.

\begin{table}
\caption{Overview of implemented counterfactual generators. \label{tab-gen} \newline}
\centering
\fontsize{7}{9}\selectfont
\begin{tabular}[t]{llll}
\toprule
Generator & Model Type & Search Space & Composable\\
\midrule
ClaPROAR \cite{altmeyer2023endogenous} & gradient based & feature & yes\\
CLUE \cite{antoran2020getting} & gradient based & latent & yes\\
DiCE \cite{mothilal2020explaining} & gradient based & feature & yes\\
FeatureTweak \cite{tolomei2017interpretable} & tree based & feature & no\\
Gravitational \cite{altmeyer2023endogenous} & gradient based & feature & yes\\
Greedy \cite{schut2021generating} & gradient based & feature & yes\\
GrowingSpheres \cite{laugel2017inversea} & agnostic & feature & no\\
PROBE \cite{pawelczyk2022probabilistically} & gradient based & feature & no\\
REVISE \cite{joshi2019realistic} & gradient based & latent & yes\\
Wachter \cite{wachter2017counterfactual} & gradient based & feature & yes\\
\bottomrule
\end{tabular}
\end{table}

## Data Catalogue

To allow researchers and practitioners to test and compare counterfactual generators, the package ships with [catalogues](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/data_catalogue/) of pre-processed synthetic and real-world benchmark datasets from different domains. Real-world datasets include:

\begin{itemize}
\item Adult Census \cite{becker1996adult}
\item California Housing \cite{pace1997sparse}
\item CIFAR10 \cite{krizhevsky2009learning}
\item German Credit \cite{hoffman1994german}
\item Give Me Some Credit \cite{kaggle2011give}
\item MNIST \cite{lecun1998mnist} and Fashion MNIST \cite{xiao2017fashion}
\item UCI defaultCredit \cite{yeh2009comparisons}
\end{itemize}

Custom datasets can also be easily preprocessed as explained in the [documentation](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/data_preprocessing/).

## Plotting

The package also extends common `Plots.jl` methods to facilitate the visualization of results. Calling the generic `plot()` method on an instance of type `<:CounterfactualExplanation`, for example, generates a plot visualizing the entire counterfactual path in the feature space^[For multi-dimensional input data, standard dimensionality reduction techniques are used to compress the data. In this case, the classifier's decision boundary is approximated through a Nearest Neighbour model. This is still somewhat experimental and will be improved in the future.]. We will see several examples of this below.

# Basic Usage {#sec-use}

In the following, we begin our exploration of the package functionality with a simple example. We then demonstrate how more advanced generators can be easily composed and show how users can impose mutability constraints on features. Finally, we also briefly explore the topics of counterfactual evaluation and benchmarking.

## A Simple Generic Generator {#sec-simple}

Code \ref{lst:simple} below provides a complete example demonstrating how the framework presented in @sec-method can be implemented through our package. Using a synthetic data set with linearly separable features we first fit a linear classifier (line \ref{line:simple-class}). Next, we define the target class (line \ref{line:simple-t}) and then draw a random sample from the other class (line \ref{line:simple-x}). Finally, we instantiate a generic generator (line \ref{line:simple-gen}) and run the counterfactual search (line \ref{line:simple-search}). @fig-binary illustrates the resulting counterfactual path in the two-dimensional feature space. Features go through iterative perturbations until the desired confidence level is reached as illustrated by the contour in the background, which shows the softmax output for the target class.

```{julia}
# Data and Classifier:
counterfactual_data = load_linearly_separable()
M = fit_model(counterfactual_data, :Linear)

# Factual and Target:
yhat = predict_label(M, counterfactual_data)
target = 2    # target label
candidates = findall(vec(yhat) .!= target)
chosen = rand(candidates)
x = select_factual(counterfactual_data, chosen)

# Counterfactual search:
generator = GenericGenerator()
ce = generate_counterfactual(
    x, target, counterfactual_data, M, generator)
```

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:simple}, caption={Standard workflow for generating counterfactuals.}] 
# Data and Classifier:
counterfactual_data = load_linearly_separable()
M = fit_model(counterfactual_data, :Linear) @\label{line:simple-class}@

# Factual and Target:
yhat = predict_label(M, counterfactual_data)
target = 2    # target label @\label{line:simple-t}@
candidates = findall(vec(yhat) .!= target)
chosen = rand(candidates)
x = select_factual(counterfactual_data, chosen) @\label{line:simple-x}@

# Counterfactual search:
generator = GenericGenerator() @\label{line:simple-gen}@
ce = generate_counterfactual(
    x, target, counterfactual_data, M, generator) @\label{line:simple-search}@
\end{lstlisting}

```{julia}
binary_data = deepcopy(counterfactual_data)     # store to reuse
```

```{julia}
plt = Plots.plot(ce, legend=:topright)
display(plt)
savefig(plt, "paper/www/ce_binary.png")
```

![Counterfactual path using generic counterfactual generator for conventional binary classifier.](www/ce_binary.png){#fig-binary width=20pc height=15pc}

In this simple example, the generic generator produces a valid counterfactual, since the decision boundary is crossed and the predicted label is flipped. But the counterfactual is not plausible: it does not appear to be generated by the same DGP as the observed data in the target class. This is because the generic generator does not take into account any of the desiderata mentioned in @sec-method, except for the distance to the factual sample.

## Composing Generators {#sec-gen-comp}

To address these issues, we can leverage the ideas underlying some of the more advanced counterfactual generators introduced above. In particular, we will now show how easy it is to [compose custom generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/generators/) that blend different ideas through user-friendly macros.

Suppose we wanted to address the desiderata of plausibility and diversity. We could do this by blending ideas underlying the *DiCE* generator with the *REVISE* generator. Formally, the corresponding search objective would be defined as follows,

$$
\mathbf{Z}^\prime = \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^{L \times K}} \{  {\ell(M(f(\mathbf{Z}^\prime)),t)} + \lambda \cdot {\text{diversity}(f(\mathbf{Z}^\prime)) }  \} 
$$ {#eq-comp}

where $\mathbf{X}^\prime$ is an $L$-dimensional array of counterfactuals, $f: \mathcal{Z}^{L \times K} \mapsto \mathcal{X}^{L \times D}$ is a function that maps the $L \times K$-dimensional latent space $\mathcal{Z}$ to the $L \times D$-dimensional feature space $\mathcal{X}$ and $\text{diversity}(\cdot)$ is the penalty proposed by Mothilal et al. (2020) \cite{mothilal2020explaining} that induces diverse sets of counterfactuals. As in @eq-solution, $\ell$ is the loss function, $M$ is the black-box model, $t$ is the target class, and $\lambda$ is the strength of the penalty.

Code \ref{lst:composed} demonstrates how @eq-comp can be seamlessly translated into Julia code. We begin by instantiating a `GradientBasedGenerator` in line \ref{line:composed-init}. Next, we use chained macros for composition: firstly, we define the counterfactual search `@objective` corresponding to *DiCE* in line \ref{line:composed-dice}; secondly, we define the latent space search strategy corresponding to *REVISE* using the `@search_latent_space` macro in line \ref{line:composed-latent}; finally, we specify our prefered optimisation method using the `@with_optimiser` macro in line \ref{line:composed-adam}.

```{julia}
# Composition:
generator = GradientBasedGenerator()
@chain generator begin
    @objective logitcrossentropy + 0.2ddp_diversity     # DiCE (Mothilal et al. 2020)                  
    @search_latent_space                                # REVISE (Joshi et al. 2019)
    @with_optimiser Adam(0.005)
end

# Counterfactual search:
counterfactuals = generate_counterfactual(
    x, target, counterfactual_data, M, generator;
    num_counterfactuals=5, 
    converge_when=:generator_conditions,
)
# Plotting
plt = Plots.plot(counterfactuals)
display(plt)
savefig(plt, "paper/www/binary_advanced.png")
```

\begin{lstlisting}[language=Julia, escapechar=§, numbers=left, label={lst:composed}, caption={Composing a custom generator.}]
generator = GradientBasedGenerator() §\label{line:composed-init}§
@chain generator begin
    @objective logitcrossentropy 
      + 0.2ddp_diversity §\label{line:composed-dice}§
    @search_latent_space §\label{line:composed-latent}§
    @with_optimiser Adam(0.005) §\label{line:composed-adam}§
end
\end{lstlisting}

In this case, the counterfactual search is performed in the latent space of a Variational Autoencoder (VAE) that is automatically trained on the observed data. It is important to specify the keyword argument `num_counterfactuals` of the `generate_counterfactual` to some value higher than $1$ (default), to ensure that the diversity penalty is effective. The resulting counterfactual path is shown in @fig-binary-advanced below. We observe that the resulting counterfactuals are diverse and the majority of them are plausible.

![Counterfactual path using the *DiCE* generator.](www/binary_advanced.png){#fig-binary-advanced width=20pc height=15pc}

## Mutability Constraints {#sec-mut}

In practice, features usually cannot be perturbed arbitrarily. Suppose, for example, that one of the features used by a bank to predict the creditworthiness of its clients is *age*. If a counterfactual explanation for the prediction model indicates that older clients should "grow younger" to improve their creditworthiness, then this is an interesting insight (it reveals age bias), but the provided recourse is not actionable. In such cases, we may want to constrain the mutability of features. To illustrate how this can be implemented in our package, we will continue with the example from above.

Mutability can be defined in terms of four different options: 1) the feature is mutable in both directions, 2) the feature can only increase (e.g. *age*), 3) the feature can only decrease (e.g. *time left* until your next deadline) and 4) the feature is not mutable (e.g. *skin colour*, *ethnicity*, ...). To specify which category a feature belongs to, users can pass a vector of symbols containing the mutability constraints at the pre-processing stage. For each feature one can choose from these four options: `:both` (mutable in both directions), `:increase` (only up), `:decrease` (only down) and `:none` (immutable). By default, `nothing` is passed to that keyword argument and it is assumed that all features are mutable in both directions.^[Mutability constraints are not yet implemented for Latent Space search.]

We can impose that the first feature is immutable as follows: `counterfactual_data.mutability = [:none, :both]`. The resulting counterfactual path is shown in @fig-mutability below. Since only the second feature can be perturbed, the sample can only move along the vertical axis. In this case, the counterfactual search does not yield a valid counterfactual, since the target class is not reached.

```{julia}
counterfactual_data.mutability = [:none, :both]
```

```{julia}
# Define generator:
generator = GenericGenerator(opt=Descent(0.05))
# Generate recourse:
ce = generate_counterfactual(x, target, counterfactual_data, M, generator)
```

```{julia}
#| output: true

plt = Plots.plot(ce)
display(plt)
savefig(plt, joinpath(www_path, "constraint_mutability.png"))
```

```{julia}
counterfactual_data.mutability = [:both, :both]
counterfactual_data.standardize=true
```

![Counterfactual path with immutable feature.](www/constraint_mutability.png){#fig-mutability width=20pc height=15pc}

## Evaluation and Benchmarking {#sec-eval}

The package also makes it easy to [evaluate](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/evaluation/) counterfactuals with respect to many of the desiderata mentioned above. For example, users may want to infer how costly the provided recourse is to individuals. To this end, we can measure the distance of the counterfactual from its original value. The API call to compute the distance metric defined in Wachter et al. (2017) \cite{wachter2017counterfactual}, for instance, is as simple as `evaluate(ce; measure=distance_mad)`, where `ce` can also be a vector of `CounterfactualExplanation`s.

Additionally, the package provides a [benchmarking](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/benchmarking/) framework that allows users to compare the performance of different generators on a given dataset. In @fig-bmk we show the results of a benchmark comparing several generators in terms of the average cost and implausibility of the generated counterfactuals. The cost is proxied by the L1-norm of the difference between the factual and counterfactual features, while implausibility is measured by the distance of the counterfactuals from samples in the target class. The results illustrate that there is a tradeoff between minimizing costs to individuals and generating plausible counterfactuals.

```{julia}
# Models:
model_dict = Dict(
    :Linear => M,
)

# Generators:
generator_dict = Dict(
    :Gravitational => GravitationalGenerator(),
    :Greedy => GreedyGenerator(),
    :REVISE => REVISEGenerator(),
    :Wachter => WachterGenerator(),
)

# Measures:
measures = [
    CounterfactualExplanations.Objectives.distance_l1,
    CounterfactualExplanations.Objectives.distance_from_targets,
]

# Benchmark:
bmk = benchmark(
    counterfactual_data; 
    models=model_dict, 
    generators=generator_dict, 
    measure=measures,
    suppress_training=true, dataname="Linearly Separable",
    n_individuals=5,
    target=target,
    initialization=:identity,
    converge_when=:generator_conditions,
    store_ce=true,
)
CSV.write(joinpath(output_path, "bmk.csv"), bmk())
```

```{julia}
# Plotting:
df = @chain bmk() begin
    @mutate(variable = ifelse.(variable .== "distance_l1", "Cost", variable))
    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
end
using CairoMakie
using AlgebraOfGraphics: data, visual, mapping, draw, BoxPlot
plt = data(df) * visual(BoxPlot) * 
    mapping(:generator, :value => "Value", col=:model, row=:variable, color=:generator => "Generator")
plt = draw(
    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
    facet=(; linkyaxes=:none)
)   
display(plt)
save(joinpath(www_path, "bmk.png"), plt, px_per_unit = 3)
```

![Benchmarking results for different generators.](www/bmk.png){#fig-bmk width=20pc height=15pc}

# Customization and Extensibility {#sec-custom}

One of our priorities has been to make our package customizable and extensible. In the long term, we aim to add support for more default models and counterfactual generators. In the short term, it is designed to allow users to integrate models and generators themselves. These community efforts will facilitate our long-term goals.

## Adding Custom Models {#sec-custom-mod}

At the high level, only two steps are necessary to make any supervised learning model compatible with our package:

\begin{unnumlist}
\item \textbf{Subtyping}: We need to subtype the \texttt{AbstractFittedModel}.
\item \textbf{Dispatch}: The functions \texttt{logits} and \texttt{probs} need to be extended through custom methods for the model in question.
\end{unnumlist}

To demonstrate how this can be done in practice, we will reiterate here how native support for [`Flux.jl`](https://fluxml.ai/) (\cite{innes2018flux}) deep learning models was enabled.^[Flux models are now natively supported by our package and can be instantiated by calling `FluxModel()`.] Once again we use synthetic data for an illustrative example. Code \ref{lst:nn} below builds a simple model architecture that can be used for a multi-class prediction task. Note how outputs from the final layer are not passed through a softmax activation function, since the counterfactual loss is evaluated with respect to logits as we discussed earlier. The model is trained with dropout.

```{julia}
# Data:
counterfactual_data = load_multi_class()
y = counterfactual_data.y
X = counterfactual_data.X

# Flux model setup: 
using Flux
using MLUtils
dl = DataLoader((X,y), batchsize=1)
n_hidden = 32
output_dim = size(y,1)
input_dim = 2
activation = σ
model = Flux.Chain(
    Dense(input_dim, n_hidden, activation),
    Dropout(0.1),
    Dense(n_hidden, output_dim)
)  
loss(x, y) = Flux.Losses.logitcrossentropy(model(x), y)

# Flux model training:
using Flux.Optimise: update!, Adam
opt = Adam()
epochs = 50
for epoch = 1:epochs
  for d in dl
    gs = gradient(Flux.params(model)) do
      l = loss(d...)
    end
    update!(opt, Flux.params(model), gs)
  end
end
```

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:nn}, caption={A simple neural network model.}]
n_hidden = 32
output_dim = length(unique(y))
input_dim = 2
model = Chain(
    Dense(input_dim, n_hidden, activation),
    Dropout(0.1),
    Dense(n_hidden, output_dim)
)  
\end{lstlisting}

```{julia}
# Step 1)
struct MyFluxModel <: AbstractDifferentiableModel
    model::Any
    likelihood::Symbol
end

# Step 2)
# import functions in order to extend
import CounterfactualExplanations.Models: logits
import CounterfactualExplanations.Models: probs 
logits(M::MyFluxModel, X::AbstractArray) = M.model(X)
probs(M::MyFluxModel, X::AbstractArray) = softmax(logits(M, X))
M = MyFluxModel(model, :classification_multi)
```

Code \ref{lst:mymodel} below implements the two steps that were necessary to make Flux models compatible with the package. In line \ref{line:mymodel-subtype} we declare our new struct as a subtype of `AbstractDifferentiableModel`, which itself is an abstract subtype of `AbstractFittedModel`.^[Note that in line \ref{line:mymodel-likelihood} we also provide a field determining the likelihood. This is optional and only used internally to determine which loss function to use in the counterfactual search. If this field is not provided to the model, the loss function needs to be explicitly supplied to the generator.] Computing logits amounts to just calling the model on inputs. Predicted probabilities for labels can be computed by passing logits through the softmax function.

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:mymodel}, caption={A wrapper for Flux models.}]
# Step 1)
struct MyFluxModel <: AbstractDifferentiableModel @\label{line:mymodel-subtype}@
    model::Any
    likelihood::Symbol @\label{line:mymodel-likelihood}@
end

# Step 2)
# import functions in order to extend
import CounterfactualExplanations.Models: logits
import CounterfactualExplanations.Models: probs 
logits(M::MyFluxModel, X::AbstractArray) = M.model(X)
probs(M::MyFluxModel, X::AbstractArray) = softmax(logits(M, X))
M = MyFluxModel(model)
\end{lstlisting}

```{julia}
# Randomly selected factual:
target = 2
factual = 4
chosen = rand(findall(predict_label(M, counterfactual_data) .== factual))
x = select_factual(
    counterfactual_data, chosen)

# Counterfactual search:
generator = GenericGenerator(opt=Flux.Descent(0.1))
ce = generate_counterfactual(
    x, target, counterfactual_data, M, generator)
```

The API call for generating counterfactuals for our new model is the same as before. @fig-multi shows the resulting counterfactual path for a randomly chosen sample. In this case, the contour shows the predicted probability that the input is in the target class ($t=2$).

```{julia}
#| output: true

plt = Plots.plot(ce)
display(plt)
savefig(plt, "paper/www/ce_multi.png")
```

![Counterfactual path using generic counterfactual generator for multi-class classifier.](www/ce_multi.png){#fig-multi width=20pc height=15pc}

## Adding Custom Generators {#sec-custom-gen}

In some cases, composability may not be sufficient to implement specific logics underlying certain counterfactual generators. In such cases, users may want to implement custom generators. To illustrate how this can be done we will consider a simple extension of our `GenericGenerator`. As we have seen above, Counterfactual Explanations are not unique. In light of this, we might be interested in quantifying the uncertainty around the generated counterfactuals \cite{delaney2021uncertainty}. One idea could be, to use dropout to randomly switch features on and off in each iteration. Without dwelling further on the merit of this idea, we will now briefly show how this can be implemented.

### A Generator with Dropout

Code \ref{lst:dropout} below implements two important steps: 1) create an abstract subtype of the `AbstractGradientBasedGenerator` and 2) create a constructor with an additional field for the dropout probability.

```{julia}
# Abstract suptype:
abstract type AbstractDropoutGenerator <: AbstractGradientBasedGenerator end

# Constructor:
struct DropoutGenerator <: AbstractDropoutGenerator
    loss::Function # loss function
    penalty::Function
    λ::AbstractFloat # strength of penalty
    latent_space::Bool
    opt::Any # optimizer
    p_dropout::AbstractFloat # dropout rate
end

# Instantiate:
generator = DropoutGenerator(
    logitbinarycrossentropy,
    distance_l1,
    0.1,
    false,
    Flux.Optimise.Descent(0.1),
    0.5,
)
```

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:dropout}, caption={Building a custom generator with dropout.}]
# Abstract suptype:
abstract type AbstractDropoutGenerator <: AbstractGradientBasedGenerator end
# Constructor:
struct DropoutGenerator <: AbstractDropoutGenerator
    loss::Symbol # loss function
    complexity::Function # complexity function
    @$\lambda$@::AbstractFloat # strength of penalty
    decision_threshold::Union{Nothing,AbstractFloat} 
    opt::Any # optimizer
    @$\tau$@::AbstractFloat # tolerance for convergence
    p_dropout::AbstractFloat # dropout rate
end
\end{lstlisting}

Next, in Code \ref{lst:generate} we define how feature perturbations are generated for our custom dropout generator: in particular, we extend the relevant function through a method that implements the dropout logic.

```{julia}
using CounterfactualExplanations.Generators
using StatsBase
function Generators.generate_perturbations(
    generator::AbstractDropoutGenerator, 
    ce::CounterfactualExplanation
)
    s′ = deepcopy(ce.s′)
    new_s′ = Generators.propose_state(generator, ce)
    Δs′ = new_s′ - s′ # gradient step

    # Dropout:
    set_to_zero = sample(
        1:length(Δs′),
        Int(round(generator.p_dropout*length(Δs′))),
        replace=false
    )
    Δs′[set_to_zero] .= 0
    return Δs′
end
```

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:generate}, caption={Generating feature perturbations with dropout.}]
using CounterfactualExplanations.Generators
function Generators.generate_perturbations(
    generator::AbstractDropoutGenerator, 
    ce::CouterfactualExplanation
)
    @$s^\prime$@ = deepcopy(ce.@$s^\prime$@)
    new_@$s^\prime$@ = Generators.propose_state(
        generator, ce)
    @$\Delta s^\prime$@ = new_@$s^\prime$@ - @$s^\prime$@ # gradient step
    # Dropout:
    set_to_zero = sample(
        1:length(@$\Delta s^\prime$@),
        Int(round(generator.p_dropout*length(@$\Delta s^\prime$@))),
        replace=false
    )
    @$\Delta s^\prime$@[set_to_zero] .= 0
    return @$\Delta s^\prime$@
end
\end{lstlisting}

Finally, we proceed to generate counterfactuals in the same way we always do. The resulting counterfactual path is shown in @fig-dropout.

```{julia}
# Data and Classifier:
counterfactual_data = binary_data
M = fit_model(counterfactual_data, :DeepEnsemble)

# Factual and Target:
yhat = predict_label(M, counterfactual_data)
target = 2    # target label
candidates = findall(vec(yhat) .!= target)
chosen = rand(candidates)
x = select_factual(counterfactual_data, chosen)

# Counterfactual search:
ce = generate_counterfactual(
    x, target, counterfactual_data, M, generator;
    num_counterfactuals=5)
```

```{julia}
#| output: true

plt = Plots.plot(ce)
display(plt)
savefig(plt, joinpath(www_path, "dropout.png"))
```

![Counterfactual path for a generator with dropout.](www/dropout.png){#fig-dropout width=20pc height=15pc}

# A Real-World Examples {#sec-emp}

Now that we have explained the basic functionality of `CounterfactualExplanations.jl` through some synthetic examples, it is time to work through examples involving real-world data.

## Give Me Some Credit

The *Give Me Some Credit* dataset is one of the tabular real-world datasets that ship with the package \cite{kaggle2011give}. It can be used to train a binary classifier to predict whether a borrower is likely to experience financial difficulties in the next two years. In particular, we have an output variable $y \in \{0=\texttt{no stress},1=\texttt{stress}\}$ and a feature matrix $X$ that includes socio-demographic variables like `age` and `income`. A retail bank might use such a classifier to determine if potential borrowers should receive credit or not.

```{julia}
Random.seed!(2023)
counterfactual_data = load_gmsc(10000)
train_data, test_data = CounterfactualExplanations.DataPreprocessing.train_test_split(counterfactual_data)
flux_training_params.batchsize = 50
flux_training_params.verbose = true
M = fit_model(train_data, :MLP; dropout=true, n_hidden=64, n_layers=4)
ŷ = predict_label(M, train_data)
# Randomly selected factual:
target_class = 1
non_target_class = 0
x = select_factual(train_data,rand(findall(vec(ŷ).==non_target_class),10))
CounterfactualExplanations.reset!(flux_training_params)
Serialization.serialize(joinpath(output_path,"credit_classifier.jls"), M)
```

For the classification task, we use a Multi-Layer Perceptron with dropout regularization. Using the Gravitational generator \cite{altmeyer2023endogenous} we will generate counterfactuals for ten randomly chosen individuals that would be denied credit based on our pre-trained model. Concerning the mutability of features, we only impose that the `age` cannot be decreased.

```{julia}
#| output: false

using Markdown
Markdown.parse("""
"The F1-score of our pre-trained classifier on test data is: $(round(model_evaluation(M, test_data)[1], digits=2))"
""")
```

```{julia}
input_names = Symbol.([
  "unsecured_lines",
  "age",
  "times_past_due",
  "debt_ratio",
  "income",
  "number_loans",
  "times_late",
  "number_mortgages",
  "times_past_due_2",
  "number_dependents",
])
_mut = [:both for i in 1:length(input_names)]
_mut[2] = :increase
counterfactual_data.mutability = _mut
x1 = :income   # Amount of given credit
x2 = :age
```

```{julia}
generator = GravitationalGenerator(
    opt = Descent(0.1)
)
counterfactuals = generate_counterfactual(
  x, target_class, counterfactual_data, M, generator;
  initialization = :identity, decision_threshold = 0.9,
)
```

```{julia}
#| output: true

# Plotting:
plt = Plots.plot()

map(enumerate(counterfactuals)) do (idx, x)

    x_origin = x.x
    x_origin = selectdim(x_origin,3,1)
    x_origin = vcat(x_origin[findall(input_names.==x1)], x_origin[findall(input_names.==x2)])

    _x = CounterfactualExplanations.counterfactual(x)
    _x = selectdim(_x,3,1)
    _x = vcat(_x[findall(input_names.==x1)], _x[findall(input_names.==x2)])

    if idx == 1
        non_target_label = "Loan denied"
        target_label = "Loan provided"
    else
        non_target_label = ""
        target_label = ""
    end

    Plots.scatter!(
        plt, 
        [x_origin[1]],
        [x_origin[2]],
        xlabel = x1,
        ylabel = x2,
        colour = non_target_class,
        label = non_target_label,
        ms = 10
    )

    # Marker
    Plots.scatter!(
        plt, 
        [_x[1]],
        [_x[2]],
        colour = target_class,
        label = target_label,
        ms = 10
    )

    # Arrow
    Plots.plot!(
        plt, 
        vcat(x_origin[1],_x[1]), 
        vcat(x_origin[2],_x[2]), 
        arrow=true, color=:black, linewidth=2, label=""
    )

end

using Plots.PlotMeasures
plt = Plots.plot(plt, size=(500, 200), legend=:topright, xlabel="Income", ylabel="Age", margin=5mm, ylims=(-2.5,3.0))
display(plt)
savefig(plt, joinpath(www_path, "credit.png"))
```

@fig-credit shows the resulting counterfactuals proposed by Wachter in the two-dimensional feature space spanned by the `age` and `income` variables. An increase in income and age is recommended for the majority of individuals, which seems plausible: both age and income are typically positively related to creditworthiness.

![Give Me Some Credit: counterfactuals for would-be borrowers proposed by the Gravitational Generator.](www/credit.png){#fig-credit width=20pc height=8pc}

## MNIST

For our second example, we will look at image data. The MNIST dataset contains 60,000 training samples of handwritten digits in the form of 28x28 pixel grey-scale images \cite{lecun1998mnist}. Each image is associated with a label indicating the digit (0-9) that the image represents. The data makes for an interesting case study of CE because humans have a good idea of what plausible counterfactuals of digits look like. For example, if you were asked to pick up an eraser and turn the digit in the left panel of @fig-mnist into a four (4) you would know exactly what to do: just erase the top part.

```{julia}
Random.seed!(2023)
using CounterfactualExplanations.Models: load_mnist_mlp, load_mnist_ensemble, load_mnist_vae
counterfactual_data = load_mnist()
X, y = unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()
vae = load_mnist_vae()
vae_weak = load_mnist_vae(;strong=false)
Serialization.serialize(joinpath(output_path,"mnist_classifier.jls"), M)
Serialization.serialize(joinpath(output_path,"mnist_vae.jls"), vae)
Serialization.serialize(joinpath(output_path,"mnist_vae_weak.jls"), vae_weak)
```

On the model side, we will use a simple multi-layer perceptron (MLP). Code \ref{lst:mnist-setup} loads the data and the pre-trained MLP. It also loads two pre-trained Variational Auto-Encoders, which will be used by our counterfactual generator of choice for this task: *REVISE*.

```{julia}
#| output: true
using Markdown
test_data = load_mnist_test()
Markdown.parse("""
"The F1-score of our pre-trained image classifier on test data is: $(round(model_evaluation(M, test_data)[1], digits=2))"
""")
```

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:mnist-setup}, caption={Loading pre-trained models and data for MNIST.}]
counterfactual_data = load_mnist()
X, y = unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()
vae = load_mnist_vae()
vae_weak = load_mnist_vae(;strong=false)
\end{lstlisting}

The proposed counterfactuals are shown in @fig-mnist. In the case in which *REVISE* has access to an expressive VAE (centre), the result looks convincing: the perturbed image does look like it represents a four (4). In terms of explainability, we may conclude that removing the top part of the handwritten nine (9) leads the black-box model to predict that the perturbed image represents a four (4). We should note, however, that the quality of counterfactuals produced by *REVISE* hinges on the performance of the underlying generative model, as demonstrated by the result on the right. In this case, *REVISE* uses a weak VAE and the resulting counterfactual is invalid. In light of this, we recommend using Latent Space search with care.

```{julia}
# Randomly selected factual:
x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==9))],input_dim,1)
counterfactual_data.generative_model = vae # assign generative model
input_dim = size(X,1)
target = 4
factual = predict_label(M, counterfactual_data, x)[1]
γ = 0.99

# Define generator:
generator = REVISEGenerator(
  opt = Descent(0.1),
  λ=0.01
)
# Generate recourse:
ce = generate_counterfactual(x, target, counterfactual_data, M, generator; decision_threshold = γ)
counterfactual_data = deepcopy(counterfactual_data)
counterfactual_data.generative_model = vae_weak
ce_weak = generate_counterfactual(x, target, counterfactual_data, M, generator; decision_threshold = γ)
```

```{julia}
using MLDatasets: convert2image
using Images
image_size = 300
bg_color = :blue
p1 = Plots.plot(
    convert2image(MNIST, reshape(x,28,28)),
    axis=nothing, 
    size=(image_size, image_size),
    title="Factual"
)
savefig(p1, joinpath(www_path, "mnist_factual.png"))
target_prob = round(target_probs(ce)[1],digits=2)
ŷ = CounterfactualExplanations.counterfactual_label(ce)[1]
p2 = Plots.plot(
    convert2image(MNIST, reshape(CounterfactualExplanations.counterfactual(ce),28,28)),
    axis=nothing, 
    size=(image_size, image_size),
    title="REVISE (strong VAE)"
)
savefig(p2, joinpath(www_path, "mnist_vae_strong.png"))
target_prob = round(target_probs(ce_weak)[1],digits=2)
ŷ = CounterfactualExplanations.counterfactual_label(ce_weak)[1]
p3 = Plots.plot(
    convert2image(MNIST, reshape(CounterfactualExplanations.counterfactual(ce_weak),28,28)),
    axis=nothing, 
    size=(image_size, image_size),
    title="REVISE (weak VAE)"
)
savefig(p3, joinpath(www_path, "mnist_vae_weak.png"))
display(p1)
display(p2)
display(p3)
plt = Plots.plot(p1,p2,p3;size=(image_size * 3,image_size), layout=(1,3))
display(plt)
savefig(plt, joinpath(www_path, "mnist_$(factual)to$(target)_latent.png"))
```

![Counterfactual explanations for MNIST using a Latent Space generator: turning a nine (9) into a four (4).](www/mnist_9to4_latent.png){#fig-mnist width=20pc height=6.66666pc}

# Discussion and Outlook {#sec-outlook}

We believe that this package in its current form offers a valuable contribution to ongoing efforts towards XAI in Julia. That being said, there is significant scope for future developments, which we briefly outline in this final section.

## Candidate models and generators

The package supports various models and generators either natively or through minimal augmentation. In future work, we would like to prioritize the addition of further predictive models and generators. Concerning the former, it would be useful to add native support for any supervised models built in `MLJ.jl`, an extensive Machine Learning framework for Julia \cite{blaom2020mlj}. This may also involve adding support for regression models as well as additional non-differentiable models. In terms of counterfactual generators, there is a list of recent methodologies that we would like to implement including MINT \cite{karimi2021algorithmic}, ROAR \cite{upadhyay2021robust} and FACE \cite{poyiadzi2020face}.

## Additional datasets

For benchmarking and testing purposes it will be crucial to add more datasets to our library. We have so far prioritized tabular datasets that have typically been used in the literature on counterfactual explanations including *Adult*, *Give Me Some Credit* and *German Credit* \cite{karimi2020survey}. There is scope for adding data sources that have so far not been explored much in this context including additional image datasets as well as audio, natural language and time-series data.

# Concluding remarks {#sec-conclude}

`CounterfactualExplanation.jl` is a package for generating Counterfactual Explanations and Algorithmic Recourse in Julia. Through various synthetic and real-world examples, we have demonstrated the basic usage of the package as well as its extensibility. The package has already served us in our research to benchmark various methodological approaches to Counterfactual Explanations and Algorithmic Recourse. We therefore strongly believe that it should help other practitioners and researchers in their own efforts towards Trustworthy AI.

We envision this package to one day constitute the go-to place for explaining arbitrary predictive models through an extensive suite of counterfactual generators. As a major next step, we aim to make our library as compatible as possible with the popular [`MLJ.jl`](https://alan-turing-institute.github.io/MLJ.jl/dev/) package for machine learning in Julia.  We invite the Julia community to contribute to these goals through usage, open challenge and active development.

# Acknowledgements {#sec-ack}

We are immensely grateful to the group of TU Delft students who contributed huge improvements to this package as part of a university project in 2023: Rauno Arike, Simon Kasdorp, Lauri Kesküll, Mariusz Kicior, Vincent Pikand. We also want to thank the broader Julia community for being welcoming and open and for supporting research contributions like this one. Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING---TU Delft collaboration.

### 2023-11-15 examples/CounterfactualExplanations.jl/paper.qmd
almost done

@@
 ---
+title: "Explaining Black-Box Models through Counterfactuals"
+keywords:
+  - Julia
+  - Explainable Artificial Intelligence
+  - Counterfactual Explanations
+  - Algorithmic Recourse
+authors:
+  - name: Patrick Altmeyer
+    orcid: 0000-0003-4726-8613
+    affiliation: Delft University of Technology
+  - name: Arie van Deursen
+    affiliation: Delft University of Technology
+  - name: Cynthia C. S. Liem
+    affiliation: Delft University of Technology
+abstract: |
+    We present `CounterfactualExplanations.jl`: a package for generating Counterfactual Explanations (CE) and Algorithmic Recourse (AR) for black-box models in Julia.
+CE explain how inputs into a model need to change to yield specific model predictions.
+Explanations that involve realistic and actionable changes can be used to provide AR: a set of proposed actions for individuals to change an undesirable outcome for the better.
+In this article, we discuss the usefulness of CE for Explainable Artificial Intelligence and demonstrate the functionality of our package.
+The package is straightforward to use and designed with a focus on customization and extensibility.
+We envision it to one day be the go-to place for explaining arbitrary predictive models in Julia through a diverse suite of counterfactual generators.
 format:
@@
 Machine Learning models like Deep Neural Networks have become so complex and opaque over recent years that they are generally considered black-box systems.
-This lack of transparency exacerbates several other problems typically associated with these models: they tend to be unstable \cite{goodfellow2014explaining}, encode existing biases \cite{buolamwini2018gender} and learn representations that are surprising or even counter-intuitive from a human perspective \cite{sturm2014simple}.
+This lack of transparency exacerbates several other problems typically associated with these models: they tend to be unstable [@goodfellow2014explaining], encode existing biases [@buolamwini2018gender] and learn representations that are surprising or even counter-intuitive from a human perspective [@buolamwini2018gender].
 Nonetheless, they often form the basis for data-driven decision-making systems in real-world applications.
@@
 human stakeholders---that fail to understand the behaviour of their agent---i.e.
-the black-box system \cite{borch2022machine}.
+the black-box system [@borch2022machine].
 The group of principals may include programmers, product managers and other decision-makers who develop and operate the system as well as those individuals ultimately subject to the decisions made by the system.
@@
 
-> “You cannot appeal to (algorithms). They do not listen. Nor do they bend.” \cite{oneil2016weapons}
+> “You cannot appeal to (algorithms). They do not listen. Nor do they bend.” [@oneil2016weapons]
 
@@
 The following literature surveys provide more detail: Arrieta et al.
-(2020) provide a broad overview of XAI \cite{arrieta2020explainable}; Fan et al.
-(2020) focus on explainability in the context of deep learning \cite{fan2020interpretability}; and finally, Karimi et al.
-(2020) \cite{karimi2020survey} and Verma et al.
-(2020) \cite{verma2020counterfactual} offer detailed reviews of the literature on Counterfactual Explanations and Algorithmic Recourse (see also \cite{molnar2020interpretable} and \cite{varshney2022trustworthy}).
-Miller (2019) explicitly discusses the concept of explainability from the perspective of a social scientist \cite{miller2019explanation}.
+(2020) provide a broad overview of XAI [@arrieta2020explainable]; Fan et al.
+(2020) focus on explainability in the context of deep learning [@fan2020interpretability]; and finally, Karimi et al.
+(2020) [@karimi2020survey] and Verma et al.
+(2020) @verma2020counterfactual offer detailed reviews of the literature on Counterfactual Explanations and Algorithmic Recourse (see also @molnar2020interpretable and @varshney2022trustworthy).
+Miller (2019) explicitly discusses the concept of explainability from the perspective of a social scientist [@miller2019explanation].
 
@@
 These terms are often used interchangeably, but this can lead to confusion.
-We find the distinction made in \cite{rudin2019stop} useful: Interpretable AI involves models that are inherently interpretable and transparent such as general additive models (GAM), decision trees and rule-based models; Explainable AI involves models that are not inherently interpretable but require additional tools to be explainable to humans.
+We find the distinction made in @rudin2019stop useful: Interpretable AI involves models that are inherently interpretable and transparent such as general additive models (GAM), decision trees and rule-based models; Explainable AI involves models that are not inherently interpretable but require additional tools to be explainable to humans.
 Examples of the latter include Ensembles, Support Vector Machines and Deep Neural Networks.
-Some would argue that we best avoid the second category of models altogether and instead focus solely on interpretable AI \cite{rudin2019stop}.
+Some would argue that we best avoid the second category of models altogether and instead focus solely on interpretable AI @rudin2019stop.
 While we agree that initial efforts should always be geared towards interpretable models, avoiding black boxes altogether would entail missed opportunities and anyway is probably not very realistic at this point.
 For that reason, we expect the need for XAI to persist in the medium term.
-Explainable AI can further be broadly divided into **global** and **local** explainability: the former is concerned with explaining the average behaviour of a model, while the latter involves explanations for individual predictions \cite{molnar2020interpretable}.
+Explainable AI can further be broadly divided into **global** and **local** explainability: the former is concerned with explaining the average behaviour of a model, while the latter involves explanations for individual predictions [@molnar2020interpretable].
 Tools for global explainability include partial dependence plots (PDP), which involve the computation of marginal effects through Monte Carlo, and global surrogates.
@@
 
-A recent study has shown that even experienced data scientists tend to put too much trust in explanations produced by LIME and SHAP \cite{kaur2020interpreting}.
-Another recent work has shown that both methods can be easily fooled: they depend on random input perturbations, a property that can be abused by adverse agents to essentially whitewash strongly biased black-box models \cite{slack2020fooling}.
-In related work, the same authors find that while gradient-based Counterfactual Explanations can also be manipulated, there is a straightforward way to protect against this in practice \cite{slack2021counterfactual}.
+A recent study has shown that even experienced data scientists tend to put too much trust in explanations produced by LIME and SHAP [@kaur2020interpreting].
+Another recent work has shown that both methods can be easily fooled: they depend on random input perturbations, a property that can be abused by adverse agents to essentially whitewash strongly biased black-box models [@slack2020fooling].
+In related work, the same authors find that while gradient-based Counterfactual Explanations can also be manipulated, there is a straightforward way to protect against this in practice [@slack2021counterfactual].
 In the context of quality assessment, it is also worth noting that---contrary to surrogate explainers---CE always achieve full fidelity by construction: counterfactuals are searched with respect to the black-box classifier, not some proxy for it.
@@
 
-Counterfactual search involves feature perturbations: we are interested in understanding how we need to change individual attributes in order to change the model output to a desired value or label \cite{molnar2020interpretable}.
+Counterfactual search involves feature perturbations: we are interested in understanding how we need to change individual attributes in order to change the model output to a desired value or label [@molnar2020interpretable].
 Typically the underlying methodology is presented in the context of binary classification: $M: \mathcal{X} \mapsto \mathcal{Y}$ where $\mathcal{X}\subset\mathbb{R}^D$ and $\mathcal{Y}=\{0,1\}$.
 Further, let $t=1$ be the target class and let $x$ denote the factual feature vector of some individual sample outside of the target class, so $y=M(x)=0$.
-We follow this convention here, though it should be noted that the ideas presented here also carry over to multi-class problems and regression \cite{molnar2020interpretable}.
-
-The counterfactual search objective originally proposed by Wachter et al. (2017) \cite{wachter2017counterfactual} is as follows
+We follow this convention here, though it should be noted that the ideas presented here also carry over to multi-class problems and regression [@molnar2020interpretable].
+
+The counterfactual search objective originally proposed by @wachter2017counterfactual is as follows
 
@@
 
-Numerous extensions to this simple approach have been developed since CE were first proposed in 2017 (see \cite{verma2020counterfactual} and \cite{karimi2020survey} for surveys).
+Numerous extensions to this simple approach have been developed since CE were first proposed in 2017 (see @verma2020counterfactual and [@karimi2020survey] for surveys).
 The various approaches largely differ in that they use different flavours of search objective defined in @eq-solution.
 Different penalties are often used to address many of the desirable properties of effective CE that have been set out.
-These desiderata include: **proximity** --- the distance between factual and counterfactual features should be small \cite{wachter2017counterfactual}; **actionability** --- the proposed recourse should be actionable (\cite{ustun2019actionable}, \cite{poyiadzi2020face}); **plausibility** --- the counterfactual explanation should be plausible to a human (\cite{joshi2019realistic}, \cite{schut2021generating}); **sparsity** --- the counterfactual explanation should involve as few individual feature changes as possible \cite{schut2021generating}; **robustness** --- the counterfactual explanation should be robust to domain and model shifts \cite{upadhyay2021robust}; **diversity** --- ideally multiple diverse counterfactuals should be provided \cite{mothilal2020explaining}; and **causality** --- counterfactuals should respect the structural causal model underlying the data generating process (\cite{karimi2020algorithmic},\cite{karimi2021algorithmic}).
+These desiderata include: **proximity** --- the distance between factual and counterfactual features should be small [@wachter2017counterfactual]; **actionability** --- the proposed recourse should be actionable [@ustun2019actionable, @poyiadzi2020face]; **plausibility** --- the counterfactual explanation should be plausible to a human [@joshi2019realistic, @schut2021generating]; **sparsity** --- the counterfactual explanation should involve as few individual feature changes as possible [@schut2021generating]; **robustness** --- the counterfactual explanation should be robust to domain and model shifts [@upadhyay2021robust]; **diversity** --- ideally multiple diverse counterfactuals should be provided [@mothilal2020explaining]; and **causality** --- counterfactuals should respect the structural causal model underlying the data generating process [@karimi2020algorithmic,@karimi2021algorithmic].
 
@@
 To the best of our knowledge, the package introduced here provides the first implementation of Counterfactual Explanations in Julia and therefore represents a novel contribution to the community.
-As for other programming languages, we are only aware of one other unifying framework: the Python library [CARLA](https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest) \cite{pawelczyk2021carla}.^[While we were writing this paper, the `R` package `counterfactuals` was released \cite{dandl2023counterfactuals}.
+As for other programming languages, we are only aware of one other unifying framework: the Python library [CARLA](https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest) [@pawelczyk2021carla].^[While we were writing this paper, the `R` package `counterfactuals` was released [@dandl2023counterfactuals].
 The developers seem to also envision a unifying framework, but the project appears to still be in its early stages.] In addition to that, there exists open-source code for some specific approaches to CE that have been proposed in recent years.
 The approach-specific implementations that we have been able to find are generally well-documented, but exclusively in Python.
-For example, a PyTorch implementation of a greedy generator for Bayesian models proposed in \cite{schut2021generating} has been released.
-As another example, the popular [InterpretML](https://github.com/interpretml) library includes an implementation of a diverse counterfactual generator \cite{mothilal2020explaining}.
+For example, a PyTorch implementation of a greedy generator for Bayesian models proposed in @schut2021generating has been released.
+As another example, the popular [InterpretML](https://github.com/interpretml) library includes an implementation of a diverse counterfactual generator [@mothilal2020explaining].
 
@@
 
-The package currently offers native support for models built and trained in [Flux](https://fluxml.ai/) \cite{innes2018flux} as well as a small subset of models made available through [MLJ](https://alan-turing-institute.github.io/MLJ.jl/dev/) \cite{blaom2020mlj}.
+The package currently offers native support for models built and trained in [Flux](https://fluxml.ai/) [@innes2018flux] as well as a small subset of models made available through [MLJ](https://alan-turing-institute.github.io/MLJ.jl/dev/) [@blaom2020mlj].
 While in general it is assumed that users resort to this package to explain their pre-trained models, we provide a simple API call to train the following [models](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/):
 
-\begin{itemize}
-\item Linear Classifier (Logistic Regression and Multinomial Logit)
-\item Multi-Layer Perceptron (Deep Neural Network)
-\item Deep Ensemble \cite{lakshminarayanan2016simple}
-\item Decision Tree, Random Forest, Gradient Boosted Trees
-\end{itemize}
+- Linear Classifier (Logistic Regression and Multinomial Logit)
+- Multi-Layer Perceptron (Deep Neural Network)
+- Deep Ensemble @lakshminarayanan2016simple
+- Decision Tree, Random Forest, Gradient Boosted Trees
 
@@
 
-A large and growing number of counterfactual [generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/) have already been implemented in our package (Table \ref{tab-gen}).
+A large and growing number of counterfactual [generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/) have already been implemented in our package (@tab-gen).
 At a high level, we distinguish generators in terms of their compatible model types, their default search space, and their composability.
@@
 Concerning the search space, it is possible to search counterfactuals in a lower-dimensional latent embedding of the feature space that implicitly encodes the data-generating process (DGP).
-To learn the latent embedding, existing work has typically relied on generative models or existing causal knowledge (\cite{joshi2019realistic}, \cite{karimi2021algorithmic}).
+To learn the latent embedding, existing work has typically relied on generative models or existing causal knowledge [@joshi2019realistic, @karimi2021algorithmic].
 While this notion is compatible with all of our gradient-based generators, only some generators search a latent space by default.
@@
 Beyond these broad technical distinctions, generators largely differ in terms of how they address the various desiderata mentioned above: *ClapROAR* aims to preserve the classifier, i.e.
-to generate counterfactuals that are robust to endogenous model shifts \cite{altmeyer2023endogenous}; *CLUE* searches plausible counterfactuals in the latent embedding of a generative model by explicitly minimising predictive entropy \cite{antoran2020getting}; *DiCE* is designed to generate multiple, maximally diverse counterfactuals \cite{mothilal2020explaining}; *FeatureTweak* leverages the internals of decision trees to search counterfactuals on a feature-by-feature basis, finding the counterfactual that tweaks the features in the least costly way \cite{tolomei2017interpretable}; *Gravitational* aims to generate plausible and robust counterfactuals by minimising the distance to observed samples in the target class \cite{altmeyer2023endogenous}; *Greedy* aims to generate plausible counterfactuals by implicitly minimising predictive uncertainty of Bayesian classifiers \cite{schut2021generating}; *GrowingSpheres* is model-agnostic, relying solely on identifying nearest neighbours of counterfactuals in the target class by gradually increasing the search radius and then moving counterfactuals in that direction \cite{laugel2017inversea}; *PROBE* generates probabilistically robust counterfactuals \cite{pawelczyk2022probabilistically}; *REVISE* addresses the need for plausibility by searching counterfactuals in the latent embedding of a Variational Autoencoder (VAE) \cite{joshi2019realistic}; *Wachter* is the baseline approach that only penalises the distance to the original sample \cite{wachter2017counterfactual}.
-
-\begin{table}
-\caption{Overview of implemented counterfactual generators. \label{tab-gen} \newline}
-\centering
-\fontsize{7}{9}\selectfont
-\begin{tabular}[t]{llll}
+to generate counterfactuals that are robust to endogenous model shifts [@altmeyer2023endogenous]; *CLUE* searches plausible counterfactuals in the latent embedding of a generative model by explicitly minimising predictive entropy [@antoran2020getting]; *DiCE* is designed to generate multiple, maximally diverse counterfactuals [@mothilal2020explaining]; *FeatureTweak* leverages the internals of decision trees to search counterfactuals on a feature-by-feature basis, finding the counterfactual that tweaks the features in the least costly way [@tolomei2017interpretable]; *Gravitational* aims to generate plausible and robust counterfactuals by minimising the distance to observed samples in the target class [@altmeyer2023endogenous]; *Greedy* aims to generate plausible counterfactuals by implicitly minimising predictive uncertainty of Bayesian classifiers [@schut2021generating]; *GrowingSpheres* is model-agnostic, relying solely on identifying nearest neighbours of counterfactuals in the target class by gradually increasing the search radius and then moving counterfactuals in that direction[ @laugel2017inversea]; *PROBE* generates probabilistically robust counterfactuals [@pawelczyk2022probabilistically]; *REVISE* addresses the need for plausibility by searching counterfactuals in the latent embedding of a Variational Autoencoder (VAE) [@joshi2019realistic]; *Wachter* is the baseline approach that only penalises the distance to the original sample [@wachter2017counterfactual].
+
+::: {#tab-gen}
+
+```{=latex}
+\begin{tabular}{llll}
 \toprule
@@
 \midrule
-ClaPROAR \cite{altmeyer2023endogenous} & gradient based & feature & yes\\
-CLUE \cite{antoran2020getting} & gradient based & latent & yes\\
-DiCE \cite{mothilal2020explaining} & gradient based & feature & yes\\
-FeatureTweak \cite{tolomei2017interpretable} & tree based & feature & no\\
-Gravitational \cite{altmeyer2023endogenous} & gradient based & feature & yes\\
-Greedy \cite{schut2021generating} & gradient based & feature & yes\\
-GrowingSpheres \cite{laugel2017inversea} & agnostic & feature & no\\
-PROBE \cite{pawelczyk2022probabilistically} & gradient based & feature & no\\
-REVISE \cite{joshi2019realistic} & gradient based & latent & yes\\
-Wachter \cite{wachter2017counterfactual} & gradient based & feature & yes\\
+ClaPROAR [@altmeyer2023endogenous] & gradient based & feature & yes\\
+CLUE [@antoran2020getting] & gradient based & latent & yes\\
+DiCE [@mothilal2020explaining] & gradient based & feature & yes\\
+FeatureTweak [@tolomei2017interpretable] & tree based & feature & no\\
+Gravitational [@altmeyer2023endogenous] & gradient based & feature & yes\\
+Greedy [@schut2021generating] & gradient based & feature & yes\\
+GrowingSpheres [@laugel2017inversea] & agnostic & feature & no\\
+PROBE [@pawelczyk2022probabilistically] & gradient based & feature & no\\
+REVISE [@joshi2019realistic] & gradient based & latent & yes\\
+Wachter [@wachter2017counterfactual] & gradient based & feature & yes\\
 \bottomrule
 \end{tabular}
-\end{table}
+```
+
+Overview of implemented counterfactual generators.
+
+:::
 
@@
 
-\begin{itemize}
-\item Adult Census \cite{becker1996adult}
-\item California Housing \cite{pace1997sparse}
-\item CIFAR10 \cite{krizhevsky2009learning}
-\item German Credit \cite{hoffman1994german}
-\item Give Me Some Credit \cite{kaggle2011give}
-\item MNIST \cite{lecun1998mnist} and Fashion MNIST \cite{xiao2017fashion}
-\item UCI defaultCredit \cite{yeh2009comparisons}
-\end{itemize}
+- Adult Census [@becker1996adult]
+- California Housing [@pace1997sparse]
+- CIFAR10 [@krizhevsky2009learning]
+- German Credit [@hoffman1994german]
+- Give Me Some Credit [@kaggle2011give]
+- MNIST [@lecun1998mnist] and Fashion MNIST [@xiao2017fashion]
+- UCI defaultCredit [@yeh2009comparisons]
 
@@
 
-where $\mathbf{X}^\prime$ is an $L$-dimensional array of counterfactuals, $f: \mathcal{Z}^{L \times K} \mapsto \mathcal{X}^{L \times D}$ is a function that maps the $L \times K$-dimensional latent space $\mathcal{Z}$ to the $L \times D$-dimensional feature space $\mathcal{X}$ and $\text{diversity}(\cdot)$ is the penalty proposed by Mothilal et al.
-(2020) \cite{mothilal2020explaining} that induces diverse sets of counterfactuals.
+where $\mathbf{X}^\prime$ is an $L$-dimensional array of counterfactuals, $f: \mathcal{Z}^{L \times K} \mapsto \mathcal{X}^{L \times D}$ is a function that maps the $L \times K$-dimensional latent space $\mathcal{Z}$ to the $L \times D$-dimensional feature space $\mathcal{X}$ and $\text{diversity}(\cdot)$ is the penalty proposed by @mothilal2020explaining that induces diverse sets of counterfactuals.
 As in @eq-solution, $\ell$ is the loss function, $M$ is the black-box model, $t$ is the target class, and $\lambda$ is the strength of the penalty.
@@
 To this end, we can measure the distance of the counterfactual from its original value.
-The API call to compute the distance metric defined in Wachter et al.
-(2017) \cite{wachter2017counterfactual}, for instance, is as simple as `evaluate(ce; measure=distance_mad)`, where `ce` can also be a vector of `CounterfactualExplanation`s.
+The API call to compute the distance metric defined in @wachter2017counterfactual, for instance, is as simple as `evaluate(ce; measure=distance_mad)`, where `ce` can also be a vector of `CounterfactualExplanation`s.
 
@@
 
-\begin{unnumlist}
-\item \textbf{Subtyping}: We need to subtype the \texttt{AbstractFittedModel}.
-\item \textbf{Dispatch}: The functions \texttt{logits} and \texttt{probs} need to be extended through custom methods for the model in question.
-\end{unnumlist}
-
-To demonstrate how this can be done in practice, we will reiterate here how native support for [`Flux.jl`](https://fluxml.ai/) (\cite{innes2018flux}) deep learning models was enabled.^[Flux models are now natively supported by our package and can be instantiated by calling `FluxModel()`.] Once again we use synthetic data for an illustrative example.
+- \textbf{Subtyping}: We need to subtype the \texttt{AbstractFittedModel}.
+- \textbf{Dispatch}: The functions \texttt{logits} and \texttt{probs} need to be extended through custom methods for the model in question.
+
+To demonstrate how this can be done in practice, we will reiterate here how native support for [`Flux.jl`](https://fluxml.ai/) [@innes2018flux] deep learning models was enabled.^[Flux models are now natively supported by our package and can be instantiated by calling `FluxModel()`.] Once again we use synthetic data for an illustrative example.
 Code \ref{lst:nn} below builds a simple model architecture that can be used for a multi-class prediction task.
@@
 As we have seen above, Counterfactual Explanations are not unique.
-In light of this, we might be interested in quantifying the uncertainty around the generated counterfactuals \cite{delaney2021uncertainty}.
+In light of this, we might be interested in quantifying the uncertainty around the generated counterfactuals [@delaney2021uncertainty].
 One idea could be, to use dropout to randomly switch features on and off in each iteration.
@@
 
-The *Give Me Some Credit* dataset is one of the tabular real-world datasets that ship with the package \cite{kaggle2011give}.
+The *Give Me Some Credit* dataset is one of the tabular real-world datasets that ship with the package [@kaggle2011give].
 It can be used to train a binary classifier to predict whether a borrower is likely to experience financial difficulties in the next two years.
@@
 For the classification task, we use a Multi-Layer Perceptron with dropout regularization.
-Using the Gravitational generator \cite{altmeyer2023endogenous} we will generate counterfactuals for ten randomly chosen individuals that would be denied credit based on our pre-trained model.
+Using the Gravitational generator [@altmeyer2023endogenous] we will generate counterfactuals for ten randomly chosen individuals that would be denied credit based on our pre-trained model.
 Concerning the mutability of features, we only impose that the `age` cannot be decreased.
@@
 For our second example, we will look at image data.
-The MNIST dataset contains 60,000 training samples of handwritten digits in the form of 28x28 pixel grey-scale images \cite{lecun1998mnist}.
+The MNIST dataset contains 60,000 training samples of handwritten digits in the form of 28x28 pixel grey-scale images [@lecun1998mnist].
 Each image is associated with a label indicating the digit (0-9) that the image represents.
@@
 In future work, we would like to prioritize the addition of further predictive models and generators.
-Concerning the former, it would be useful to add native support for any supervised models built in `MLJ.jl`, an extensive Machine Learning framework for Julia \cite{blaom2020mlj}.
+Concerning the former, it would be useful to add native support for any supervised models built in `MLJ.jl`, an extensive Machine Learning framework for Julia [@blaom2020mlj].
 This may also involve adding support for regression models as well as additional non-differentiable models.
-In terms of counterfactual generators, there is a list of recent methodologies that we would like to implement including MINT \cite{karimi2021algorithmic}, ROAR \cite{upadhyay2021robust} and FACE \cite{poyiadzi2020face}.
+In terms of counterfactual generators, there is a list of recent methodologies that we would like to implement including MINT [@karimi2021algorithmic], ROAR [@upadhyay2021robust] and FACE [@poyiadzi2020face].
 
@@
 For benchmarking and testing purposes it will be crucial to add more datasets to our library.
-We have so far prioritized tabular datasets that have typically been used in the literature on counterfactual explanations including *Adult*, *Give Me Some Credit* and *German Credit* \cite{karimi2020survey}.
+We have so far prioritized tabular datasets that have typically been used in the literature on counterfactual explanations including *Adult*, *Give Me Some Credit* and *German Credit* [@karimi2020survey].
 There is scope for adding data sources that have so far not been explored much in this context including additional image datasets as well as audio, natural language and time-series data.
@@
 Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING---TU Delft collaboration.
+
+# References

### 2023-11-15 examples/CounterfactualExplanations.jl/paper.qmd
ce done

@@
 
-A large and growing number of counterfactual [generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/) have already been implemented in our package (@tab-gen).
+A large and growing number of counterfactual [generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/) have already been implemented in our package (@tbl-gen).
 At a high level, we distinguish generators in terms of their compatible model types, their default search space, and their composability.
@@
 
-::: {#tab-gen}
-
-```{=latex}
-\begin{tabular}{llll}
-\toprule
-Generator & Model Type & Search Space & Composable\\
-\midrule
-ClaPROAR [@altmeyer2023endogenous] & gradient based & feature & yes\\
-CLUE [@antoran2020getting] & gradient based & latent & yes\\
-DiCE [@mothilal2020explaining] & gradient based & feature & yes\\
-FeatureTweak [@tolomei2017interpretable] & tree based & feature & no\\
-Gravitational [@altmeyer2023endogenous] & gradient based & feature & yes\\
-Greedy [@schut2021generating] & gradient based & feature & yes\\
-GrowingSpheres [@laugel2017inversea] & agnostic & feature & no\\
-PROBE [@pawelczyk2022probabilistically] & gradient based & feature & no\\
-REVISE [@joshi2019realistic] & gradient based & latent & yes\\
-Wachter [@wachter2017counterfactual] & gradient based & feature & yes\\
-\bottomrule
-\end{tabular}
-```
-
-Overview of implemented counterfactual generators.
-
-:::
+\onecolumn
+
+| Generator | Model Type | Search Space | Composable |
+|---|---|---|---|
+| ClaPROAR [@altmeyer2023endogenous] | gradient based | feature | yes |
+| CLUE [@antoran2020getting] | gradient based | latent | yes |
+| DiCE [@mothilal2020explaining] | gradient based | feature | yes |
+| FeatureTweak [@tolomei2017interpretable] | tree based | feature | no |
+| Gravitational [@altmeyer2023endogenous] | gradient based | feature | yes |
+| Greedy [@schut2021generating] | gradient based | feature | yes |
+| GrowingSpheres [@laugel2017inversea] | agnostic | feature | no |
+| PROBE [@pawelczyk2022probabilistically] | gradient based | feature | no |
+| REVISE [@joshi2019realistic] | gradient based | latent | yes |
+| Wachter [@wachter2017counterfactual] | gradient based | feature | yes |
+
+: Overview of implemented counterfactual generators. {#tbl-gen}
+
+\twocolumn
 

### 2023-11-16 examples/CounterfactualExplanations.jl/paper.qmd

@@
 format:
-  juliacon-proceedings-pdf: default
+  juliacon-proceedings-pdf:
+    keep-tex: true  
+  juliacon-proceedings-html: default
 bibliography: ref.bib

### 2023-11-16 examples/CounterfactualExplanations.jl/paper.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%

\documentclass{juliacon}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage[]{tgtermes}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{orcidlink}
\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Explaining Black-Box Models through Counterfactuals},
  pdfkeywords={Julia, Explainable Artificial
Intelligence, Counterfactual Explanations, Algorithmic Recourse},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Explaining Black-Box Models through Counterfactuals}

\author[1]{Patrick Altmeyer}
\author[1]{Arie van Deursen}
\author[1]{Cynthia C. S. Liem}
\affil[1]{Delft University of Technology}
\date{}
\begin{document}
\maketitle

% Abstract
\begin{abstract}

We present \texttt{CounterfactualExplanations.jl}: a package for
generating Counterfactual Explanations (CE) and Algorithmic Recourse
(AR) for black-box models in Julia. CE explain how inputs into a model
need to change to yield specific model predictions. Explanations that
involve realistic and actionable changes can be used to provide AR: a
set of proposed actions for individuals to change an undesirable outcome
for the better. In this article, we discuss the usefulness of CE for
Explainable Artificial Intelligence and demonstrate the functionality of
our package. The package is straightforward to use and designed with a
focus on customization and extensibility. We envision it to one day be
the go-to place for explaining arbitrary predictive models in Julia
through a diverse suite of counterfactual generators.
\end{abstract}

% Keywords
\JCONkeywords{Julia, Explainable Artificial Intelligence, Counterfactual
Explanations, Algorithmic Recourse}

% Hypersetup
\hypersetup{
    pdftitle = {Explaining Black-Box Models through Counterfactuals},
    pdfsubject = {JuliaCon \@year Proceedings},
    pdfauthor = {},
    pdfkeywords = {Julia, Explainable Artificial
Intelligence, Counterfactual Explanations, Algorithmic Recourse},
}

\setcounter{page}{1}

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, frame hidden, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi

\setstretch{1}
\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

Machine Learning models like Deep Neural Networks have become so complex
and opaque over recent years that they are generally considered
black-box systems. This lack of transparency exacerbates several other
problems typically associated with these models: they tend to be
unstable (Goodfellow, Shlens, and Szegedy 2014), encode existing biases
(Buolamwini and Gebru 2018) and learn representations that are
surprising or even counter-intuitive from a human perspective
(Buolamwini and Gebru 2018). Nonetheless, they often form the basis for
data-driven decision-making systems in real-world applications.

As others have pointed out, this scenario gives rise to an undesirable
principal-agent problem involving a group of principals---i.e.~human
stakeholders---that fail to understand the behaviour of their
agent---i.e.~the black-box system (Borch 2022). The group of principals
may include programmers, product managers and other decision-makers who
develop and operate the system as well as those individuals ultimately
subject to the decisions made by the system. In practice, decisions made
by black-box systems are typically left unchallenged since the group of
principals cannot scrutinize them:

\begin{quote}
``You cannot appeal to (algorithms). They do not listen. Nor do they
bend.'' (O'Neil 2016)
\end{quote}

In light of all this, a quickly growing body of literature on
Explainable Artificial Intelligence (XAI) has emerged. Counterfactual
Explanations fall into this broad category. They can help human
stakeholders make sense of the systems they develop, use or endure: they
explain how inputs into a system need to change for it to produce
different decisions. Explainability benefits internal as well as
external quality assurance. Explanations that involve plausible and
actionable changes can be used for Algorithmic Recourse (AR): they offer
the group of principals a way to not only understand their agent's
behaviour but also adjust or react to it.

The availability of open-source software to explain black-box models
through counterfactuals is still limited. Through the work presented
here, we aim to close that gap and thereby contribute to broader
community efforts towards XAI. We envision this package to one day be
the go-to place for Counterfactual Explanations in Julia. Thanks to
Julia's unique support for interoperability with foreign programming
languages we believe that this library may also benefit the broader
machine learning and data science community.

Our package provides a simple and intuitive interface to generate CE for
many standard classification models trained in Julia, as well as in
Python and R. It comes with detailed documentation involving various
illustrative example datasets, models and counterfactual generators for
binary and multi-class prediction tasks. A carefully designed package
architecture allows for a seamless extension of the package
functionality through custom generators and models.

The remainder of this article is structured as follows:
Section~\ref{sec-related} presents related work on XAI as well as a
brief overview of the methodological framework underlying CE.
Section~\ref{sec-arch} introduces the Julia package and its high-level
architecture. Section~\ref{sec-use} presents several basic and advanced
usage examples. In Section~\ref{sec-custom} we demonstrate how the
package functionality can be customized and extended. To illustrate its
practical usability, we explore examples involving real-world data in
Section~\ref{sec-emp}. Finally, we also discuss the current limitations
of our package, as well as its future outlook in
Section~\ref{sec-outlook}. Section~\ref{sec-conclude} concludes.

\hypertarget{sec-related}{%
\section{Background and related work}\label{sec-related}}

In this section, we first briefly introduce the broad field of
Explainable AI, before narrowing it down to Counterfactual Explanations.
We introduce the methodological framework and finally point to existing
open-source software.

\hypertarget{literature-on-explainable-ai}{%
\subsection{Literature on Explainable
AI}\label{literature-on-explainable-ai}}

The field of XAI is still relatively young and made up of a variety of
subdomains, definitions, concepts and taxonomies. Covering all of these
is beyond the scope of this article, so we will focus only on high-level
concepts. The following literature surveys provide more detail: Arrieta
et al.~(2020) provide a broad overview of XAI (Arrieta et al. 2020); Fan
et al.~(2020) focus on explainability in the context of deep learning
(Fan, Xiong, and Wang 2020); and finally, Karimi et al.~(2020) (Karimi,
Barthe, et al. 2020) and Verma et al.~(2020) Verma, Dickerson, and Hines
(2020) offer detailed reviews of the literature on Counterfactual
Explanations and Algorithmic Recourse (see also Molnar (2020) and
Varshney (2022)). Miller (2019) explicitly discusses the concept of
explainability from the perspective of a social scientist (Miller 2019).

The first broad distinction we want to make here is between
\textbf{Interpretable} and \textbf{Explainable} AI. These terms are
often used interchangeably, but this can lead to confusion. We find the
distinction made in Rudin (2019) useful: Interpretable AI involves
models that are inherently interpretable and transparent such as general
additive models (GAM), decision trees and rule-based models; Explainable
AI involves models that are not inherently interpretable but require
additional tools to be explainable to humans. Examples of the latter
include Ensembles, Support Vector Machines and Deep Neural Networks.
Some would argue that we best avoid the second category of models
altogether and instead focus solely on interpretable AI Rudin (2019).
While we agree that initial efforts should always be geared towards
interpretable models, avoiding black boxes altogether would entail
missed opportunities and anyway is probably not very realistic at this
point. For that reason, we expect the need for XAI to persist in the
medium term. Explainable AI can further be broadly divided into
\textbf{global} and \textbf{local} explainability: the former is
concerned with explaining the average behaviour of a model, while the
latter involves explanations for individual predictions (Molnar 2020).
Tools for global explainability include partial dependence plots (PDP),
which involve the computation of marginal effects through Monte Carlo,
and global surrogates. A surrogate model is an interpretable model that
is trained to explain the predictions of a black-box model.

Counterfactual Explanations fall into the category of local methods:
they explain how individual predictions change in response to individual
feature perturbations. Among the most popular alternatives to
Counterfactual Explanations are local surrogate explainers including
Local Interpretable Model-agnostic Explanations (LIME) and Shapley
additive explanations (SHAP). Since explanations produced by LIME and
SHAP typically involve simple feature importance plots, they arguably
rely on reasonably interpretable features at the very least. Contrary to
Counterfactual Explanations, for example, it is not obvious how to apply
LIME and SHAP to high-dimensional image data. Nonetheless, local
surrogate explainers are among the most widely used XAI tools today,
potentially because they are easy to interpret and implemented in
popular programming languages. Proponents of surrogate explainers also
commonly mention that there is a straightforward way to assess their
reliability: a surrogate model that generates predictions in line with
those produced by the black-box model is said to have high
\textbf{fidelity} and therefore considered reliable. As intuitive as
this notion may be, it also points to an obvious shortfall of surrogate
explainers: even a high-fidelity surrogate model that produces the same
predictions as the black-box model 99 per cent of the time is useless
and potentially misleading for every 1 out of 100 individual
predictions.

A recent study has shown that even experienced data scientists tend to
put too much trust in explanations produced by LIME and SHAP (Kaur et
al. 2020). Another recent work has shown that both methods can be easily
fooled: they depend on random input perturbations, a property that can
be abused by adverse agents to essentially whitewash strongly biased
black-box models (Slack et al. 2020). In related work, the same authors
find that while gradient-based Counterfactual Explanations can also be
manipulated, there is a straightforward way to protect against this in
practice (Slack et al. 2021). In the context of quality assessment, it
is also worth noting that---contrary to surrogate explainers---CE always
achieve full fidelity by construction: counterfactuals are searched with
respect to the black-box classifier, not some proxy for it. That being
said, CE should also be used with care and research around them is still
in its early stages.

\hypertarget{sec-method}{%
\subsection{A framework for Counterfactual
Explanations}\label{sec-method}}

Counterfactual search involves feature perturbations: we are interested
in understanding how we need to change individual attributes in order to
change the model output to a desired value or label (Molnar 2020).
Typically the underlying methodology is presented in the context of
binary classification: \(M: \mathcal{X} \mapsto \mathcal{Y}\) where
\(\mathcal{X}\subset\mathbb{R}^D\) and \(\mathcal{Y}=\{0,1\}\). Further,
let \(t=1\) be the target class and let \(x\) denote the factual feature
vector of some individual sample outside of the target class, so
\(y=M(x)=0\). We follow this convention here, though it should be noted
that the ideas presented here also carry over to multi-class problems
and regression (Molnar 2020).

The counterfactual search objective originally proposed by Wachter,
Mittelstadt, and Russell (2017) is as follows

\begin{equation}\protect\hypertarget{eq-obj}{}{
\min_{x^\prime \in \mathcal{X}} h(x^\prime) \ \ \ \mbox{s. t.} \ \ \ M(x^\prime) = t
}\label{eq-obj}\end{equation}

where \(h(\cdot)\) quantifies how complex or costly it is to go from the
factual \(x\) to the counterfactual \(x^\prime\). To simplify things we
can restate this constrained objective as the following unconstrained
and differentiable problem:

\begin{equation}\protect\hypertarget{eq-solution}{}{
x^\prime = \arg \min_{x^\prime}  \ell(M(x^\prime),t) + \lambda h(x^\prime)
}\label{eq-solution}\end{equation}

Here \(\ell\) denotes some loss function targeting the deviation between
the target label and the predicted label and \(\lambda\) governs the
strength of the complexity penalty. Provided we have gradient access for
the black-box model \(M\) the solution to this problem can be found
through gradient descent. This generic framework lays the foundation for
most state-of-the-art approaches to counterfactual search and is also
used as the baseline approach in our package. The hyperparameter
\(\lambda\) is typically tuned through grid search or in some sense
pre-determined by the nature of the problem. Conventional choices for
\(\ell\) include margin-based losses like cross-entropy loss and hinge
loss. It is worth pointing out that the loss function is typically
computed with respect to logits rather than predicted probabilities, a
convention that we have chosen to follow.\footnote{Implementations of
  loss functions with respect to logits are often numerically more
  stable. For example, the \texttt{logitbinarycrossentropy(ŷ,\ y)}
  implementation in \texttt{Flux.Losses} (used here) is more stable than
  the mathematically equivalent \texttt{binarycrossentropy(ŷ,\ y)}.}

Numerous extensions to this simple approach have been developed since CE
were first proposed in 2017 (see Verma, Dickerson, and Hines (2020) and
(Karimi, Barthe, et al. 2020) for surveys). The various approaches
largely differ in that they use different flavours of search objective
defined in Equation~\ref{eq-solution}. Different penalties are often
used to address many of the desirable properties of effective CE that
have been set out. These desiderata include: \textbf{proximity} --- the
distance between factual and counterfactual features should be small
(Wachter, Mittelstadt, and Russell 2017); \textbf{actionability} --- the
proposed recourse should be actionable Poyiadzi et al. (2020);
\textbf{plausibility} --- the counterfactual explanation should be
plausible to a human Schut et al. (2021); \textbf{sparsity} --- the
counterfactual explanation should involve as few individual feature
changes as possible (Schut et al. 2021); \textbf{robustness} --- the
counterfactual explanation should be robust to domain and model shifts
(Upadhyay, Joshi, and Lakkaraju 2021); \textbf{diversity} --- ideally
multiple diverse counterfactuals should be provided (Mothilal, Sharma,
and Tan 2020); and \textbf{causality} --- counterfactuals should respect
the structural causal model underlying the data generating process
Karimi, Schölkopf, and Valera (2021).

Beyond gradient-based counterfactual search, which has been the main
focus in our development so far, various methodologies have been
proposed that can handle non-differentiable models like decision trees.
We have implemented some of these approaches and will discuss them
further in Section~\ref{sec-gen}.

\hypertarget{existing-software}{%
\subsection{Existing software}\label{existing-software}}

To the best of our knowledge, the package introduced here provides the
first implementation of Counterfactual Explanations in Julia and
therefore represents a novel contribution to the community. As for other
programming languages, we are only aware of one other unifying
framework: the Python library
\href{https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest}{CARLA}
(Pawelczyk et al. 2021).\footnote{While we were writing this paper, the
  \texttt{R} package \texttt{counterfactuals} was released (Dandl et al.
  2023). The developers seem to also envision a unifying framework, but
  the project appears to still be in its early stages.} In addition to
that, there exists open-source code for some specific approaches to CE
that have been proposed in recent years. The approach-specific
implementations that we have been able to find are generally
well-documented, but exclusively in Python. For example, a PyTorch
implementation of a greedy generator for Bayesian models proposed in
Schut et al. (2021) has been released. As another example, the popular
\href{https://github.com/interpretml}{InterpretML} library includes an
implementation of a diverse counterfactual generator (Mothilal, Sharma,
and Tan 2020).

Generally speaking, software development in the space of XAI has largely
focused on various global methods and surrogate explainers:
implementations of PDP, LIME and SHAP are available for both Python
(e.g.~\href{https://github.com/marcotcr/lime}{\texttt{lime}},
\href{https://github.com/slundberg/shap}{\texttt{shap}}) and R
(e.g.~\href{https://cran.r-project.org/web/packages/lime/index.html}{\texttt{lime}},
\href{https://cran.r-project.org/web/packages/lime/index.html}{\texttt{iml}},
\href{https://modeloriented.github.io/shapper/}{\texttt{shapper}},
\href{https://github.com/bgreenwell/fastshap}{\texttt{fastshap}}). In
the Julia space, there exist two packages related to XAI: firstly,
\href{https://github.com/nredell/ShapML.jl}{\texttt{ShapML.jl}}, which
provides a fast implementation of SHAP; and, secondly,
\href{https://github.com/adrhill/ExplainableAI.jl}{\texttt{ExplainableAI.jl}},
which enables users to easily visualise gradients and activation maps
for \texttt{Flux.jl} models. We also should not fail to mention the
comprehensive
\href{https://docs.interpretable.ai/stable/IAIBase/data/}{Interpretable
AI} infrastructure, which focuses exclusively on interpretable models.

Arguably the current availability of tools for explaining black-box
models in Julia is limited, but it appears that the community is
invested in changing that. The team behind \texttt{MLJ.jl}, for example,
recruited contributors for a project about both Interpretable and
Explainable AI in 2022.\footnote{For details, see the Google Summer of
  Code 2022 project proposal:
  \url{https://julialang.org/jsoc/gsoc/MLJ/\#interpretable_machine_learning_in_julia}.}
With our work on Counterfactual Explanations we hope to contribute to
these efforts. We think that because of its unique transparency the
Julia language naturally lends itself towards building Trustworthy AI
systems.

\hypertarget{sec-arch}{%
\section{\texorpdfstring{Introducing:
\texttt{CounterfactualExplanations.jl}}{Introducing: CounterfactualExplanations.jl}}\label{sec-arch}}

Figure~\ref{fig-arch} provides an overview of the package architecture.
It is built around two core modules that are designed to be as
extensible as possible through dispatch: 1) \texttt{Models} is concerned
with making any arbitrary model compatible with the package; 2)
\texttt{Generators} is used to implement counterfactual search
algorithms. The core function of the
package---\texttt{generate\_counterfactual}---uses an instance of type
\texttt{\textless{}:AbstractFittedModel} produced by the \texttt{Models}
module and an instance of type \texttt{\textless{}:AbstractGenerator}
produced by the \texttt{Generators} module. Relating this to the
methodology outlined in Section~\ref{sec-method}, the former instance
corresponds to the model \(M\), while the latter defines the rules for
the counterfactual search (Equation~\ref{eq-solution}).

\begin{figure}

{\centering \includegraphics[width=3.33333in,height=2.38095in]{www/pkg_architecture.png}

}

\caption{\label{fig-arch}High-level schematic overview of package
architecture. Modules are shown in red, structs in green and functions
in purple.}

\end{figure}

\hypertarget{models}{%
\subsection{Models}\label{models}}

The package currently offers native support for models built and trained
in \href{https://fluxml.ai/}{Flux} (Innes 2018) as well as a small
subset of models made available through
\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ} (Blaom
et al. 2020). While in general it is assumed that users resort to this
package to explain their pre-trained models, we provide a simple API
call to train the following
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/}{models}:

\begin{itemize}
\tightlist
\item
  Linear Classifier (Logistic Regression and Multinomial Logit)
\item
  Multi-Layer Perceptron (Deep Neural Network)
\item
  Deep Ensemble Lakshminarayanan, Pritzel, and Blundell (2016)
\item
  Decision Tree, Random Forest, Gradient Boosted Trees
\end{itemize}

As we demonstrate below, it is straightforward to extend the package
through custom models. Support for \texttt{torch} models trained in
Python or R is also available.\footnote{We are currently relying on
  \texttt{PythonCall.jl} and \texttt{RCall.jl} and this functionality is
  still somewhat brittle. Since this is more of an edge case, we may
  move this feature into its own package in the future.}

\hypertarget{sec-gen}{%
\subsection{Generators}\label{sec-gen}}

A large and growing number of counterfactual
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/}{generators}
have already been implemented in our package (Table~\ref{tbl-gen}). At a
high level, we distinguish generators in terms of their compatible model
types, their default search space, and their composability. All
``gradient-based'' generators are compatible with differentiable models,
e.g.~\texttt{Flux} and \texttt{torch}, while ``tree-based'' generators
are only applicable to models that involve decision trees. Concerning
the search space, it is possible to search counterfactuals in a
lower-dimensional latent embedding of the feature space that implicitly
encodes the data-generating process (DGP). To learn the latent
embedding, existing work has typically relied on generative models or
existing causal knowledge Karimi, Schölkopf, and Valera (2021). While
this notion is compatible with all of our gradient-based generators,
only some generators search a latent space by default. Finally,
composability implies that the given generator can be blended with any
other composable generator, which we discuss in
Section~\ref{sec-gen-comp}.

Beyond these broad technical distinctions, generators largely differ in
terms of how they address the various desiderata mentioned above:
\emph{ClapROAR} aims to preserve the classifier, i.e.~to generate
counterfactuals that are robust to endogenous model shifts (Altmeyer et
al. 2023); \emph{CLUE} searches plausible counterfactuals in the latent
embedding of a generative model by explicitly minimising predictive
entropy (Antorán et al. 2020); \emph{DiCE} is designed to generate
multiple, maximally diverse counterfactuals (Mothilal, Sharma, and Tan
2020); \emph{FeatureTweak} leverages the internals of decision trees to
search counterfactuals on a feature-by-feature basis, finding the
counterfactual that tweaks the features in the least costly way (Tolomei
et al. 2017); \emph{Gravitational} aims to generate plausible and robust
counterfactuals by minimising the distance to observed samples in the
target class (Altmeyer et al. 2023); \emph{Greedy} aims to generate
plausible counterfactuals by implicitly minimising predictive
uncertainty of Bayesian classifiers (Schut et al. 2021);
\emph{GrowingSpheres} is model-agnostic, relying solely on identifying
nearest neighbours of counterfactuals in the target class by gradually
increasing the search radius and then moving counterfactuals in that
direction(Laugel et al. 2017); \emph{PROBE} generates probabilistically
robust counterfactuals (Pawelczyk et al. 2022); \emph{REVISE} addresses
the need for plausibility by searching counterfactuals in the latent
embedding of a Variational Autoencoder (VAE) (Joshi et al. 2019);
\emph{Wachter} is the baseline approach that only penalises the distance
to the original sample (Wachter, Mittelstadt, and Russell 2017).

\onecolumn

\hypertarget{tbl-gen}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}@{}}
\caption{\label{tbl-gen}Overview of implemented counterfactual
generators.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ClaPROAR (Altmeyer et al. 2023) & gradient based & feature & yes \\
CLUE (Antorán et al. 2020) & gradient based & latent & yes \\
DiCE (Mothilal, Sharma, and Tan 2020) & gradient based & feature &
yes \\
FeatureTweak (Tolomei et al. 2017) & tree based & feature & no \\
Gravitational (Altmeyer et al. 2023) & gradient based & feature & yes \\
Greedy (Schut et al. 2021) & gradient based & feature & yes \\
GrowingSpheres (Laugel et al. 2017) & agnostic & feature & no \\
PROBE (Pawelczyk et al. 2022) & gradient based & feature & no \\
REVISE (Joshi et al. 2019) & gradient based & latent & yes \\
Wachter (Wachter, Mittelstadt, and Russell 2017) & gradient based &
feature & yes \\
\end{longtable}

\twocolumn

\hypertarget{data-catalogue}{%
\subsection{Data Catalogue}\label{data-catalogue}}

To allow researchers and practitioners to test and compare
counterfactual generators, the package ships with
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/data_catalogue/}{catalogues}
of pre-processed synthetic and real-world benchmark datasets from
different domains. Real-world datasets include:

\begin{itemize}
\tightlist
\item
  Adult Census (Barry Becker 1996)
\item
  California Housing (Pace and Barry 1997)
\item
  CIFAR10 (Krizhevsky 2009)
\item
  German Credit (Hoffman 1994)
\item
  Give Me Some Credit (Kaggle 2011)
\item
  MNIST (LeCun 1998) and Fashion MNIST (Xiao, Rasul, and Vollgraf 2017)
\item
  UCI defaultCredit (Yeh and Lien 2009)
\end{itemize}

Custom datasets can also be easily preprocessed as explained in the
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/data_preprocessing/}{documentation}.

\hypertarget{plotting}{%
\subsection{Plotting}\label{plotting}}

The package also extends common \texttt{Plots.jl} methods to facilitate
the visualization of results. Calling the generic \texttt{plot()} method
on an instance of type \texttt{\textless{}:CounterfactualExplanation},
for example, generates a plot visualizing the entire counterfactual path
in the feature space\footnote{For multi-dimensional input data, standard
  dimensionality reduction techniques are used to compress the data. In
  this case, the classifier's decision boundary is approximated through
  a Nearest Neighbour model. This is still somewhat experimental and
  will be improved in the future.}. We will see several examples of this
below.

\hypertarget{sec-use}{%
\section{Basic Usage}\label{sec-use}}

In the following, we begin our exploration of the package functionality
with a simple example. We then demonstrate how more advanced generators
can be easily composed and show how users can impose mutability
constraints on features. Finally, we also briefly explore the topics of
counterfactual evaluation and benchmarking.

\hypertarget{sec-simple}{%
\subsection{A Simple Generic Generator}\label{sec-simple}}

Code \ref{lst:simple} below provides a complete example demonstrating
how the framework presented in Section~\ref{sec-method} can be
implemented through our package. Using a synthetic data set with
linearly separable features we first fit a linear classifier (line
\ref{line:simple-class}). Next, we define the target class (line
\ref{line:simple-t}) and then draw a random sample from the other class
(line \ref{line:simple-x}). Finally, we instantiate a generic generator
(line \ref{line:simple-gen}) and run the counterfactual search (line
\ref{line:simple-search}). Figure~\ref{fig-binary} illustrates the
resulting counterfactual path in the two-dimensional feature space.
Features go through iterative perturbations until the desired confidence
level is reached as illustrated by the contour in the background, which
shows the softmax output for the target class.

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:simple}, caption={Standard workflow for generating counterfactuals.}] 
# Data and Classifier:
counterfactual_data = load_linearly_separable()
M = fit_model(counterfactual_data, :Linear) @\label{line:simple-class}@

# Factual and Target:
yhat = predict_label(M, counterfactual_data)
target = 2    # target label @\label{line:simple-t}@
candidates = findall(vec(yhat) .!= target)
chosen = rand(candidates)
x = select_factual(counterfactual_data, chosen) @\label{line:simple-x}@

# Counterfactual search:
generator = GenericGenerator() @\label{line:simple-gen}@
ce = generate_counterfactual(
    x, target, counterfactual_data, M, generator) @\label{line:simple-search}@
\end{lstlisting}

\begin{figure}

{\centering \includegraphics[width=3.33333in,height=2.5in]{www/ce_binary.png}

}

\caption{\label{fig-binary}Counterfactual path using generic
counterfactual generator for conventional binary classifier.}

\end{figure}

In this simple example, the generic generator produces a valid
counterfactual, since the decision boundary is crossed and the predicted
label is flipped. But the counterfactual is not plausible: it does not
appear to be generated by the same DGP as the observed data in the
target class. This is because the generic generator does not take into
account any of the desiderata mentioned in Section~\ref{sec-method},
except for the distance to the factual sample.

\hypertarget{sec-gen-comp}{%
\subsection{Composing Generators}\label{sec-gen-comp}}

To address these issues, we can leverage the ideas underlying some of
the more advanced counterfactual generators introduced above. In
particular, we will now show how easy it is to
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/generators/}{compose
custom generators} that blend different ideas through user-friendly
macros.

Suppose we wanted to address the desiderata of plausibility and
diversity. We could do this by blending ideas underlying the \emph{DiCE}
generator with the \emph{REVISE} generator. Formally, the corresponding
search objective would be defined as follows,

\begin{equation}\protect\hypertarget{eq-comp}{}{
\mathbf{Z}^\prime = \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^{L \times K}} \{  {\ell(M(f(\mathbf{Z}^\prime)),t)} + \lambda \cdot {\text{diversity}(f(\mathbf{Z}^\prime)) }  \} 
}\label{eq-comp}\end{equation}

where \(\mathbf{X}^\prime\) is an \(L\)-dimensional array of
counterfactuals,
\(f: \mathcal{Z}^{L \times K} \mapsto \mathcal{X}^{L \times D}\) is a
function that maps the \(L \times K\)-dimensional latent space
\(\mathcal{Z}\) to the \(L \times D\)-dimensional feature space
\(\mathcal{X}\) and \(\text{diversity}(\cdot)\) is the penalty proposed
by Mothilal, Sharma, and Tan (2020) that induces diverse sets of
counterfactuals. As in Equation~\ref{eq-solution}, \(\ell\) is the loss
function, \(M\) is the black-box model, \(t\) is the target class, and
\(\lambda\) is the strength of the penalty.

Code \ref{lst:composed} demonstrates how Equation~\ref{eq-comp} can be
seamlessly translated into Julia code. We begin by instantiating a
\texttt{GradientBasedGenerator} in line \ref{line:composed-init}. Next,
we use chained macros for composition: firstly, we define the
counterfactual search \texttt{@objective} corresponding to \emph{DiCE}
in line \ref{line:composed-dice}; secondly, we define the latent space
search strategy corresponding to \emph{REVISE} using the
\texttt{@search\_latent\_space} macro in line
\ref{line:composed-latent}; finally, we specify our prefered
optimisation method using the \texttt{@with\_optimiser} macro in line
\ref{line:composed-adam}.

\begin{lstlisting}[language=Julia, escapechar=§, numbers=left, label={lst:composed}, caption={Composing a custom generator.}]
generator = GradientBasedGenerator() §\label{line:composed-init}§
@chain generator begin
    @objective logitcrossentropy 
      + 0.2ddp_diversity §\label{line:composed-dice}§
    @search_latent_space §\label{line:composed-latent}§
    @with_optimiser Adam(0.005) §\label{line:composed-adam}§
end
\end{lstlisting}

In this case, the counterfactual search is performed in the latent space
of a Variational Autoencoder (VAE) that is automatically trained on the
observed data. It is important to specify the keyword argument
\texttt{num\_counterfactuals} of the \texttt{generate\_counterfactual}
to some value higher than \(1\) (default), to ensure that the diversity
penalty is effective. The resulting counterfactual path is shown in
Figure~\ref{fig-binary-advanced} below. We observe that the resulting
counterfactuals are diverse and the majority of them are plausible.

\begin{figure}

{\centering \includegraphics[width=3.33333in,height=2.5in]{www/binary_advanced.png}

}

\caption{\label{fig-binary-advanced}Counterfactual path using the
\emph{DiCE} generator.}

\end{figure}

\hypertarget{sec-mut}{%
\subsection{Mutability Constraints}\label{sec-mut}}

In practice, features usually cannot be perturbed arbitrarily. Suppose,
for example, that one of the features used by a bank to predict the
creditworthiness of its clients is \emph{age}. If a counterfactual
explanation for the prediction model indicates that older clients should
``grow younger'' to improve their creditworthiness, then this is an
interesting insight (it reveals age bias), but the provided recourse is
not actionable. In such cases, we may want to constrain the mutability
of features. To illustrate how this can be implemented in our package,
we will continue with the example from above.

Mutability can be defined in terms of four different options: 1) the
feature is mutable in both directions, 2) the feature can only increase
(e.g.~\emph{age}), 3) the feature can only decrease (e.g.~\emph{time
left} until your next deadline) and 4) the feature is not mutable
(e.g.~\emph{skin colour}, \emph{ethnicity}, \ldots). To specify which
category a feature belongs to, users can pass a vector of symbols
containing the mutability constraints at the pre-processing stage. For
each feature one can choose from these four options: \texttt{:both}
(mutable in both directions), \texttt{:increase} (only up),
\texttt{:decrease} (only down) and \texttt{:none} (immutable). By
default, \texttt{nothing} is passed to that keyword argument and it is
assumed that all features are mutable in both directions.\footnote{Mutability
  constraints are not yet implemented for Latent Space search.}

We can impose that the first feature is immutable as follows:
\texttt{counterfactual\_data.mutability\ =\ {[}:none,\ :both{]}}. The
resulting counterfactual path is shown in Figure~\ref{fig-mutability}
below. Since only the second feature can be perturbed, the sample can
only move along the vertical axis. In this case, the counterfactual
search does not yield a valid counterfactual, since the target class is
not reached.

\begin{figure}

{\centering \includegraphics[width=3.33333in,height=2.5in]{www/constraint_mutability.png}

}

\caption{\label{fig-mutability}Counterfactual path with immutable
feature.}

\end{figure}

\hypertarget{sec-eval}{%
\subsection{Evaluation and Benchmarking}\label{sec-eval}}

The package also makes it easy to
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/evaluation/}{evaluate}
counterfactuals with respect to many of the desiderata mentioned above.
For example, users may want to infer how costly the provided recourse is
to individuals. To this end, we can measure the distance of the
counterfactual from its original value. The API call to compute the
distance metric defined in Wachter, Mittelstadt, and Russell (2017), for
instance, is as simple as \texttt{evaluate(ce;\ measure=distance\_mad)},
where \texttt{ce} can also be a vector of
\texttt{CounterfactualExplanation}s.

Additionally, the package provides a
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/benchmarking/}{benchmarking}
framework that allows users to compare the performance of different
generators on a given dataset. In Figure~\ref{fig-bmk} we show the
results of a benchmark comparing several generators in terms of the
average cost and implausibility of the generated counterfactuals. The
cost is proxied by the L1-norm of the difference between the factual and
counterfactual features, while implausibility is measured by the
distance of the counterfactuals from samples in the target class. The
results illustrate that there is a tradeoff between minimizing costs to
individuals and generating plausible counterfactuals.

\begin{figure}

{\centering \includegraphics[width=3.33333in,height=2.5in]{www/bmk.png}

}

\caption{\label{fig-bmk}Benchmarking results for different generators.}

\end{figure}

\hypertarget{sec-custom}{%
\section{Customization and Extensibility}\label{sec-custom}}

One of our priorities has been to make our package customizable and
extensible. In the long term, we aim to add support for more default
models and counterfactual generators. In the short term, it is designed
to allow users to integrate models and generators themselves. These
community efforts will facilitate our long-term goals.

\hypertarget{sec-custom-mod}{%
\subsection{Adding Custom Models}\label{sec-custom-mod}}

At the high level, only two steps are necessary to make any supervised
learning model compatible with our package:

\begin{itemize}
\tightlist
\item
  \textbf{Subtyping}: We need to subtype the
  \texttt{AbstractFittedModel}.
\item
  \textbf{Dispatch}: The functions \texttt{logits} and \texttt{probs}
  need to be extended through custom methods for the model in question.
\end{itemize}

To demonstrate how this can be done in practice, we will reiterate here
how native support for \href{https://fluxml.ai/}{\texttt{Flux.jl}}
(Innes 2018) deep learning models was enabled.\footnote{Flux models are
  now natively supported by our package and can be instantiated by
  calling \texttt{FluxModel()}.} Once again we use synthetic data for an
illustrative example. Code \ref{lst:nn} below builds a simple model
architecture that can be used for a multi-class prediction task. Note
how outputs from the final layer are not passed through a softmax
activation function, since the counterfactual loss is evaluated with
respect to logits as we discussed earlier. The model is trained with
dropout.

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:nn}, caption={A simple neural network model.}]
n_hidden = 32
output_dim = length(unique(y))
input_dim = 2
model = Chain(
    Dense(input_dim, n_hidden, activation),
    Dropout(0.1),
    Dense(n_hidden, output_dim)
)  
\end{lstlisting}

Code \ref{lst:mymodel} below implements the two steps that were
necessary to make Flux models compatible with the package. In line
\ref{line:mymodel-subtype} we declare our new struct as a subtype of
\texttt{AbstractDifferentiableModel}, which itself is an abstract
subtype of \texttt{AbstractFittedModel}.\footnote{Note that in line
  \ref{line:mymodel-likelihood} we also provide a field determining the
  likelihood. This is optional and only used internally to determine
  which loss function to use in the counterfactual search. If this field
  is not provided to the model, the loss function needs to be explicitly
  supplied to the generator.} Computing logits amounts to just calling
the model on inputs. Predicted probabilities for labels can be computed
by passing logits through the softmax function.

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:mymodel}, caption={A wrapper for Flux models.}]
# Step 1)
struct MyFluxModel <: AbstractDifferentiableModel @\label{line:mymodel-subtype}@
    model::Any
    likelihood::Symbol @\label{line:mymodel-likelihood}@
end

# Step 2)
# import functions in order to extend
import CounterfactualExplanations.Models: logits
import CounterfactualExplanations.Models: probs 
logits(M::MyFluxModel, X::AbstractArray) = M.model(X)
probs(M::MyFluxModel, X::AbstractArray) = softmax(logits(M, X))
M = MyFluxModel(model)
\end{lstlisting}

The API call for generating counterfactuals for our new model is the
same as before. Figure~\ref{fig-multi} shows the resulting
counterfactual path for a randomly chosen sample. In this case, the
contour shows the predicted probability that the input is in the target
class (\(t=2\)).

\begin{figure}

{\centering \includegraphics[width=3.33333in,height=2.5in]{www/ce_multi.png}

}

\caption{\label{fig-multi}Counterfactual path using generic
counterfactual generator for multi-class classifier.}

\end{figure}

\hypertarget{sec-custom-gen}{%
\subsection{Adding Custom Generators}\label{sec-custom-gen}}

In some cases, composability may not be sufficient to implement specific
logics underlying certain counterfactual generators. In such cases,
users may want to implement custom generators. To illustrate how this
can be done we will consider a simple extension of our
\texttt{GenericGenerator}. As we have seen above, Counterfactual
Explanations are not unique. In light of this, we might be interested in
quantifying the uncertainty around the generated counterfactuals
(Delaney, Greene, and Keane 2021). One idea could be, to use dropout to
randomly switch features on and off in each iteration. Without dwelling
further on the merit of this idea, we will now briefly show how this can
be implemented.

\hypertarget{a-generator-with-dropout}{%
\subsubsection{A Generator with
Dropout}\label{a-generator-with-dropout}}

Code \ref{lst:dropout} below implements two important steps: 1) create
an abstract subtype of the \texttt{AbstractGradientBasedGenerator} and
2) create a constructor with an additional field for the dropout
probability.

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:dropout}, caption={Building a custom generator with dropout.}]
# Abstract suptype:
abstract type AbstractDropoutGenerator <: AbstractGradientBasedGenerator end
# Constructor:
struct DropoutGenerator <: AbstractDropoutGenerator
    loss::Symbol # loss function
    complexity::Function # complexity function
    @$\lambda$@::AbstractFloat # strength of penalty
    decision_threshold::Union{Nothing,AbstractFloat} 
    opt::Any # optimizer
    @$\tau$@::AbstractFloat # tolerance for convergence
    p_dropout::AbstractFloat # dropout rate
end
\end{lstlisting}

Next, in Code \ref{lst:generate} we define how feature perturbations are
generated for our custom dropout generator: in particular, we extend the
relevant function through a method that implements the dropout logic.

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:generate}, caption={Generating feature perturbations with dropout.}]
using CounterfactualExplanations.Generators
function Generators.generate_perturbations(
    generator::AbstractDropoutGenerator, 
    ce::CouterfactualExplanation
)
    @$s^\prime$@ = deepcopy(ce.@$s^\prime$@)
    new_@$s^\prime$@ = Generators.propose_state(
        generator, ce)
    @$\Delta s^\prime$@ = new_@$s^\prime$@ - @$s^\prime$@ # gradient step
    # Dropout:
    set_to_zero = sample(
        1:length(@$\Delta s^\prime$@),
        Int(round(generator.p_dropout*length(@$\Delta s^\prime$@))),
        replace=false
    )
    @$\Delta s^\prime$@[set_to_zero] .= 0
    return @$\Delta s^\prime$@
end
\end{lstlisting}

Finally, we proceed to generate counterfactuals in the same way we
always do. The resulting counterfactual path is shown in
Figure~\ref{fig-dropout}.

\begin{figure}

{\centering \includegraphics[width=3.33333in,height=2.5in]{www/dropout.png}

}

\caption{\label{fig-dropout}Counterfactual path for a generator with
dropout.}

\end{figure}

\hypertarget{sec-emp}{%
\section{A Real-World Examples}\label{sec-emp}}

Now that we have explained the basic functionality of
\texttt{CounterfactualExplanations.jl} through some synthetic examples,
it is time to work through examples involving real-world data.

\hypertarget{give-me-some-credit}{%
\subsection{Give Me Some Credit}\label{give-me-some-credit}}

The \emph{Give Me Some Credit} dataset is one of the tabular real-world
datasets that ship with the package (Kaggle 2011). It can be used to
train a binary classifier to predict whether a borrower is likely to
experience financial difficulties in the next two years. In particular,
we have an output variable
\(y \in \{0=\texttt{no stress},1=\texttt{stress}\}\) and a feature
matrix \(X\) that includes socio-demographic variables like \texttt{age}
and \texttt{income}. A retail bank might use such a classifier to
determine if potential borrowers should receive credit or not.

For the classification task, we use a Multi-Layer Perceptron with
dropout regularization. Using the Gravitational generator (Altmeyer et
al. 2023) we will generate counterfactuals for ten randomly chosen
individuals that would be denied credit based on our pre-trained model.
Concerning the mutability of features, we only impose that the
\texttt{age} cannot be decreased.

Figure~\ref{fig-credit} shows the resulting counterfactuals proposed by
Wachter in the two-dimensional feature space spanned by the \texttt{age}
and \texttt{income} variables. An increase in income and age is
recommended for the majority of individuals, which seems plausible: both
age and income are typically positively related to creditworthiness.

\begin{figure}

{\centering \includegraphics[width=3.33333in,height=1.33333in]{www/credit.png}

}

\caption{\label{fig-credit}Give Me Some Credit: counterfactuals for
would-be borrowers proposed by the Gravitational Generator.}

\end{figure}

\hypertarget{mnist}{%
\subsection{MNIST}\label{mnist}}

For our second example, we will look at image data. The MNIST dataset
contains 60,000 training samples of handwritten digits in the form of
28x28 pixel grey-scale images (LeCun 1998). Each image is associated
with a label indicating the digit (0-9) that the image represents. The
data makes for an interesting case study of CE because humans have a
good idea of what plausible counterfactuals of digits look like. For
example, if you were asked to pick up an eraser and turn the digit in
the left panel of Figure~\ref{fig-mnist} into a four (4) you would know
exactly what to do: just erase the top part.

On the model side, we will use a simple multi-layer perceptron (MLP).
Code \ref{lst:mnist-setup} loads the data and the pre-trained MLP. It
also loads two pre-trained Variational Auto-Encoders, which will be used
by our counterfactual generator of choice for this task: \emph{REVISE}.

\begin{lstlisting}[language=Julia, escapechar=@, numbers=left, label={lst:mnist-setup}, caption={Loading pre-trained models and data for MNIST.}]
counterfactual_data = load_mnist()
X, y = unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()
vae = load_mnist_vae()
vae_weak = load_mnist_vae(;strong=false)
\end{lstlisting}

The proposed counterfactuals are shown in Figure~\ref{fig-mnist}. In the
case in which \emph{REVISE} has access to an expressive VAE (centre),
the result looks convincing: the perturbed image does look like it
represents a four (4). In terms of explainability, we may conclude that
removing the top part of the handwritten nine (9) leads the black-box
model to predict that the perturbed image represents a four (4). We
should note, however, that the quality of counterfactuals produced by
\emph{REVISE} hinges on the performance of the underlying generative
model, as demonstrated by the result on the right. In this case,
\emph{REVISE} uses a weak VAE and the resulting counterfactual is
invalid. In light of this, we recommend using Latent Space search with
care.

\begin{figure}

{\centering \includegraphics[width=3.33333in,height=1.11111in]{www/mnist_9to4_latent.png}

}

\caption{\label{fig-mnist}Counterfactual explanations for MNIST using a
Latent Space generator: turning a nine (9) into a four (4).}

\end{figure}

\hypertarget{sec-outlook}{%
\section{Discussion and Outlook}\label{sec-outlook}}

We believe that this package in its current form offers a valuable
contribution to ongoing efforts towards XAI in Julia. That being said,
there is significant scope for future developments, which we briefly
outline in this final section.

\hypertarget{candidate-models-and-generators}{%
\subsection{Candidate models and
generators}\label{candidate-models-and-generators}}

The package supports various models and generators either natively or
through minimal augmentation. In future work, we would like to
prioritize the addition of further predictive models and generators.
Concerning the former, it would be useful to add native support for any
supervised models built in \texttt{MLJ.jl}, an extensive Machine
Learning framework for Julia (Blaom et al. 2020). This may also involve
adding support for regression models as well as additional
non-differentiable models. In terms of counterfactual generators, there
is a list of recent methodologies that we would like to implement
including MINT (Karimi, Schölkopf, and Valera 2021), ROAR (Upadhyay,
Joshi, and Lakkaraju 2021) and FACE (Poyiadzi et al. 2020).

\hypertarget{additional-datasets}{%
\subsection{Additional datasets}\label{additional-datasets}}

For benchmarking and testing purposes it will be crucial to add more
datasets to our library. We have so far prioritized tabular datasets
that have typically been used in the literature on counterfactual
explanations including \emph{Adult}, \emph{Give Me Some Credit} and
\emph{German Credit} (Karimi, Barthe, et al. 2020). There is scope for
adding data sources that have so far not been explored much in this
context including additional image datasets as well as audio, natural
language and time-series data.

\hypertarget{sec-conclude}{%
\section{Concluding remarks}\label{sec-conclude}}

\texttt{CounterfactualExplanation.jl} is a package for generating
Counterfactual Explanations and Algorithmic Recourse in Julia. Through
various synthetic and real-world examples, we have demonstrated the
basic usage of the package as well as its extensibility. The package has
already served us in our research to benchmark various methodological
approaches to Counterfactual Explanations and Algorithmic Recourse. We
therefore strongly believe that it should help other practitioners and
researchers in their own efforts towards Trustworthy AI.

We envision this package to one day constitute the go-to place for
explaining arbitrary predictive models through an extensive suite of
counterfactual generators. As a major next step, we aim to make our
library as compatible as possible with the popular
\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{\texttt{MLJ.jl}}
package for machine learning in Julia. We invite the Julia community to
contribute to these goals through usage, open challenge and active
development.

\hypertarget{sec-ack}{%
\section{Acknowledgements}\label{sec-ack}}

We are immensely grateful to the group of TU Delft students who
contributed huge improvements to this package as part of a university
project in 2023: Rauno Arike, Simon Kasdorp, Lauri Kesküll, Mariusz
Kicior, Vincent Pikand. We also want to thank the broader Julia
community for being welcoming and open and for supporting research
contributions like this one. Some of the members of TU Delft were
partially funded by ICAI AI for Fintech Research, an ING---TU Delft
collaboration.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2023endogenous}{}}%
Altmeyer, Patrick, Giovan Angela, Aleksander Buszydlik, Karol Dobiczek,
Arie van Deursen, and Cynthia Liem. 2023. {``Endogenous {Macrodynamics}
in {Algorithmic} {Recourse}.''} In \emph{First {IEEE} {Conference} on
{Secure} and {Trustworthy} {Machine} {Learning}}.
\url{https://doi.org/10.1109/satml54575.2023.00036}.

\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José
Miguel Hernández-Lobato. 2020. {``Getting a Clue: {A} Method for
Explaining Uncertainty Estimates.''}
\url{https://arxiv.org/abs/2006.06848}.

\leavevmode\vadjust pre{\hypertarget{ref-arrieta2020explainable}{}}%
Arrieta, Alejandro Barredo, Natalia Diaz-Rodriguez, Javier Del Ser,
Adrien Bennetot, Siham Tabik, Alberto Barbado, Salvador Garcia, et al.
2020. {``Explainable {Artificial Intelligence} ({XAI}): {Concepts},
Taxonomies, Opportunities and Challenges Toward Responsible {AI}.''}
\emph{Information Fusion} 58: 82--115.
\url{https://doi.org/10.1016/j.inffus.2019.12.012}.

\leavevmode\vadjust pre{\hypertarget{ref-becker1996adult}{}}%
Barry Becker, Ronny Kohavi. 1996. {``Adult.''} UCI Machine Learning
Repository. \url{https://doi.org/10.24432/C5XW20}.

\leavevmode\vadjust pre{\hypertarget{ref-blaom2020mlj}{}}%
Blaom, Anthony D., Franz Kiraly, Thibaut Lienart, Yiannis Simillides,
Diego Arenas, and Sebastian J. Vollmer. 2020. {``{MLJ}: {A Julia}
Package for Composable Machine Learning.''} \emph{Journal of Open Source
Software} 5 (55): 2704. \url{https://doi.org/10.21105/joss.02704}.

\leavevmode\vadjust pre{\hypertarget{ref-borch2022machine}{}}%
Borch, Christian. 2022. {``Machine Learning, Knowledge Risk, and
Principal-Agent Problems in Automated Trading.''} \emph{Technology in
Society}, 101852. \url{https://doi.org/10.1016/j.techsoc.2021.101852}.

\leavevmode\vadjust pre{\hypertarget{ref-buolamwini2018gender}{}}%
Buolamwini, Joy, and Timnit Gebru. 2018. {``Gender Shades:
{Intersectional} Accuracy Disparities in Commercial Gender
Classification.''} In \emph{Conference on Fairness, Accountability and
Transparency}, 77--91. {PMLR}.

\leavevmode\vadjust pre{\hypertarget{ref-dandl2023counterfactuals}{}}%
Dandl, Susanne, Andreas Hofheinz, Martin Binder, Bernd Bischl, and
Giuseppe Casalicchio. 2023. {``Counterfactuals: {An} {R} {Package} for
{Counterfactual} {Explanation} {Methods}.''} arXiv.
\url{http://arxiv.org/abs/2304.06569}.

\leavevmode\vadjust pre{\hypertarget{ref-delaney2021uncertainty}{}}%
Delaney, Eoin, Derek Greene, and Mark T. Keane. 2021. {``Uncertainty
{Estimation} and {Out}-of-{Distribution} {Detection} for
{Counterfactual} {Explanations}: {Pitfalls} and {Solutions}.''} arXiv.
\url{http://arxiv.org/abs/2107.09734}.

\leavevmode\vadjust pre{\hypertarget{ref-fan2020interpretability}{}}%
Fan, Fenglei, Jinjun Xiong, and Ge Wang. 2020. {``On Interpretability of
Artificial Neural Networks.''} \url{https://arxiv.org/abs/2001.02522}.

\leavevmode\vadjust pre{\hypertarget{ref-goodfellow2014explaining}{}}%
Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
{``Explaining and Harnessing Adversarial Examples.''}
\url{https://arxiv.org/abs/1412.6572}.

\leavevmode\vadjust pre{\hypertarget{ref-hoffman1994german}{}}%
Hoffman, Hans. 1994. {``German {Credit Data}.''}
\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}.

\leavevmode\vadjust pre{\hypertarget{ref-innes2018flux}{}}%
Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
\emph{Journal of Open Source Software} 3 (25): 602.
\url{https://doi.org/10.21105/joss.00602}.

\leavevmode\vadjust pre{\hypertarget{ref-joshi2019realistic}{}}%
Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
Actionable Explanations in Black-Box Decision Making Systems.''}
\url{https://arxiv.org/abs/1907.09615}.

\leavevmode\vadjust pre{\hypertarget{ref-kaggle2011give}{}}%
Kaggle. 2011. {``Give Me Some Credit, {Improve} on the State of the Art
in Credit Scoring by Predicting the Probability That Somebody Will
Experience Financial Distress in the Next Two Years.''} {Kaggle}.
\url{https://www.kaggle.com/c/GiveMeSomeCredit}.

\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
Karimi, Amir-Hossein, Gilles Barthe, Bernhard Schölkopf, and Isabel
Valera. 2020. {``A Survey of Algorithmic Recourse: Definitions,
Formulations, Solutions, and Prospects.''}
\url{https://arxiv.org/abs/2010.04050}.

\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
Karimi, Amir-Hossein, Bernhard Schölkopf, and Isabel Valera. 2021.
{``Algorithmic Recourse: From Counterfactual Explanations to
Interventions.''} In \emph{Proceedings of the 2021 {ACM Conference} on
{Fairness}, {Accountability}, and {Transparency}}, 353--62.

\leavevmode\vadjust pre{\hypertarget{ref-karimi2020algorithmic}{}}%
Karimi, Amir-Hossein, Julius Von Kügelgen, Bernhard Schölkopf, and
Isabel Valera. 2020. {``Algorithmic Recourse Under Imperfect Causal
Knowledge: A Probabilistic Approach.''}
\url{https://arxiv.org/abs/2006.06831}.

\leavevmode\vadjust pre{\hypertarget{ref-kaur2020interpreting}{}}%
Kaur, Harmanpreet, Harsha Nori, Samuel Jenkins, Rich Caruana, Hanna
Wallach, and Jennifer Wortman Vaughan. 2020. {``Interpreting
Interpretability: Understanding Data Scientists' Use of Interpretability
Tools for Machine Learning.''} In \emph{Proceedings of the 2020 {CHI}
Conference on Human Factors in Computing Systems}, 1--14.
\url{https://doi.org/10.1145/3313831.3376219}.

\leavevmode\vadjust pre{\hypertarget{ref-krizhevsky2009learning}{}}%
Krizhevsky, A. 2009. {``Learning {Multiple} {Layers} of {Features} from
{Tiny} {Images}.''} In.
\url{https://www.semanticscholar.org/paper/Learning-Multiple-Layers-of-Features-from-Tiny-Krizhevsky/5d90f06bb70a0a3dced62413346235c02b1aa086}.

\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2016.
{``Simple and Scalable Predictive Uncertainty Estimation Using Deep
Ensembles.''} \url{https://arxiv.org/abs/1612.01474}.

\leavevmode\vadjust pre{\hypertarget{ref-laugel2017inversea}{}}%
Laugel, Thibault, Marie-Jeanne Lesot, Christophe Marsala, Xavier Renard,
and Marcin Detyniecki. 2017. {``Inverse {Classification} for
{Comparison}-Based {Interpretability} in {Machine} {Learning}.''} arXiv.
\url{https://doi.org/10.48550/arXiv.1712.08443}.

\leavevmode\vadjust pre{\hypertarget{ref-lecun1998mnist}{}}%
LeCun, Yann. 1998. {``The {MNIST} Database of Handwritten Digits.''}

\leavevmode\vadjust pre{\hypertarget{ref-miller2019explanation}{}}%
Miller, Tim. 2019. {``Explanation in Artificial Intelligence: {Insights}
from the Social Sciences.''} \emph{Artificial Intelligence} 267: 1--38.
\url{https://doi.org/10.1016/j.artint.2018.07.007}.

\leavevmode\vadjust pre{\hypertarget{ref-molnar2020interpretable}{}}%
Molnar, Christoph. 2020. \emph{Interpretable Machine Learning}. {Lulu.
com}.

\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
Mothilal, Ramaravind K, Amit Sharma, and Chenhao Tan. 2020.
{``Explaining Machine Learning Classifiers Through Diverse
Counterfactual Explanations.''} In \emph{Proceedings of the 2020
{Conference} on {Fairness}, {Accountability}, and {Transparency}},
607--17. \url{https://doi.org/10.1145/3351095.3372850}.

\leavevmode\vadjust pre{\hypertarget{ref-oneil2016weapons}{}}%
O'Neil, Cathy. 2016. \emph{Weapons of Math Destruction: {How} Big Data
Increases Inequality and Threatens Democracy}. {Crown}.

\leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
Pace, R Kelley, and Ronald Barry. 1997. {``Sparse Spatial
Autoregressions.''} \emph{Statistics \& Probability Letters} 33 (3):
291--97. \url{https://doi.org/10.1016/s0167-7152(96)00140-x}.

\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
Pawelczyk, Martin, Sascha Bielawski, Johannes van den Heuvel, Tobias
Richter, and Gjergji Kasneci. 2021. {``Carla: A Python Library to
Benchmark Algorithmic Recourse and Counterfactual Explanation
Algorithms.''} \url{https://arxiv.org/abs/2108.00783}.

\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2022probabilistically}{}}%
Pawelczyk, Martin, Teresa Datta, Johannes van-den-Heuvel, Gjergji
Kasneci, and Himabindu Lakkaraju. 2022. {``Probabilistically {Robust}
{Recourse}: {Navigating} the {Trade}-Offs Between {Costs} and
{Robustness} in {Algorithmic} {Recourse}.''} \emph{arXiv Preprint
arXiv:2203.06768}.

\leavevmode\vadjust pre{\hypertarget{ref-poyiadzi2020face}{}}%
Poyiadzi, Rafael, Kacper Sokol, Raul Santos-Rodriguez, Tijl De Bie, and
Peter Flach. 2020. {``{FACE}: {Feasible} and Actionable Counterfactual
Explanations.''} In \emph{Proceedings of the {AAAI}/{ACM Conference} on
{AI}, {Ethics}, and {Society}}, 344--50.

\leavevmode\vadjust pre{\hypertarget{ref-rudin2019stop}{}}%
Rudin, Cynthia. 2019. {``Stop Explaining Black Box Machine Learning
Models for High Stakes Decisions and Use Interpretable Models
Instead.''} \emph{Nature Machine Intelligence} 1 (5): 206--15.
\url{https://doi.org/10.1038/s42256-019-0048-x}.

\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
Sacaleanu, Yarin Gal, et al. 2021. {``Generating {Interpretable
Counterfactual Explanations By Implicit Minimisation} of {Epistemic} and
{Aleatoric Uncertainties}.''} In \emph{International {Conference} on
{Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.

\leavevmode\vadjust pre{\hypertarget{ref-slack2021counterfactual}{}}%
Slack, Dylan, Anna Hilgard, Himabindu Lakkaraju, and Sameer Singh. 2021.
{``Counterfactual Explanations Can Be Manipulated.''} \emph{Advances in
Neural Information Processing Systems} 34.

\leavevmode\vadjust pre{\hypertarget{ref-slack2020fooling}{}}%
Slack, Dylan, Sophie Hilgard, Emily Jia, Sameer Singh, and Himabindu
Lakkaraju. 2020. {``Fooling Lime and Shap: {Adversarial} Attacks on Post
Hoc Explanation Methods.''} In \emph{Proceedings of the {AAAI}/{ACM
Conference} on {AI}, {Ethics}, and {Society}}, 180--86.

\leavevmode\vadjust pre{\hypertarget{ref-tolomei2017interpretable}{}}%
Tolomei, Gabriele, Fabrizio Silvestri, Andrew Haines, and Mounia Lalmas.
2017. {``Interpretable {Predictions} of {Tree}-Based {Ensembles} via
{Actionable} {Feature} {Tweaking}.''} In \emph{Proceedings of the 23rd
{ACM} {SIGKDD} {International} {Conference} on {Knowledge} {Discovery}
and {Data} {Mining}}, 465--74.
\url{https://doi.org/10.1145/3097983.3098039}.

\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021robust}{}}%
Upadhyay, Sohini, Shalmali Joshi, and Himabindu Lakkaraju. 2021.
{``Towards {Robust} and {Reliable Algorithmic Recourse}.''}
\url{https://arxiv.org/abs/2102.13620}.

\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
Ustun, Berk, Alexander Spangher, and Yang Liu. 2019. {``Actionable
Recourse in Linear Classification.''} In \emph{Proceedings of the
{Conference} on {Fairness}, {Accountability}, and {Transparency}},
10--19. \url{https://doi.org/10.1145/3287560.3287566}.

\leavevmode\vadjust pre{\hypertarget{ref-varshney2022trustworthy}{}}%
Varshney, Kush R. 2022. \emph{Trustworthy {Machine Learning}}.
{Chappaqua, NY, USA}: {Independently Published}.

\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
Verma, Sahil, John Dickerson, and Keegan Hines. 2020. {``Counterfactual
Explanations for Machine Learning: {A} Review.''}
\url{https://arxiv.org/abs/2010.10596}.

\leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
{``Counterfactual Explanations Without Opening the Black Box:
{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
841. \url{https://doi.org/10.2139/ssrn.3063289}.

\leavevmode\vadjust pre{\hypertarget{ref-xiao2017fashion}{}}%
Xiao, Han, Kashif Rasul, and Roland Vollgraf. 2017. {``Fashion-{MNIST}:
A {Novel} {Image} {Dataset} for {Benchmarking} {Machine} {Learning}
{Algorithms}.''} arXiv. \url{https://doi.org/10.48550/arXiv.1708.07747}.

\leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
Yeh, I-Cheng, and Che-hui Lien. 2009. {``The Comparisons of Data Mining
Techniques for the Predictive Accuracy of Probability of Default of
Credit Card Clients.''} \emph{Expert Systems with Applications} 36 (2):
2473--80. \url{https://doi.org/10.1016/j.eswa.2007.12.020}.

\end{CSLReferences}

\end{document}

### 2023-11-16 examples/CounterfactualExplanations.jl/paper.tex
ok

@@
 
+% Keywords command
+\providecommand{\JCONkeywords}[1]
+{
+  \small	
+  \section*{Keywords} #1
+}
 \usepackage{orcidlink}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, frame hidden, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
 

### 2023-11-16 examples/CounterfactualExplanations.jl/paper.tex
suppressing line 142 in cls related to subcaption

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 

### 2023-11-16 examples/CounterfactualExplanations.jl/paper.tex
comment out lines 736 to 747

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 

### 2023-11-16 examples/CounterfactualExplanations.jl/paper.tex
this seems to work

@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \definecolor{mypink}{RGB}{219, 48, 122}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
 \makeatletter
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \makeatother
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi
-
 \setstretch{1}
-\hypertarget{sec-intro}{%
-\section{Introduction}\label{sec-intro}}
+\section{Introduction}\label{sec-intro}
 
@@
 
-\hypertarget{sec-related}{%
-\section{Background and related work}\label{sec-related}}
+\section{Background and related work}\label{sec-related}
 
@@
 
-\hypertarget{literature-on-explainable-ai}{%
 \subsection{Literature on Explainable
-AI}\label{literature-on-explainable-ai}}
+AI}\label{literature-on-explainable-ai}
 
@@
 
-\hypertarget{sec-method}{%
 \subsection{A framework for Counterfactual
-Explanations}\label{sec-method}}
+Explanations}\label{sec-method}
 
@@
 
-\begin{equation}\protect\hypertarget{eq-obj}{}{
+\begin{equation}\phantomsection\label{eq-obj}{
 \min_{x^\prime \in \mathcal{X}} h(x^\prime) \ \ \ \mbox{s. t.} \ \ \ M(x^\prime) = t
-}\label{eq-obj}\end{equation}
+}\end{equation}
 
@@
 
-\begin{equation}\protect\hypertarget{eq-solution}{}{
+\begin{equation}\phantomsection\label{eq-solution}{
 x^\prime = \arg \min_{x^\prime}  \ell(M(x^\prime),t) + \lambda h(x^\prime)
-}\label{eq-solution}\end{equation}
+}\end{equation}
 
@@
 
-\hypertarget{existing-software}{%
-\subsection{Existing software}\label{existing-software}}
+\subsection{Existing software}\label{existing-software}
 
@@
 
-\hypertarget{sec-arch}{%
 \section{\texorpdfstring{Introducing:
-\texttt{CounterfactualExplanations.jl}}{Introducing: CounterfactualExplanations.jl}}\label{sec-arch}}
+\texttt{CounterfactualExplanations.jl}}{Introducing: CounterfactualExplanations.jl}}\label{sec-arch}
 
@@
 
-{\centering \includegraphics[width=3.33333in,height=2.38095in]{www/pkg_architecture.png}
+\centering{
+
+\includegraphics[width=3.33333in,height=2.38095in]{www/pkg_architecture.png}
 
@@
 
-\end{figure}
-
-\hypertarget{models}{%
-\subsection{Models}\label{models}}
+\end{figure}%
+
+\subsection{Models}\label{models}
 
@@
 
-\hypertarget{sec-gen}{%
-\subsection{Generators}\label{sec-gen}}
+\subsection{Generators}\label{sec-gen}
 
@@
 
-\hypertarget{tbl-gen}{}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}@{}}
-\caption{\label{tbl-gen}Overview of implemented counterfactual
-generators.}\tabularnewline
+\caption{Overview of implemented counterfactual
+generators.}\label{tbl-gen}\tabularnewline
 \toprule\noalign{}
@@
 
-\hypertarget{data-catalogue}{%
-\subsection{Data Catalogue}\label{data-catalogue}}
+\subsection{Data Catalogue}\label{data-catalogue}
 
@@
 
-\hypertarget{plotting}{%
-\subsection{Plotting}\label{plotting}}
+\subsection{Plotting}\label{plotting}
 
@@
 
-\hypertarget{sec-use}{%
-\section{Basic Usage}\label{sec-use}}
+\section{Basic Usage}\label{sec-use}
 
@@
 
-\hypertarget{sec-simple}{%
-\subsection{A Simple Generic Generator}\label{sec-simple}}
+\subsection{A Simple Generic Generator}\label{sec-simple}
 
@@
 
-{\centering \includegraphics[width=3.33333in,height=2.5in]{www/ce_binary.png}
+\centering{
+
+\includegraphics[width=3.33333in,height=2.5in]{www/ce_binary.png}
 
@@
 
-\end{figure}
+\end{figure}%
 
@@
 
-\hypertarget{sec-gen-comp}{%
-\subsection{Composing Generators}\label{sec-gen-comp}}
+\subsection{Composing Generators}\label{sec-gen-comp}
 
@@
 
-\begin{equation}\protect\hypertarget{eq-comp}{}{
+\begin{equation}\phantomsection\label{eq-comp}{
 \mathbf{Z}^\prime = \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^{L \times K}} \{  {\ell(M(f(\mathbf{Z}^\prime)),t)} + \lambda \cdot {\text{diversity}(f(\mathbf{Z}^\prime)) }  \} 
-}\label{eq-comp}\end{equation}
+}\end{equation}
 
@@
 
-{\centering \includegraphics[width=3.33333in,height=2.5in]{www/binary_advanced.png}
+\centering{
+
+\includegraphics[width=3.33333in,height=2.5in]{www/binary_advanced.png}
 
@@
 
-\end{figure}
-
-\hypertarget{sec-mut}{%
-\subsection{Mutability Constraints}\label{sec-mut}}
+\end{figure}%
+
+\subsection{Mutability Constraints}\label{sec-mut}
 
@@
 
-{\centering \includegraphics[width=3.33333in,height=2.5in]{www/constraint_mutability.png}
+\centering{
+
+\includegraphics[width=3.33333in,height=2.5in]{www/constraint_mutability.png}
 
@@
 
-\end{figure}
-
-\hypertarget{sec-eval}{%
-\subsection{Evaluation and Benchmarking}\label{sec-eval}}
+\end{figure}%
+
+\subsection{Evaluation and Benchmarking}\label{sec-eval}
 
@@
 
-{\centering \includegraphics[width=3.33333in,height=2.5in]{www/bmk.png}
+\centering{
+
+\includegraphics[width=3.33333in,height=2.5in]{www/bmk.png}
 
@@
 
-\end{figure}
-
-\hypertarget{sec-custom}{%
-\section{Customization and Extensibility}\label{sec-custom}}
+\end{figure}%
+
+\section{Customization and Extensibility}\label{sec-custom}
 
@@
 
-\hypertarget{sec-custom-mod}{%
-\subsection{Adding Custom Models}\label{sec-custom-mod}}
+\subsection{Adding Custom Models}\label{sec-custom-mod}
 
@@
 
-{\centering \includegraphics[width=3.33333in,height=2.5in]{www/ce_multi.png}
+\centering{
+
+\includegraphics[width=3.33333in,height=2.5in]{www/ce_multi.png}
 
@@
 
-\end{figure}
-
-\hypertarget{sec-custom-gen}{%
-\subsection{Adding Custom Generators}\label{sec-custom-gen}}
+\end{figure}%
+
+\subsection{Adding Custom Generators}\label{sec-custom-gen}
 
@@
 
-\hypertarget{a-generator-with-dropout}{%
-\subsubsection{A Generator with
-Dropout}\label{a-generator-with-dropout}}
+\subsubsection{A Generator with Dropout}\label{a-generator-with-dropout}
 
@@
 
-{\centering \includegraphics[width=3.33333in,height=2.5in]{www/dropout.png}
+\centering{
+
+\includegraphics[width=3.33333in,height=2.5in]{www/dropout.png}
 
@@
 
-\end{figure}
-
-\hypertarget{sec-emp}{%
-\section{A Real-World Examples}\label{sec-emp}}
+\end{figure}%
+
+\section{A Real-World Examples}\label{sec-emp}
 
@@
 
-\hypertarget{give-me-some-credit}{%
-\subsection{Give Me Some Credit}\label{give-me-some-credit}}
+\subsection{Give Me Some Credit}\label{give-me-some-credit}
 
@@
 
-{\centering \includegraphics[width=3.33333in,height=1.33333in]{www/credit.png}
+\centering{
+
+\includegraphics[width=3.33333in,height=1.33333in]{www/credit.png}
 
@@
 
-\end{figure}
-
-\hypertarget{mnist}{%
-\subsection{MNIST}\label{mnist}}
+\end{figure}%
+
+\subsection{MNIST}\label{mnist}
 
@@
 
-{\centering \includegraphics[width=3.33333in,height=1.11111in]{www/mnist_9to4_latent.png}
+\centering{
+
+\includegraphics[width=3.33333in,height=1.11111in]{www/mnist_9to4_latent.png}
 
@@
 
-\end{figure}
-
-\hypertarget{sec-outlook}{%
-\section{Discussion and Outlook}\label{sec-outlook}}
+\end{figure}%
+
+\section{Discussion and Outlook}\label{sec-outlook}
 
@@
 
-\hypertarget{candidate-models-and-generators}{%
 \subsection{Candidate models and
-generators}\label{candidate-models-and-generators}}
+generators}\label{candidate-models-and-generators}
 
@@
 
-\hypertarget{additional-datasets}{%
-\subsection{Additional datasets}\label{additional-datasets}}
+\subsection{Additional datasets}\label{additional-datasets}
 
@@
 
-\hypertarget{sec-conclude}{%
-\section{Concluding remarks}\label{sec-conclude}}
+\section{Concluding remarks}\label{sec-conclude}
 
@@
 
-\hypertarget{sec-ack}{%
-\section{Acknowledgements}\label{sec-ack}}
+\section{Acknowledgements}\label{sec-ack}
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2023endogenous}{}}%
+\bibitem[\citeproctext]{ref-altmeyer2023endogenous}
 Altmeyer, Patrick, Giovan Angela, Aleksander Buszydlik, Karol Dobiczek,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
+\bibitem[\citeproctext]{ref-antoran2020getting}
 Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-arrieta2020explainable}{}}%
+\bibitem[\citeproctext]{ref-arrieta2020explainable}
 Arrieta, Alejandro Barredo, Natalia Diaz-Rodriguez, Javier Del Ser,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-becker1996adult}{}}%
+\bibitem[\citeproctext]{ref-becker1996adult}
 Barry Becker, Ronny Kohavi. 1996. {``Adult.''} UCI Machine Learning
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-blaom2020mlj}{}}%
+\bibitem[\citeproctext]{ref-blaom2020mlj}
 Blaom, Anthony D., Franz Kiraly, Thibaut Lienart, Yiannis Simillides,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-borch2022machine}{}}%
+\bibitem[\citeproctext]{ref-borch2022machine}
 Borch, Christian. 2022. {``Machine Learning, Knowledge Risk, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-buolamwini2018gender}{}}%
+\bibitem[\citeproctext]{ref-buolamwini2018gender}
 Buolamwini, Joy, and Timnit Gebru. 2018. {``Gender Shades:
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-dandl2023counterfactuals}{}}%
+\bibitem[\citeproctext]{ref-dandl2023counterfactuals}
 Dandl, Susanne, Andreas Hofheinz, Martin Binder, Bernd Bischl, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-delaney2021uncertainty}{}}%
+\bibitem[\citeproctext]{ref-delaney2021uncertainty}
 Delaney, Eoin, Derek Greene, and Mark T. Keane. 2021. {``Uncertainty
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fan2020interpretability}{}}%
+\bibitem[\citeproctext]{ref-fan2020interpretability}
 Fan, Fenglei, Jinjun Xiong, and Ge Wang. 2020. {``On Interpretability of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-goodfellow2014explaining}{}}%
+\bibitem[\citeproctext]{ref-goodfellow2014explaining}
 Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hoffman1994german}{}}%
+\bibitem[\citeproctext]{ref-hoffman1994german}
 Hoffman, Hans. 1994. {``German {Credit Data}.''}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-innes2018flux}{}}%
+\bibitem[\citeproctext]{ref-innes2018flux}
 Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-joshi2019realistic}{}}%
+\bibitem[\citeproctext]{ref-joshi2019realistic}
 Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kaggle2011give}{}}%
+\bibitem[\citeproctext]{ref-kaggle2011give}
 Kaggle. 2011. {``Give Me Some Credit, {Improve} on the State of the Art
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
+\bibitem[\citeproctext]{ref-karimi2020survey}
 Karimi, Amir-Hossein, Gilles Barthe, Bernhard Schölkopf, and Isabel
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
+\bibitem[\citeproctext]{ref-karimi2021algorithmic}
 Karimi, Amir-Hossein, Bernhard Schölkopf, and Isabel Valera. 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2020algorithmic}{}}%
+\bibitem[\citeproctext]{ref-karimi2020algorithmic}
 Karimi, Amir-Hossein, Julius Von Kügelgen, Bernhard Schölkopf, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kaur2020interpreting}{}}%
+\bibitem[\citeproctext]{ref-kaur2020interpreting}
 Kaur, Harmanpreet, Harsha Nori, Samuel Jenkins, Rich Caruana, Hanna
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-krizhevsky2009learning}{}}%
+\bibitem[\citeproctext]{ref-krizhevsky2009learning}
 Krizhevsky, A. 2009. {``Learning {Multiple} {Layers} of {Features} from
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
+\bibitem[\citeproctext]{ref-lakshminarayanan2016simple}
 Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2016.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-laugel2017inversea}{}}%
+\bibitem[\citeproctext]{ref-laugel2017inversea}
 Laugel, Thibault, Marie-Jeanne Lesot, Christophe Marsala, Xavier Renard,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lecun1998mnist}{}}%
+\bibitem[\citeproctext]{ref-lecun1998mnist}
 LeCun, Yann. 1998. {``The {MNIST} Database of Handwritten Digits.''}
 
-\leavevmode\vadjust pre{\hypertarget{ref-miller2019explanation}{}}%
+\bibitem[\citeproctext]{ref-miller2019explanation}
 Miller, Tim. 2019. {``Explanation in Artificial Intelligence: {Insights}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-molnar2020interpretable}{}}%
+\bibitem[\citeproctext]{ref-molnar2020interpretable}
 Molnar, Christoph. 2020. \emph{Interpretable Machine Learning}. {Lulu.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
+\bibitem[\citeproctext]{ref-mothilal2020explaining}
 Mothilal, Ramaravind K, Amit Sharma, and Chenhao Tan. 2020.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-oneil2016weapons}{}}%
+\bibitem[\citeproctext]{ref-oneil2016weapons}
 O'Neil, Cathy. 2016. \emph{Weapons of Math Destruction: {How} Big Data
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
+\bibitem[\citeproctext]{ref-pace1997sparse}
 Pace, R Kelley, and Ronald Barry. 1997. {``Sparse Spatial
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
+\bibitem[\citeproctext]{ref-pawelczyk2021carla}
 Pawelczyk, Martin, Sascha Bielawski, Johannes van den Heuvel, Tobias
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2022probabilistically}{}}%
+\bibitem[\citeproctext]{ref-pawelczyk2022probabilistically}
 Pawelczyk, Martin, Teresa Datta, Johannes van-den-Heuvel, Gjergji
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-poyiadzi2020face}{}}%
+\bibitem[\citeproctext]{ref-poyiadzi2020face}
 Poyiadzi, Rafael, Kacper Sokol, Raul Santos-Rodriguez, Tijl De Bie, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rudin2019stop}{}}%
+\bibitem[\citeproctext]{ref-rudin2019stop}
 Rudin, Cynthia. 2019. {``Stop Explaining Black Box Machine Learning
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
+\bibitem[\citeproctext]{ref-schut2021generating}
 Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-slack2021counterfactual}{}}%
+\bibitem[\citeproctext]{ref-slack2021counterfactual}
 Slack, Dylan, Anna Hilgard, Himabindu Lakkaraju, and Sameer Singh. 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-slack2020fooling}{}}%
+\bibitem[\citeproctext]{ref-slack2020fooling}
 Slack, Dylan, Sophie Hilgard, Emily Jia, Sameer Singh, and Himabindu
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tolomei2017interpretable}{}}%
+\bibitem[\citeproctext]{ref-tolomei2017interpretable}
 Tolomei, Gabriele, Fabrizio Silvestri, Andrew Haines, and Mounia Lalmas.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021robust}{}}%
+\bibitem[\citeproctext]{ref-upadhyay2021robust}
 Upadhyay, Sohini, Shalmali Joshi, and Himabindu Lakkaraju. 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
+\bibitem[\citeproctext]{ref-ustun2019actionable}
 Ustun, Berk, Alexander Spangher, and Yang Liu. 2019. {``Actionable
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-varshney2022trustworthy}{}}%
+\bibitem[\citeproctext]{ref-varshney2022trustworthy}
 Varshney, Kush R. 2022. \emph{Trustworthy {Machine Learning}}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
+\bibitem[\citeproctext]{ref-verma2020counterfactual}
 Verma, Sahil, John Dickerson, and Keegan Hines. 2020. {``Counterfactual
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
+\bibitem[\citeproctext]{ref-wachter2017counterfactual}
 Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-xiao2017fashion}{}}%
+\bibitem[\citeproctext]{ref-xiao2017fashion}
 Xiao, Han, Kashif Rasul, and Roland Vollgraf. 2017. {``Fashion-{MNIST}:
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
+\bibitem[\citeproctext]{ref-yeh2009comparisons}
 Yeh, I-Cheng, and Che-hui Lien. 2009. {``The Comparisons of Data Mining

### 2023-11-16 template.qmd
this seems to work

@@
 ---
-title: Juliacon Proceedings in Quarto
+title: JuliaCon Proceedings in Quarto
 format:
@@
 
-## About this template
+# About this template
 
 This template is based on the existing [JuliaCon Proceedings LaTeX template](https://github.com/JuliaCon/JuliaConSubmission.jl).
+The rendered versions can be found here: [pdf](https://www.paltmeyer.com/quarto-juliacon-proceedings/template.pdf), [html](https://www.paltmeyer.com/quarto-juliacon-proceedings/template.html).
 
-## How to use this template
+## Basic Usage
 
-> The remainder of this document repeats relevant elements from the [JuliaCon Proceedings LaTeX template].
+To use the Quarto [extension](https://github.com/pat-alt/quarto-juliacon-proceedings) that provides this template, you can create a new project as follows:
 
-# The JuliaCon Article Class {#sec-documentclass}
+```
+quarto use template pat-alt/quarto-juliacon-proceedings
+```
 
-The `juliacon` class file preserves the standard LATEX{} interface such that any document that can be produced using the standard LATEX{} article class can also be produced with the class file.
+Alternatively, you can add the extension to an existing project:
 
-It is likely that the make up will change after file submission.
-For this reason, we ask you to ignore details such as slightly long lines, page stretching, or figures falling out of synchronization, as these details can be dealt with at a later stage.
+```
+quarto add pat-alt/quarto-juliacon-proceedings
+```
 
-Use should be made of symbolic references (\verb|\ref| ) in order to protect against late changes of order, etc.
+Then, add the format to your document options:
 
-# Executable Code
+```yaml
+format:
+  juliacon-proceedings-pdf: default
+```
 
-Ideally, we would be able to tab into Quarto's existing support for executable, cross-referenceable code chunks.
+# Code
+
+There are various options to present code using this template.
+We could leverage Quarto's existing support for executable, cross-referenceable code.
+Alternatively, we could use the existing `lstlisting` environment for Julia code.
+
+## Executable Listings  (version $\geq$ 1.4.0)
+
+Relying on Quarto's support for executable listings is the most flexible option as it allows us to use all of Quarto's features. It comes with numerous advantages:
+
+1. We can use the same code chunks for both PDF and HTML output (and more) with consistent formatting.
+2.
+Using executable code ensures that things stay up-to-date and potential errors in the code are identified early: as you render your document, the code will be executed and the output will be inserted into the document.
+3.
+The output of executable code can be cross-referenced.
+For example, we can reference a figure that is generated by a code chunk and the reference will be updated automatically when the figure number changes.
+4. Code chunks can also be hidden. In HTML, there is additional support for code folding.
+5. The code itself can be cross-referenced.
+
+From the [Quarto Docs](https://quarto.org/docs/prerelease/1.4/crossref.html#cross-referenceable-listings-of-executable-code-blocks):
 
@@
 For example, the following code block will be labeled as `lst-1` and calling `@lst-1` will render as @lst-1.
-This ensures that things stay up-to-date and potential errors in the code are identified early: as you render your document, the code will be executed and the output will be inserted into the document.
+The figure it generates will be labeled as `fig-1` and calling `@fig-1` will render as @fig-1.
 
@@
 #| lst-label: lst-1
-#| lst-cap: A listing caption
+#| lst-cap: A listing caption.
+#| label: fig-1
+#| fig-cap: A figure caption.
 #| echo: true
@@
 x = -3.0:0.01:3.0
-y = rand(length(x));
+y = rand(length(x))
+plot(x, y)
 ```
 
-Unfortunately, I don't know how to use the special environment that is already defined for Julia in this context.
+::: {.callout-note}
 
-# Code Listings
+# Current Limitations
 
-The special environment that is already defined for Julia code can still be used as before.
+Support for executable listings is only available in the development version of Quarto (version $\geq$ 1.4.0).
+It is currently not clear how to support custom formatting, for example through the existing `jlcode.sty` environment.
+
+:::
+
+## Static Listings
+
+The special environment that is already defined for Julia code is still compatible with this template.
 
@@
 \end{lstlisting}
+
+# Tables
+
+As with code, there are various options to present tables.
+Using standard Markdown syntax is the most flexible option because it is compatible with various output formats.
+Alternatively, standard LaTeX syntax can be used.
+
+## Markdown Tables
+
+Markdown tables are rendered to LaTeX using the `booktabs` package.
+This is the recommended way to create tables because it is compatible with various output formats.
+Tables can be cross-referenced and the table caption will be rendered as a LaTeX table caption.
+The table generated through the following Markdown syntax is labeled as `tbl-1` and calling `@tbl-1` will render as @tbl-1.
+
+```{.markdown}
+
+| Default | Left | Right | Center |
+|---------|:-----|------:|:------:|
+| 12      | 12   |    12 |   12   |
+| 123     | 123  |   123 |  123   |
+| 1       | 1    |     1 |   1    |
+
+: Demonstration of pipe table syntax {#tbl-1}
+
+```
+
+| Default | Left | Right | Center |
+|---------|:-----|------:|:------:|
+| 12      | 12   |    12 |   12   |
+| 123     | 123  |   123 |  123   |
+| 1       | 1    |     1 |   1    |
+
+: Demonstration of pipe table syntax {#tbl-1}

### 2023-11-16 template.qmd

@@
 
-## Executable Listings  (version $\geq$ 1.4.0)
+## Executable Listings  (version $\geq$ 1.4.0) {#sec-executable-listings}
 
@@
 
-For example, the following code block will be labeled as `lst-1` and calling `@lst-1` will render as @lst-1.
-The figure it generates will be labeled as `fig-1` and calling `@fig-1` will render as @fig-1.
+For example, the following code block will be labelled as `lst-1` and calling `@lst-1` will render as @lst-1.
+The figure it generates will be labelled as `fig-1` and calling `@fig-1` will render as @fig-1.
 
@@
 #| fig-cap: A figure caption.
+#| eval: true
 #| echo: true
@@
 Tables can be cross-referenced and the table caption will be rendered as a LaTeX table caption.
-The table generated through the following Markdown syntax is labeled as `tbl-1` and calling `@tbl-1` will render as @tbl-1.
 
-```{.markdown}
+```
 
@@
 
-| Default | Left | Right | Center |
+<!-- | Default | Left | Right | Center |
 |---------|:-----|------:|:------:|
@@
 
-: Demonstration of pipe table syntax {#tbl-1}
+: Demonstration of pipe table syntax {#tbl-1} -->
+
+::: {.callout-note}
+
+# Current Limitations
+
+This seems to currently mess with executable listing (@sec-executable-listings).
+
+:::
+
+## LaTeX Tables
+
+Standard LaTeX syntax can also be used to create tables.
+To use standard Quarto cross-referencing, the LaTeX syntax needs to be wrapped in a `div` with the `tbl` class.
+The table caption can be added as a paragraph.
+The following creates @tbl-1 which can be cross-referenced as `@tbl-1`.
+
+```
+::: {#tbl-1}
+
+\begin{tabular}{|l|l|}\hline
+Age & Frequency \\ \hline
+18--25  & 15 \\
+26--35  & 33 \\
+36--45  & 22 \\ \hline
+\end{tabular}
+
+This is a table caption.
+
+:::
+```
+
+::: {#tbl-1}
+
+\begin{tabular}{|l|l|}\hline
+Age & Frequency \\ \hline
+18--25  & 15 \\
+26--35  & 33 \\
+36--45  & 22 \\ \hline
+\end{tabular}
+
+This is a table caption.
+
+:::

### 2023-11-17 template.qmd
that's pretty close

@@
 
+Alternatively, we can use the existing `lstlisting` environment for Julia code using the `julia` language option.
+This is the same environment that is used for Julia code in the existing LaTeX template.
+Under the hood, this uses a lua filter to convert the code to LaTeX.
+The code creates @fig-2 which can be cross-referenced as `@fig-2`.
+
+```{julia}
+#| label: fig-2
+#| fig-cap: A figure caption.
+#| eval: true
+#| echo: true
+
+using Plots
+
+x = -3.0:0.01:3.0
+y = cumsum(rand(length(x)))
+plot(x, y)
+```
+
+::: {.callout-note}
+
+# Current Limitations
+
+I don't know how to cross-reference the code itself.
+
+:::
+
 ## Static Listings

### 2023-11-17 examples/CounterfactualExplanations.jl/paper.tex
bst seems to have no effect

@@
 }
+
+% Two-column table
+% Lifted from https://github.com/quarto-journals/elsevier/blob/main/_extensions/elsevier/partials/_two-column-longtable.tex.
+\usepackage{float}
+\makeatletter
+\let\oldlt\longtable
+\let\endoldlt\endlongtable
+\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
+\def\longtable@i[#1]{\begin{figure}[H]
+\onecolumn
+\begin{minipage}{0.5\textwidth}
+\oldlt[#1]
+}
+\def\longtable@ii{\begin{figure}[H]
+\onecolumn
+\begin{minipage}{0.5\textwidth}
+\oldlt
+}
+\def\endlongtable{\endoldlt
+\end{minipage}
+\twocolumn
+\end{figure}}
+\makeatother
+
+% Remove whitespace after paragraphs
+\setlength{\parskip}{0pt}
+
 \usepackage{orcidlink}
@@
 problems typically associated with these models: they tend to be
-unstable (Goodfellow, Shlens, and Szegedy 2014), encode existing biases
-(Buolamwini and Gebru 2018) and learn representations that are
-surprising or even counter-intuitive from a human perspective
-(Buolamwini and Gebru 2018). Nonetheless, they often form the basis for
-data-driven decision-making systems in real-world applications.
+unstable (\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens,
+and Szegedy 2014}), encode existing biases
+(\citeproc{ref-buolamwini2018gender}{Buolamwini and Gebru 2018}) and
+learn representations that are surprising or even counter-intuitive from
+a human perspective (\citeproc{ref-buolamwini2018gender}{Buolamwini and
+Gebru 2018}). Nonetheless, they often form the basis for data-driven
+decision-making systems in real-world applications.
 
@@
 stakeholders---that fail to understand the behaviour of their
-agent---i.e.~the black-box system (Borch 2022). The group of principals
-may include programmers, product managers and other decision-makers who
-develop and operate the system as well as those individuals ultimately
-subject to the decisions made by the system. In practice, decisions made
-by black-box systems are typically left unchallenged since the group of
-principals cannot scrutinize them:
+agent---i.e.~the black-box system (\citeproc{ref-borch2022machine}{Borch
+2022}). The group of principals may include programmers, product
+managers and other decision-makers who develop and operate the system as
+well as those individuals ultimately subject to the decisions made by
+the system. In practice, decisions made by black-box systems are
+typically left unchallenged since the group of principals cannot
+scrutinize them:
 
@@
 ``You cannot appeal to (algorithms). They do not listen. Nor do they
-bend.'' (O'Neil 2016)
+bend.'' (\citeproc{ref-oneil2016weapons}{O'Neil 2016})
 \end{quote}
@@
 concepts. The following literature surveys provide more detail: Arrieta
-et al.~(2020) provide a broad overview of XAI (Arrieta et al. 2020); Fan
-et al.~(2020) focus on explainability in the context of deep learning
-(Fan, Xiong, and Wang 2020); and finally, Karimi et al.~(2020) (Karimi,
-Barthe, et al. 2020) and Verma et al.~(2020) Verma, Dickerson, and Hines
-(2020) offer detailed reviews of the literature on Counterfactual
-Explanations and Algorithmic Recourse (see also Molnar (2020) and
-Varshney (2022)). Miller (2019) explicitly discusses the concept of
-explainability from the perspective of a social scientist (Miller 2019).
+et al.~(2020) provide a broad overview of XAI
+(\citeproc{ref-arrieta2020explainable}{Arrieta et al. 2020}); Fan et
+al.~(2020) focus on explainability in the context of deep learning
+(\citeproc{ref-fan2020interpretability}{Fan, Xiong, and Wang 2020}); and
+finally, Karimi et al.~(2020) (\citeproc{ref-karimi2020survey}{Karimi,
+Barthe, et al. 2020}) and Verma et al.~(2020) Verma, Dickerson, and
+Hines (\citeproc{ref-verma2020counterfactual}{2020}) offer detailed
+reviews of the literature on Counterfactual Explanations and Algorithmic
+Recourse (see also Molnar (\citeproc{ref-molnar2020interpretable}{2020})
+and Varshney (\citeproc{ref-varshney2022trustworthy}{2022})). Miller
+(2019) explicitly discusses the concept of explainability from the
+perspective of a social scientist
+(\citeproc{ref-miller2019explanation}{Miller 2019}).
 
@@
 often used interchangeably, but this can lead to confusion. We find the
-distinction made in Rudin (2019) useful: Interpretable AI involves
-models that are inherently interpretable and transparent such as general
-additive models (GAM), decision trees and rule-based models; Explainable
-AI involves models that are not inherently interpretable but require
-additional tools to be explainable to humans. Examples of the latter
-include Ensembles, Support Vector Machines and Deep Neural Networks.
-Some would argue that we best avoid the second category of models
-altogether and instead focus solely on interpretable AI Rudin (2019).
-While we agree that initial efforts should always be geared towards
-interpretable models, avoiding black boxes altogether would entail
-missed opportunities and anyway is probably not very realistic at this
-point. For that reason, we expect the need for XAI to persist in the
-medium term. Explainable AI can further be broadly divided into
-\textbf{global} and \textbf{local} explainability: the former is
-concerned with explaining the average behaviour of a model, while the
-latter involves explanations for individual predictions (Molnar 2020).
-Tools for global explainability include partial dependence plots (PDP),
-which involve the computation of marginal effects through Monte Carlo,
-and global surrogates. A surrogate model is an interpretable model that
-is trained to explain the predictions of a black-box model.
+distinction made in Rudin (\citeproc{ref-rudin2019stop}{2019}) useful:
+Interpretable AI involves models that are inherently interpretable and
+transparent such as general additive models (GAM), decision trees and
+rule-based models; Explainable AI involves models that are not
+inherently interpretable but require additional tools to be explainable
+to humans. Examples of the latter include Ensembles, Support Vector
+Machines and Deep Neural Networks. Some would argue that we best avoid
+the second category of models altogether and instead focus solely on
+interpretable AI Rudin (\citeproc{ref-rudin2019stop}{2019}). While we
+agree that initial efforts should always be geared towards interpretable
+models, avoiding black boxes altogether would entail missed
+opportunities and anyway is probably not very realistic at this point.
+For that reason, we expect the need for XAI to persist in the medium
+term. Explainable AI can further be broadly divided into \textbf{global}
+and \textbf{local} explainability: the former is concerned with
+explaining the average behaviour of a model, while the latter involves
+explanations for individual predictions
+(\citeproc{ref-molnar2020interpretable}{Molnar 2020}). Tools for global
+explainability include partial dependence plots (PDP), which involve the
+computation of marginal effects through Monte Carlo, and global
+surrogates. A surrogate model is an interpretable model that is trained
+to explain the predictions of a black-box model.
 
@@
 A recent study has shown that even experienced data scientists tend to
-put too much trust in explanations produced by LIME and SHAP (Kaur et
-al. 2020). Another recent work has shown that both methods can be easily
-fooled: they depend on random input perturbations, a property that can
-be abused by adverse agents to essentially whitewash strongly biased
-black-box models (Slack et al. 2020). In related work, the same authors
-find that while gradient-based Counterfactual Explanations can also be
-manipulated, there is a straightforward way to protect against this in
-practice (Slack et al. 2021). In the context of quality assessment, it
-is also worth noting that---contrary to surrogate explainers---CE always
-achieve full fidelity by construction: counterfactuals are searched with
-respect to the black-box classifier, not some proxy for it. That being
-said, CE should also be used with care and research around them is still
-in its early stages.
+put too much trust in explanations produced by LIME and SHAP
+(\citeproc{ref-kaur2020interpreting}{Kaur et al. 2020}). Another recent
+work has shown that both methods can be easily fooled: they depend on
+random input perturbations, a property that can be abused by adverse
+agents to essentially whitewash strongly biased black-box models
+(\citeproc{ref-slack2020fooling}{Slack et al. 2020}). In related work,
+the same authors find that while gradient-based Counterfactual
+Explanations can also be manipulated, there is a straightforward way to
+protect against this in practice
+(\citeproc{ref-slack2021counterfactual}{Slack et al. 2021}). In the
+context of quality assessment, it is also worth noting that---contrary
+to surrogate explainers---CE always achieve full fidelity by
+construction: counterfactuals are searched with respect to the black-box
+classifier, not some proxy for it. That being said, CE should also be
+used with care and research around them is still in its early stages.
 
@@
 in understanding how we need to change individual attributes in order to
-change the model output to a desired value or label (Molnar 2020).
-Typically the underlying methodology is presented in the context of
-binary classification: \(M: \mathcal{X} \mapsto \mathcal{Y}\) where
+change the model output to a desired value or label
+(\citeproc{ref-molnar2020interpretable}{Molnar 2020}). Typically the
+underlying methodology is presented in the context of binary
+classification: \(M: \mathcal{X} \mapsto \mathcal{Y}\) where
 \(\mathcal{X}\subset\mathbb{R}^D\) and \(\mathcal{Y}=\{0,1\}\). Further,
@@
 that the ideas presented here also carry over to multi-class problems
-and regression (Molnar 2020).
+and regression (\citeproc{ref-molnar2020interpretable}{Molnar 2020}).
 
 The counterfactual search objective originally proposed by Wachter,
-Mittelstadt, and Russell (2017) is as follows
+Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}) is as follows
 
@@
 Numerous extensions to this simple approach have been developed since CE
-were first proposed in 2017 (see Verma, Dickerson, and Hines (2020) and
-(Karimi, Barthe, et al. 2020) for surveys). The various approaches
-largely differ in that they use different flavours of search objective
-defined in Equation~\ref{eq-solution}. Different penalties are often
-used to address many of the desirable properties of effective CE that
-have been set out. These desiderata include: \textbf{proximity} --- the
-distance between factual and counterfactual features should be small
-(Wachter, Mittelstadt, and Russell 2017); \textbf{actionability} --- the
-proposed recourse should be actionable Poyiadzi et al. (2020);
+were first proposed in 2017 (see Verma, Dickerson, and Hines
+(\citeproc{ref-verma2020counterfactual}{2020}) and
+(\citeproc{ref-karimi2020survey}{Karimi, Barthe, et al. 2020}) for
+surveys). The various approaches largely differ in that they use
+different flavours of search objective defined in
+Equation~\ref{eq-solution}. Different penalties are often used to
+address many of the desirable properties of effective CE that have been
+set out. These desiderata include: \textbf{proximity} --- the distance
+between factual and counterfactual features should be small
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}); \textbf{actionability} --- the proposed recourse should
+be actionable Poyiadzi et al. (\citeproc{ref-poyiadzi2020face}{2020});
 \textbf{plausibility} --- the counterfactual explanation should be
-plausible to a human Schut et al. (2021); \textbf{sparsity} --- the
+plausible to a human Schut et al.
+(\citeproc{ref-schut2021generating}{2021}); \textbf{sparsity} --- the
 counterfactual explanation should involve as few individual feature
-changes as possible (Schut et al. 2021); \textbf{robustness} --- the
-counterfactual explanation should be robust to domain and model shifts
-(Upadhyay, Joshi, and Lakkaraju 2021); \textbf{diversity} --- ideally
-multiple diverse counterfactuals should be provided (Mothilal, Sharma,
-and Tan 2020); and \textbf{causality} --- counterfactuals should respect
-the structural causal model underlying the data generating process
-Karimi, Schölkopf, and Valera (2021).
+changes as possible (\citeproc{ref-schut2021generating}{Schut et al.
+2021}); \textbf{robustness} --- the counterfactual explanation should be
+robust to domain and model shifts
+(\citeproc{ref-upadhyay2021robust}{Upadhyay, Joshi, and Lakkaraju
+2021}); \textbf{diversity} --- ideally multiple diverse counterfactuals
+should be provided (\citeproc{ref-mothilal2020explaining}{Mothilal,
+Sharma, and Tan 2020}); and \textbf{causality} --- counterfactuals
+should respect the structural causal model underlying the data
+generating process Karimi, Schölkopf, and Valera
+(\citeproc{ref-karimi2021algorithmic}{2021}).
 
@@
 \href{https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest}{CARLA}
-(Pawelczyk et al. 2021).\footnote{While we were writing this paper, the
-  \texttt{R} package \texttt{counterfactuals} was released (Dandl et al.
-  2023). The developers seem to also envision a unifying framework, but
-  the project appears to still be in its early stages.} In addition to
-that, there exists open-source code for some specific approaches to CE
-that have been proposed in recent years. The approach-specific
-implementations that we have been able to find are generally
-well-documented, but exclusively in Python. For example, a PyTorch
-implementation of a greedy generator for Bayesian models proposed in
-Schut et al. (2021) has been released. As another example, the popular
-\href{https://github.com/interpretml}{InterpretML} library includes an
-implementation of a diverse counterfactual generator (Mothilal, Sharma,
-and Tan 2020).
+(\citeproc{ref-pawelczyk2021carla}{Pawelczyk et al. 2021}).\footnote{While
+  we were writing this paper, the \texttt{R} package
+  \texttt{counterfactuals} was released
+  (\citeproc{ref-dandl2023counterfactuals}{Dandl et al. 2023}). The
+  developers seem to also envision a unifying framework, but the project
+  appears to still be in its early stages.} In addition to that, there
+exists open-source code for some specific approaches to CE that have
+been proposed in recent years. The approach-specific implementations
+that we have been able to find are generally well-documented, but
+exclusively in Python. For example, a PyTorch implementation of a greedy
+generator for Bayesian models proposed in Schut et al.
+(\citeproc{ref-schut2021generating}{2021}) has been released. As another
+example, the popular \href{https://github.com/interpretml}{InterpretML}
+library includes an implementation of a diverse counterfactual generator
+(\citeproc{ref-mothilal2020explaining}{Mothilal, Sharma, and Tan 2020}).
 
@@
 The package currently offers native support for models built and trained
-in \href{https://fluxml.ai/}{Flux} (Innes 2018) as well as a small
-subset of models made available through
-\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ} (Blaom
-et al. 2020). While in general it is assumed that users resort to this
-package to explain their pre-trained models, we provide a simple API
-call to train the following
+in \href{https://fluxml.ai/}{Flux} (\citeproc{ref-innes2018flux}{Innes
+2018}) as well as a small subset of models made available through
+\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ}
+(\citeproc{ref-blaom2020mlj}{Blaom et al. 2020}). While in general it is
+assumed that users resort to this package to explain their pre-trained
+models, we provide a simple API call to train the following
 \href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/}{models}:
@@
 \item
-  Deep Ensemble Lakshminarayanan, Pritzel, and Blundell (2016)
+  Deep Ensemble Lakshminarayanan, Pritzel, and Blundell
+  (\citeproc{ref-lakshminarayanan2016simple}{2016})
 \item
@@
 embedding, existing work has typically relied on generative models or
-existing causal knowledge Karimi, Schölkopf, and Valera (2021). While
-this notion is compatible with all of our gradient-based generators,
-only some generators search a latent space by default. Finally,
-composability implies that the given generator can be blended with any
-other composable generator, which we discuss in
-Section~\ref{sec-gen-comp}.
+existing causal knowledge Karimi, Schölkopf, and Valera
+(\citeproc{ref-karimi2021algorithmic}{2021}). While this notion is
+compatible with all of our gradient-based generators, only some
+generators search a latent space by default. Finally, composability
+implies that the given generator can be blended with any other
+composable generator, which we discuss in Section~\ref{sec-gen-comp}.
 
@@
 \emph{ClapROAR} aims to preserve the classifier, i.e.~to generate
-counterfactuals that are robust to endogenous model shifts (Altmeyer et
-al. 2023); \emph{CLUE} searches plausible counterfactuals in the latent
-embedding of a generative model by explicitly minimising predictive
-entropy (Antorán et al. 2020); \emph{DiCE} is designed to generate
-multiple, maximally diverse counterfactuals (Mothilal, Sharma, and Tan
-2020); \emph{FeatureTweak} leverages the internals of decision trees to
-search counterfactuals on a feature-by-feature basis, finding the
-counterfactual that tweaks the features in the least costly way (Tolomei
-et al. 2017); \emph{Gravitational} aims to generate plausible and robust
+counterfactuals that are robust to endogenous model shifts
+(\citeproc{ref-altmeyer2023endogenous}{Altmeyer et al. 2023});
+\emph{CLUE} searches plausible counterfactuals in the latent embedding
+of a generative model by explicitly minimising predictive entropy
+(\citeproc{ref-antoran2020getting}{Antorán et al. 2020}); \emph{DiCE} is
+designed to generate multiple, maximally diverse counterfactuals
+(\citeproc{ref-mothilal2020explaining}{Mothilal, Sharma, and Tan 2020});
+\emph{FeatureTweak} leverages the internals of decision trees to search
+counterfactuals on a feature-by-feature basis, finding the
+counterfactual that tweaks the features in the least costly way
+(\citeproc{ref-tolomei2017interpretable}{Tolomei et al. 2017});
+\emph{Gravitational} aims to generate plausible and robust
 counterfactuals by minimising the distance to observed samples in the
-target class (Altmeyer et al. 2023); \emph{Greedy} aims to generate
-plausible counterfactuals by implicitly minimising predictive
-uncertainty of Bayesian classifiers (Schut et al. 2021);
+target class (\citeproc{ref-altmeyer2023endogenous}{Altmeyer et al.
+2023}); \emph{Greedy} aims to generate plausible counterfactuals by
+implicitly minimising predictive uncertainty of Bayesian classifiers
+(\citeproc{ref-schut2021generating}{Schut et al. 2021});
 \emph{GrowingSpheres} is model-agnostic, relying solely on identifying
@@
 increasing the search radius and then moving counterfactuals in that
-direction(Laugel et al. 2017); \emph{PROBE} generates probabilistically
-robust counterfactuals (Pawelczyk et al. 2022); \emph{REVISE} addresses
-the need for plausibility by searching counterfactuals in the latent
-embedding of a Variational Autoencoder (VAE) (Joshi et al. 2019);
+direction(\citeproc{ref-laugel2017inversea}{Laugel et al. 2017});
+\emph{PROBE} generates probabilistically robust counterfactuals
+(\citeproc{ref-pawelczyk2022probabilistically}{Pawelczyk et al. 2022});
+\emph{REVISE} addresses the need for plausibility by searching
+counterfactuals in the latent embedding of a Variational Autoencoder
+(VAE) (\citeproc{ref-joshi2019realistic}{Joshi et al. 2019});
 \emph{Wachter} is the baseline approach that only penalises the distance
-to the original sample (Wachter, Mittelstadt, and Russell 2017).
-
-\onecolumn
+to the original sample
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}).
 
@@
 \endlastfoot
-ClaPROAR (Altmeyer et al. 2023) & gradient based & feature & yes \\
-CLUE (Antorán et al. 2020) & gradient based & latent & yes \\
-DiCE (Mothilal, Sharma, and Tan 2020) & gradient based & feature &
-yes \\
-FeatureTweak (Tolomei et al. 2017) & tree based & feature & no \\
-Gravitational (Altmeyer et al. 2023) & gradient based & feature & yes \\
-Greedy (Schut et al. 2021) & gradient based & feature & yes \\
-GrowingSpheres (Laugel et al. 2017) & agnostic & feature & no \\
-PROBE (Pawelczyk et al. 2022) & gradient based & feature & no \\
-REVISE (Joshi et al. 2019) & gradient based & latent & yes \\
-Wachter (Wachter, Mittelstadt, and Russell 2017) & gradient based &
-feature & yes \\
+ClaPROAR (\citeproc{ref-altmeyer2023endogenous}{Altmeyer et al. 2023}) &
+gradient based & feature & yes \\
+CLUE (\citeproc{ref-antoran2020getting}{Antorán et al. 2020}) & gradient
+based & latent & yes \\
+DiCE (\citeproc{ref-mothilal2020explaining}{Mothilal, Sharma, and Tan
+2020}) & gradient based & feature & yes \\
+FeatureTweak (\citeproc{ref-tolomei2017interpretable}{Tolomei et al.
+2017}) & tree based & feature & no \\
+Gravitational (\citeproc{ref-altmeyer2023endogenous}{Altmeyer et al.
+2023}) & gradient based & feature & yes \\
+Greedy (\citeproc{ref-schut2021generating}{Schut et al. 2021}) &
+gradient based & feature & yes \\
+GrowingSpheres (\citeproc{ref-laugel2017inversea}{Laugel et al. 2017}) &
+agnostic & feature & no \\
+PROBE (\citeproc{ref-pawelczyk2022probabilistically}{Pawelczyk et al.
+2022}) & gradient based & feature & no \\
+REVISE (\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}) & gradient
+based & latent & yes \\
+Wachter (\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt,
+and Russell 2017}) & gradient based & feature & yes \\
 \end{longtable}
-
-\twocolumn
 
@@
 \item
-  Adult Census (Barry Becker 1996)
+  Adult Census (\citeproc{ref-becker1996adult}{Barry Becker 1996})
 \item
-  California Housing (Pace and Barry 1997)
+  California Housing (\citeproc{ref-pace1997sparse}{Pace and Barry
+  1997})
 \item
-  CIFAR10 (Krizhevsky 2009)
+  CIFAR10 (\citeproc{ref-krizhevsky2009learning}{Krizhevsky 2009})
 \item
-  German Credit (Hoffman 1994)
+  German Credit (\citeproc{ref-hoffman1994german}{Hoffman 1994})
 \item
-  Give Me Some Credit (Kaggle 2011)
+  Give Me Some Credit (\citeproc{ref-kaggle2011give}{Kaggle 2011})
 \item
-  MNIST (LeCun 1998) and Fashion MNIST (Xiao, Rasul, and Vollgraf 2017)
+  MNIST (\citeproc{ref-lecun1998mnist}{LeCun 1998}) and Fashion MNIST
+  (\citeproc{ref-xiao2017fashion}{Xiao, Rasul, and Vollgraf 2017})
 \item
-  UCI defaultCredit (Yeh and Lien 2009)
+  UCI defaultCredit (\citeproc{ref-yeh2009comparisons}{Yeh and Lien
+  2009})
 \end{itemize}
@@
 \(\mathcal{X}\) and \(\text{diversity}(\cdot)\) is the penalty proposed
-by Mothilal, Sharma, and Tan (2020) that induces diverse sets of
-counterfactuals. As in Equation~\ref{eq-solution}, \(\ell\) is the loss
-function, \(M\) is the black-box model, \(t\) is the target class, and
-\(\lambda\) is the strength of the penalty.
+by Mothilal, Sharma, and Tan
+(\citeproc{ref-mothilal2020explaining}{2020}) that induces diverse sets
+of counterfactuals. As in Equation~\ref{eq-solution}, \(\ell\) is the
+loss function, \(M\) is the black-box model, \(t\) is the target class,
+and \(\lambda\) is the strength of the penalty.
 
@@
 counterfactual from its original value. The API call to compute the
-distance metric defined in Wachter, Mittelstadt, and Russell (2017), for
-instance, is as simple as \texttt{evaluate(ce;\ measure=distance\_mad)},
-where \texttt{ce} can also be a vector of
-\texttt{CounterfactualExplanation}s.
+distance metric defined in Wachter, Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}), for instance, is as
+simple as \texttt{evaluate(ce;\ measure=distance\_mad)}, where
+\texttt{ce} can also be a vector of \texttt{CounterfactualExplanation}s.
 
@@
 how native support for \href{https://fluxml.ai/}{\texttt{Flux.jl}}
-(Innes 2018) deep learning models was enabled.\footnote{Flux models are
-  now natively supported by our package and can be instantiated by
-  calling \texttt{FluxModel()}.} Once again we use synthetic data for an
-illustrative example. Code \ref{lst:nn} below builds a simple model
-architecture that can be used for a multi-class prediction task. Note
-how outputs from the final layer are not passed through a softmax
-activation function, since the counterfactual loss is evaluated with
-respect to logits as we discussed earlier. The model is trained with
-dropout.
+(\citeproc{ref-innes2018flux}{Innes 2018}) deep learning models was
+enabled.\footnote{Flux models are now natively supported by our package
+  and can be instantiated by calling \texttt{FluxModel()}.} Once again
+we use synthetic data for an illustrative example. Code \ref{lst:nn}
+below builds a simple model architecture that can be used for a
+multi-class prediction task. Note how outputs from the final layer are
+not passed through a softmax activation function, since the
+counterfactual loss is evaluated with respect to logits as we discussed
+earlier. The model is trained with dropout.
 
@@
 quantifying the uncertainty around the generated counterfactuals
-(Delaney, Greene, and Keane 2021). One idea could be, to use dropout to
-randomly switch features on and off in each iteration. Without dwelling
-further on the merit of this idea, we will now briefly show how this can
-be implemented.
+(\citeproc{ref-delaney2021uncertainty}{Delaney, Greene, and Keane
+2021}). One idea could be, to use dropout to randomly switch features on
+and off in each iteration. Without dwelling further on the merit of this
+idea, we will now briefly show how this can be implemented.
 
@@
 The \emph{Give Me Some Credit} dataset is one of the tabular real-world
-datasets that ship with the package (Kaggle 2011). It can be used to
-train a binary classifier to predict whether a borrower is likely to
-experience financial difficulties in the next two years. In particular,
-we have an output variable
-\(y \in \{0=\texttt{no stress},1=\texttt{stress}\}\) and a feature
-matrix \(X\) that includes socio-demographic variables like \texttt{age}
-and \texttt{income}. A retail bank might use such a classifier to
-determine if potential borrowers should receive credit or not.
+datasets that ship with the package
+(\citeproc{ref-kaggle2011give}{Kaggle 2011}). It can be used to train a
+binary classifier to predict whether a borrower is likely to experience
+financial difficulties in the next two years. In particular, we have an
+output variable \(y \in \{0=\texttt{no stress},1=\texttt{stress}\}\) and
+a feature matrix \(X\) that includes socio-demographic variables like
+\texttt{age} and \texttt{income}. A retail bank might use such a
+classifier to determine if potential borrowers should receive credit or
+not.
 
 For the classification task, we use a Multi-Layer Perceptron with
-dropout regularization. Using the Gravitational generator (Altmeyer et
-al. 2023) we will generate counterfactuals for ten randomly chosen
-individuals that would be denied credit based on our pre-trained model.
-Concerning the mutability of features, we only impose that the
-\texttt{age} cannot be decreased.
+dropout regularization. Using the Gravitational generator
+(\citeproc{ref-altmeyer2023endogenous}{Altmeyer et al. 2023}) we will
+generate counterfactuals for ten randomly chosen individuals that would
+be denied credit based on our pre-trained model. Concerning the
+mutability of features, we only impose that the \texttt{age} cannot be
+decreased.
 
@@
 contains 60,000 training samples of handwritten digits in the form of
-28x28 pixel grey-scale images (LeCun 1998). Each image is associated
-with a label indicating the digit (0-9) that the image represents. The
-data makes for an interesting case study of CE because humans have a
-good idea of what plausible counterfactuals of digits look like. For
-example, if you were asked to pick up an eraser and turn the digit in
-the left panel of Figure~\ref{fig-mnist} into a four (4) you would know
-exactly what to do: just erase the top part.
+28x28 pixel grey-scale images (\citeproc{ref-lecun1998mnist}{LeCun
+1998}). Each image is associated with a label indicating the digit (0-9)
+that the image represents. The data makes for an interesting case study
+of CE because humans have a good idea of what plausible counterfactuals
+of digits look like. For example, if you were asked to pick up an eraser
+and turn the digit in the left panel of Figure~\ref{fig-mnist} into a
+four (4) you would know exactly what to do: just erase the top part.
 
@@
 supervised models built in \texttt{MLJ.jl}, an extensive Machine
-Learning framework for Julia (Blaom et al. 2020). This may also involve
-adding support for regression models as well as additional
-non-differentiable models. In terms of counterfactual generators, there
-is a list of recent methodologies that we would like to implement
-including MINT (Karimi, Schölkopf, and Valera 2021), ROAR (Upadhyay,
-Joshi, and Lakkaraju 2021) and FACE (Poyiadzi et al. 2020).
+Learning framework for Julia (\citeproc{ref-blaom2020mlj}{Blaom et al.
+2020}). This may also involve adding support for regression models as
+well as additional non-differentiable models. In terms of counterfactual
+generators, there is a list of recent methodologies that we would like
+to implement including MINT
+(\citeproc{ref-karimi2021algorithmic}{Karimi, Schölkopf, and Valera
+2021}), ROAR (\citeproc{ref-upadhyay2021robust}{Upadhyay, Joshi, and
+Lakkaraju 2021}) and FACE (\citeproc{ref-poyiadzi2020face}{Poyiadzi et
+al. 2020}).
 
@@
 explanations including \emph{Adult}, \emph{Give Me Some Credit} and
-\emph{German Credit} (Karimi, Barthe, et al. 2020). There is scope for
-adding data sources that have so far not been explored much in this
-context including additional image datasets as well as audio, natural
-language and time-series data.
+\emph{German Credit} (\citeproc{ref-karimi2020survey}{Karimi, Barthe, et
+al. 2020}). There is scope for adding data sources that have so far not
+been explored much in this context including additional image datasets
+as well as audio, natural language and time-series data.
 
@@
 
+% **************GENERATED FILE, DO NOT EDIT**************
+
+\bibliographystyle{juliacon}
 \end{document}

### 2023-11-17 examples/CounterfactualExplanations.jl/paper.qmd
this seems to work

@@
 
-Numerous extensions to this simple approach have been developed since CE were first proposed in 2017 (see @verma2020counterfactual and [@karimi2020survey] for surveys).
+Numerous extensions to this simple approach have been developed since CE were first proposed in 2017 (see @verma2020counterfactual and @karimi2020survey for surveys).
 The various approaches largely differ in that they use different flavours of search objective defined in @eq-solution.
 Different penalties are often used to address many of the desirable properties of effective CE that have been set out.
-These desiderata include: **proximity** --- the distance between factual and counterfactual features should be small [@wachter2017counterfactual]; **actionability** --- the proposed recourse should be actionable [@ustun2019actionable, @poyiadzi2020face]; **plausibility** --- the counterfactual explanation should be plausible to a human [@joshi2019realistic, @schut2021generating]; **sparsity** --- the counterfactual explanation should involve as few individual feature changes as possible [@schut2021generating]; **robustness** --- the counterfactual explanation should be robust to domain and model shifts [@upadhyay2021robust]; **diversity** --- ideally multiple diverse counterfactuals should be provided [@mothilal2020explaining]; and **causality** --- counterfactuals should respect the structural causal model underlying the data generating process [@karimi2020algorithmic,@karimi2021algorithmic].
+These desiderata include: **proximity** --- the distance between factual and counterfactual features should be small [@wachter2017counterfactual]; **actionability** --- the proposed recourse should be actionable [@ustun2019actionable; @poyiadzi2020face]; **plausibility** --- the counterfactual explanation should be plausible to a human [@joshi2019realistic;@schut2021generating]; **sparsity** --- the counterfactual explanation should involve as few individual feature changes as possible [@schut2021generating]; **robustness** --- the counterfactual explanation should be robust to domain and model shifts [@upadhyay2021robust]; **diversity** --- ideally multiple diverse counterfactuals should be provided [@mothilal2020explaining]; and **causality** --- counterfactuals should respect the structural causal model underlying the data generating process [@karimi2020algorithmic;@karimi2021algorithmic].
 
@@
 Concerning the search space, it is possible to search counterfactuals in a lower-dimensional latent embedding of the feature space that implicitly encodes the data-generating process (DGP).
-To learn the latent embedding, existing work has typically relied on generative models or existing causal knowledge [@joshi2019realistic, @karimi2021algorithmic].
+To learn the latent embedding, existing work has typically relied on generative models or existing causal knowledge [@joshi2019realistic;@karimi2021algorithmic].
 While this notion is compatible with all of our gradient-based generators, only some generators search a latent space by default.
@@
 to generate counterfactuals that are robust to endogenous model shifts [@altmeyer2023endogenous]; *CLUE* searches plausible counterfactuals in the latent embedding of a generative model by explicitly minimising predictive entropy [@antoran2020getting]; *DiCE* is designed to generate multiple, maximally diverse counterfactuals [@mothilal2020explaining]; *FeatureTweak* leverages the internals of decision trees to search counterfactuals on a feature-by-feature basis, finding the counterfactual that tweaks the features in the least costly way [@tolomei2017interpretable]; *Gravitational* aims to generate plausible and robust counterfactuals by minimising the distance to observed samples in the target class [@altmeyer2023endogenous]; *Greedy* aims to generate plausible counterfactuals by implicitly minimising predictive uncertainty of Bayesian classifiers [@schut2021generating]; *GrowingSpheres* is model-agnostic, relying solely on identifying nearest neighbours of counterfactuals in the target class by gradually increasing the search radius and then moving counterfactuals in that direction[ @laugel2017inversea]; *PROBE* generates probabilistically robust counterfactuals [@pawelczyk2022probabilistically]; *REVISE* addresses the need for plausibility by searching counterfactuals in the latent embedding of a Variational Autoencoder (VAE) [@joshi2019realistic]; *Wachter* is the baseline approach that only penalises the distance to the original sample [@wachter2017counterfactual].
-
-\onecolumn
 
@@
 : Overview of implemented counterfactual generators. {#tbl-gen}
-
-\twocolumn
 

### 2023-11-17 examples/CounterfactualExplanations.jl/paper.tex
uf

--- 2023-11-16 examples/CounterfactualExplanations.jl/paper.tex
@@
 \makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 }
+
+% Two-column table
+% Lifted from https://github.com/quarto-journals/elsevier/blob/main/_extensions/elsevier/partials/_two-column-longtable.tex.
+\usepackage{float}
+\makeatletter
+\let\oldlt\longtable
+\let\endoldlt\endlongtable
+\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
+\def\longtable@i[#1]{\begin{figure}[H]
+\onecolumn
+\begin{minipage}{0.5\textwidth}
+\oldlt[#1]
+}
+\def\longtable@ii{\begin{figure}[H]
+\onecolumn
+\begin{minipage}{0.5\textwidth}
+\oldlt
+}
+\def\endlongtable{\endoldlt
+\end{minipage}
+\twocolumn
+\end{figure}}
+\makeatother
+
+% Remove whitespace after paragraphs
+\setlength{\parskip}{0pt}
+
+\bibliographystyle{juliacon}
 \usepackage{orcidlink}
@@
 \fi
+\usepackage[]{biblatex}
+\addbibresource{ref.bib}
 \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 problems typically associated with these models: they tend to be
-unstable (Goodfellow, Shlens, and Szegedy 2014), encode existing biases
-(Buolamwini and Gebru 2018) and learn representations that are
+unstable \autocite{goodfellow2014explaining}, encode existing biases
+\autocite{buolamwini2018gender} and learn representations that are
 surprising or even counter-intuitive from a human perspective
-(Buolamwini and Gebru 2018). Nonetheless, they often form the basis for
-data-driven decision-making systems in real-world applications.
+\autocite{buolamwini2018gender}. Nonetheless, they often form the basis
+for data-driven decision-making systems in real-world applications.
 
@@
 stakeholders---that fail to understand the behaviour of their
-agent---i.e.~the black-box system (Borch 2022). The group of principals
-may include programmers, product managers and other decision-makers who
-develop and operate the system as well as those individuals ultimately
-subject to the decisions made by the system. In practice, decisions made
-by black-box systems are typically left unchallenged since the group of
-principals cannot scrutinize them:
+agent---i.e.~the black-box system \autocite{borch2022machine}. The group
+of principals may include programmers, product managers and other
+decision-makers who develop and operate the system as well as those
+individuals ultimately subject to the decisions made by the system. In
+practice, decisions made by black-box systems are typically left
+unchallenged since the group of principals cannot scrutinize them:
 
@@
 ``You cannot appeal to (algorithms). They do not listen. Nor do they
-bend.'' (O'Neil 2016)
+bend.'' \autocite{oneil2016weapons}
 \end{quote}
@@
 concepts. The following literature surveys provide more detail: Arrieta
-et al.~(2020) provide a broad overview of XAI (Arrieta et al. 2020); Fan
-et al.~(2020) focus on explainability in the context of deep learning
-(Fan, Xiong, and Wang 2020); and finally, Karimi et al.~(2020) (Karimi,
-Barthe, et al. 2020) and Verma et al.~(2020) Verma, Dickerson, and Hines
-(2020) offer detailed reviews of the literature on Counterfactual
-Explanations and Algorithmic Recourse (see also Molnar (2020) and
-Varshney (2022)). Miller (2019) explicitly discusses the concept of
-explainability from the perspective of a social scientist (Miller 2019).
+et al.~(2020) provide a broad overview of XAI
+\autocite{arrieta2020explainable}; Fan et al.~(2020) focus on
+explainability in the context of deep learning
+\autocite{fan2020interpretability}; and finally, Karimi et al.~(2020)
+\autocite{karimi2020survey} and Verma et al.~(2020)
+\textcite{verma2020counterfactual} offer detailed reviews of the
+literature on Counterfactual Explanations and Algorithmic Recourse (see
+also \textcite{molnar2020interpretable} and
+\textcite{varshney2022trustworthy}). Miller (2019) explicitly discusses
+the concept of explainability from the perspective of a social scientist
+\autocite{miller2019explanation}.
 
@@
 often used interchangeably, but this can lead to confusion. We find the
-distinction made in Rudin (2019) useful: Interpretable AI involves
-models that are inherently interpretable and transparent such as general
-additive models (GAM), decision trees and rule-based models; Explainable
-AI involves models that are not inherently interpretable but require
-additional tools to be explainable to humans. Examples of the latter
-include Ensembles, Support Vector Machines and Deep Neural Networks.
-Some would argue that we best avoid the second category of models
-altogether and instead focus solely on interpretable AI Rudin (2019).
-While we agree that initial efforts should always be geared towards
-interpretable models, avoiding black boxes altogether would entail
-missed opportunities and anyway is probably not very realistic at this
-point. For that reason, we expect the need for XAI to persist in the
-medium term. Explainable AI can further be broadly divided into
-\textbf{global} and \textbf{local} explainability: the former is
-concerned with explaining the average behaviour of a model, while the
-latter involves explanations for individual predictions (Molnar 2020).
-Tools for global explainability include partial dependence plots (PDP),
-which involve the computation of marginal effects through Monte Carlo,
-and global surrogates. A surrogate model is an interpretable model that
-is trained to explain the predictions of a black-box model.
+distinction made in \textcite{rudin2019stop} useful: Interpretable AI
+involves models that are inherently interpretable and transparent such
+as general additive models (GAM), decision trees and rule-based models;
+Explainable AI involves models that are not inherently interpretable but
+require additional tools to be explainable to humans. Examples of the
+latter include Ensembles, Support Vector Machines and Deep Neural
+Networks. Some would argue that we best avoid the second category of
+models altogether and instead focus solely on interpretable AI
+\textcite{rudin2019stop}. While we agree that initial efforts should
+always be geared towards interpretable models, avoiding black boxes
+altogether would entail missed opportunities and anyway is probably not
+very realistic at this point. For that reason, we expect the need for
+XAI to persist in the medium term. Explainable AI can further be broadly
+divided into \textbf{global} and \textbf{local} explainability: the
+former is concerned with explaining the average behaviour of a model,
+while the latter involves explanations for individual predictions
+\autocite{molnar2020interpretable}. Tools for global explainability
+include partial dependence plots (PDP), which involve the computation of
+marginal effects through Monte Carlo, and global surrogates. A surrogate
+model is an interpretable model that is trained to explain the
+predictions of a black-box model.
 
@@
 A recent study has shown that even experienced data scientists tend to
-put too much trust in explanations produced by LIME and SHAP (Kaur et
-al. 2020). Another recent work has shown that both methods can be easily
-fooled: they depend on random input perturbations, a property that can
-be abused by adverse agents to essentially whitewash strongly biased
-black-box models (Slack et al. 2020). In related work, the same authors
-find that while gradient-based Counterfactual Explanations can also be
-manipulated, there is a straightforward way to protect against this in
-practice (Slack et al. 2021). In the context of quality assessment, it
-is also worth noting that---contrary to surrogate explainers---CE always
-achieve full fidelity by construction: counterfactuals are searched with
-respect to the black-box classifier, not some proxy for it. That being
-said, CE should also be used with care and research around them is still
-in its early stages.
+put too much trust in explanations produced by LIME and SHAP
+\autocite{kaur2020interpreting}. Another recent work has shown that both
+methods can be easily fooled: they depend on random input perturbations,
+a property that can be abused by adverse agents to essentially whitewash
+strongly biased black-box models \autocite{slack2020fooling}. In related
+work, the same authors find that while gradient-based Counterfactual
+Explanations can also be manipulated, there is a straightforward way to
+protect against this in practice \autocite{slack2021counterfactual}. In
+the context of quality assessment, it is also worth noting
+that---contrary to surrogate explainers---CE always achieve full
+fidelity by construction: counterfactuals are searched with respect to
+the black-box classifier, not some proxy for it. That being said, CE
+should also be used with care and research around them is still in its
+early stages.
 
@@
 in understanding how we need to change individual attributes in order to
-change the model output to a desired value or label (Molnar 2020).
-Typically the underlying methodology is presented in the context of
-binary classification: \(M: \mathcal{X} \mapsto \mathcal{Y}\) where
+change the model output to a desired value or label
+\autocite{molnar2020interpretable}. Typically the underlying methodology
+is presented in the context of binary classification:
+\(M: \mathcal{X} \mapsto \mathcal{Y}\) where
 \(\mathcal{X}\subset\mathbb{R}^D\) and \(\mathcal{Y}=\{0,1\}\). Further,
@@
 that the ideas presented here also carry over to multi-class problems
-and regression (Molnar 2020).
-
-The counterfactual search objective originally proposed by Wachter,
-Mittelstadt, and Russell (2017) is as follows
+and regression \autocite{molnar2020interpretable}.
+
+The counterfactual search objective originally proposed by
+\textcite{wachter2017counterfactual} is as follows
 
@@
 Numerous extensions to this simple approach have been developed since CE
-were first proposed in 2017 (see Verma, Dickerson, and Hines (2020) and
-(Karimi, Barthe, et al. 2020) for surveys). The various approaches
-largely differ in that they use different flavours of search objective
-defined in Equation~\ref{eq-solution}. Different penalties are often
-used to address many of the desirable properties of effective CE that
-have been set out. These desiderata include: \textbf{proximity} --- the
-distance between factual and counterfactual features should be small
-(Wachter, Mittelstadt, and Russell 2017); \textbf{actionability} --- the
-proposed recourse should be actionable Poyiadzi et al. (2020);
-\textbf{plausibility} --- the counterfactual explanation should be
-plausible to a human Schut et al. (2021); \textbf{sparsity} --- the
-counterfactual explanation should involve as few individual feature
-changes as possible (Schut et al. 2021); \textbf{robustness} --- the
-counterfactual explanation should be robust to domain and model shifts
-(Upadhyay, Joshi, and Lakkaraju 2021); \textbf{diversity} --- ideally
-multiple diverse counterfactuals should be provided (Mothilal, Sharma,
-and Tan 2020); and \textbf{causality} --- counterfactuals should respect
-the structural causal model underlying the data generating process
-Karimi, Schölkopf, and Valera (2021).
+were first proposed in 2017 (see \textcite{verma2020counterfactual} and
+\textcite{karimi2020survey} for surveys). The various approaches largely
+differ in that they use different flavours of search objective defined
+in Equation~\ref{eq-solution}. Different penalties are often used to
+address many of the desirable properties of effective CE that have been
+set out. These desiderata include: \textbf{proximity} --- the distance
+between factual and counterfactual features should be small
+\autocite{wachter2017counterfactual}; \textbf{actionability} --- the
+proposed recourse should be actionable
+\autocite{ustun2019actionable,poyiadzi2020face}; \textbf{plausibility}
+--- the counterfactual explanation should be plausible to a human
+\autocite{joshi2019realistic,schut2021generating}; \textbf{sparsity} ---
+the counterfactual explanation should involve as few individual feature
+changes as possible \autocite{schut2021generating}; \textbf{robustness}
+--- the counterfactual explanation should be robust to domain and model
+shifts \autocite{upadhyay2021robust}; \textbf{diversity} --- ideally
+multiple diverse counterfactuals should be provided
+\autocite{mothilal2020explaining}; and \textbf{causality} ---
+counterfactuals should respect the structural causal model underlying
+the data generating process
+\autocite{karimi2020algorithmic,karimi2021algorithmic}.
 
@@
 \href{https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest}{CARLA}
-(Pawelczyk et al. 2021).\footnote{While we were writing this paper, the
-  \texttt{R} package \texttt{counterfactuals} was released (Dandl et al.
-  2023). The developers seem to also envision a unifying framework, but
-  the project appears to still be in its early stages.} In addition to
-that, there exists open-source code for some specific approaches to CE
-that have been proposed in recent years. The approach-specific
-implementations that we have been able to find are generally
-well-documented, but exclusively in Python. For example, a PyTorch
-implementation of a greedy generator for Bayesian models proposed in
-Schut et al. (2021) has been released. As another example, the popular
+\autocite{pawelczyk2021carla}.\footnote{While we were writing this
+  paper, the \texttt{R} package \texttt{counterfactuals} was released
+  \autocite{dandl2023counterfactuals}. The developers seem to also
+  envision a unifying framework, but the project appears to still be in
+  its early stages.} In addition to that, there exists open-source code
+for some specific approaches to CE that have been proposed in recent
+years. The approach-specific implementations that we have been able to
+find are generally well-documented, but exclusively in Python. For
+example, a PyTorch implementation of a greedy generator for Bayesian
+models proposed in \textcite{schut2021generating} has been released. As
+another example, the popular
 \href{https://github.com/interpretml}{InterpretML} library includes an
-implementation of a diverse counterfactual generator (Mothilal, Sharma,
-and Tan 2020).
+implementation of a diverse counterfactual generator
+\autocite{mothilal2020explaining}.
 
@@
 The package currently offers native support for models built and trained
-in \href{https://fluxml.ai/}{Flux} (Innes 2018) as well as a small
-subset of models made available through
-\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ} (Blaom
-et al. 2020). While in general it is assumed that users resort to this
-package to explain their pre-trained models, we provide a simple API
-call to train the following
+in \href{https://fluxml.ai/}{Flux} \autocite{innes2018flux} as well as a
+small subset of models made available through
+\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ}
+\autocite{blaom2020mlj}. While in general it is assumed that users
+resort to this package to explain their pre-trained models, we provide a
+simple API call to train the following
 \href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/}{models}:
@@
 \item
-  Deep Ensemble Lakshminarayanan, Pritzel, and Blundell (2016)
+  Deep Ensemble \textcite{lakshminarayanan2016simple}
 \item
@@
 embedding, existing work has typically relied on generative models or
-existing causal knowledge Karimi, Schölkopf, and Valera (2021). While
-this notion is compatible with all of our gradient-based generators,
-only some generators search a latent space by default. Finally,
-composability implies that the given generator can be blended with any
-other composable generator, which we discuss in
-Section~\ref{sec-gen-comp}.
+existing causal knowledge
+\autocite{joshi2019realistic,karimi2021algorithmic}. While this notion
+is compatible with all of our gradient-based generators, only some
+generators search a latent space by default. Finally, composability
+implies that the given generator can be blended with any other
+composable generator, which we discuss in Section~\ref{sec-gen-comp}.
 
@@
 \emph{ClapROAR} aims to preserve the classifier, i.e.~to generate
-counterfactuals that are robust to endogenous model shifts (Altmeyer et
-al. 2023); \emph{CLUE} searches plausible counterfactuals in the latent
-embedding of a generative model by explicitly minimising predictive
-entropy (Antorán et al. 2020); \emph{DiCE} is designed to generate
-multiple, maximally diverse counterfactuals (Mothilal, Sharma, and Tan
-2020); \emph{FeatureTweak} leverages the internals of decision trees to
-search counterfactuals on a feature-by-feature basis, finding the
-counterfactual that tweaks the features in the least costly way (Tolomei
-et al. 2017); \emph{Gravitational} aims to generate plausible and robust
+counterfactuals that are robust to endogenous model shifts
+\autocite{altmeyer2023endogenous}; \emph{CLUE} searches plausible
+counterfactuals in the latent embedding of a generative model by
+explicitly minimising predictive entropy \autocite{antoran2020getting};
+\emph{DiCE} is designed to generate multiple, maximally diverse
+counterfactuals \autocite{mothilal2020explaining}; \emph{FeatureTweak}
+leverages the internals of decision trees to search counterfactuals on a
+feature-by-feature basis, finding the counterfactual that tweaks the
+features in the least costly way \autocite{tolomei2017interpretable};
+\emph{Gravitational} aims to generate plausible and robust
 counterfactuals by minimising the distance to observed samples in the
-target class (Altmeyer et al. 2023); \emph{Greedy} aims to generate
-plausible counterfactuals by implicitly minimising predictive
-uncertainty of Bayesian classifiers (Schut et al. 2021);
+target class \autocite{altmeyer2023endogenous}; \emph{Greedy} aims to
+generate plausible counterfactuals by implicitly minimising predictive
+uncertainty of Bayesian classifiers \autocite{schut2021generating};
 \emph{GrowingSpheres} is model-agnostic, relying solely on identifying
@@
 increasing the search radius and then moving counterfactuals in that
-direction(Laugel et al. 2017); \emph{PROBE} generates probabilistically
-robust counterfactuals (Pawelczyk et al. 2022); \emph{REVISE} addresses
-the need for plausibility by searching counterfactuals in the latent
-embedding of a Variational Autoencoder (VAE) (Joshi et al. 2019);
-\emph{Wachter} is the baseline approach that only penalises the distance
-to the original sample (Wachter, Mittelstadt, and Russell 2017).
-
-\onecolumn
+direction\autocite{laugel2017inversea}; \emph{PROBE} generates
+probabilistically robust counterfactuals
+\autocite{pawelczyk2022probabilistically}; \emph{REVISE} addresses the
+need for plausibility by searching counterfactuals in the latent
+embedding of a Variational Autoencoder (VAE)
+\autocite{joshi2019realistic}; \emph{Wachter} is the baseline approach
+that only penalises the distance to the original sample
+\autocite{wachter2017counterfactual}.
 
@@
 \endlastfoot
-ClaPROAR (Altmeyer et al. 2023) & gradient based & feature & yes \\
-CLUE (Antorán et al. 2020) & gradient based & latent & yes \\
-DiCE (Mothilal, Sharma, and Tan 2020) & gradient based & feature &
+ClaPROAR \autocite{altmeyer2023endogenous} & gradient based & feature &
 yes \\
-FeatureTweak (Tolomei et al. 2017) & tree based & feature & no \\
-Gravitational (Altmeyer et al. 2023) & gradient based & feature & yes \\
-Greedy (Schut et al. 2021) & gradient based & feature & yes \\
-GrowingSpheres (Laugel et al. 2017) & agnostic & feature & no \\
-PROBE (Pawelczyk et al. 2022) & gradient based & feature & no \\
-REVISE (Joshi et al. 2019) & gradient based & latent & yes \\
-Wachter (Wachter, Mittelstadt, and Russell 2017) & gradient based &
+CLUE \autocite{antoran2020getting} & gradient based & latent & yes \\
+DiCE \autocite{mothilal2020explaining} & gradient based & feature &
+yes \\
+FeatureTweak \autocite{tolomei2017interpretable} & tree based & feature
+& no \\
+Gravitational \autocite{altmeyer2023endogenous} & gradient based &
 feature & yes \\
+Greedy \autocite{schut2021generating} & gradient based & feature &
+yes \\
+GrowingSpheres \autocite{laugel2017inversea} & agnostic & feature &
+no \\
+PROBE \autocite{pawelczyk2022probabilistically} & gradient based &
+feature & no \\
+REVISE \autocite{joshi2019realistic} & gradient based & latent & yes \\
+Wachter \autocite{wachter2017counterfactual} & gradient based & feature
+& yes \\
 \end{longtable}
-
-\twocolumn
 
@@
 \item
-  Adult Census (Barry Becker 1996)
+  Adult Census \autocite{becker1996adult}
 \item
-  California Housing (Pace and Barry 1997)
+  California Housing \autocite{pace1997sparse}
 \item
-  CIFAR10 (Krizhevsky 2009)
+  CIFAR10 \autocite{krizhevsky2009learning}
 \item
-  German Credit (Hoffman 1994)
+  German Credit \autocite{hoffman1994german}
 \item
-  Give Me Some Credit (Kaggle 2011)
+  Give Me Some Credit \autocite{kaggle2011give}
 \item
-  MNIST (LeCun 1998) and Fashion MNIST (Xiao, Rasul, and Vollgraf 2017)
+  MNIST \autocite{lecun1998mnist} and Fashion MNIST
+  \autocite{xiao2017fashion}
 \item
-  UCI defaultCredit (Yeh and Lien 2009)
+  UCI defaultCredit \autocite{yeh2009comparisons}
 \end{itemize}
@@
 \(\mathcal{X}\) and \(\text{diversity}(\cdot)\) is the penalty proposed
-by Mothilal, Sharma, and Tan (2020) that induces diverse sets of
+by \textcite{mothilal2020explaining} that induces diverse sets of
 counterfactuals. As in Equation~\ref{eq-solution}, \(\ell\) is the loss
@@
 counterfactual from its original value. The API call to compute the
-distance metric defined in Wachter, Mittelstadt, and Russell (2017), for
+distance metric defined in \textcite{wachter2017counterfactual}, for
 instance, is as simple as \texttt{evaluate(ce;\ measure=distance\_mad)},
@@
 how native support for \href{https://fluxml.ai/}{\texttt{Flux.jl}}
-(Innes 2018) deep learning models was enabled.\footnote{Flux models are
-  now natively supported by our package and can be instantiated by
-  calling \texttt{FluxModel()}.} Once again we use synthetic data for an
-illustrative example. Code \ref{lst:nn} below builds a simple model
-architecture that can be used for a multi-class prediction task. Note
-how outputs from the final layer are not passed through a softmax
-activation function, since the counterfactual loss is evaluated with
-respect to logits as we discussed earlier. The model is trained with
-dropout.
+\autocite{innes2018flux} deep learning models was enabled.\footnote{Flux
+  models are now natively supported by our package and can be
+  instantiated by calling \texttt{FluxModel()}.} Once again we use
+synthetic data for an illustrative example. Code \ref{lst:nn} below
+builds a simple model architecture that can be used for a multi-class
+prediction task. Note how outputs from the final layer are not passed
+through a softmax activation function, since the counterfactual loss is
+evaluated with respect to logits as we discussed earlier. The model is
+trained with dropout.
 
@@
 quantifying the uncertainty around the generated counterfactuals
-(Delaney, Greene, and Keane 2021). One idea could be, to use dropout to
+\autocite{delaney2021uncertainty}. One idea could be, to use dropout to
 randomly switch features on and off in each iteration. Without dwelling
@@
 The \emph{Give Me Some Credit} dataset is one of the tabular real-world
-datasets that ship with the package (Kaggle 2011). It can be used to
-train a binary classifier to predict whether a borrower is likely to
-experience financial difficulties in the next two years. In particular,
-we have an output variable
+datasets that ship with the package \autocite{kaggle2011give}. It can be
+used to train a binary classifier to predict whether a borrower is
+likely to experience financial difficulties in the next two years. In
+particular, we have an output variable
 \(y \in \{0=\texttt{no stress},1=\texttt{stress}\}\) and a feature
@@
 For the classification task, we use a Multi-Layer Perceptron with
-dropout regularization. Using the Gravitational generator (Altmeyer et
-al. 2023) we will generate counterfactuals for ten randomly chosen
-individuals that would be denied credit based on our pre-trained model.
-Concerning the mutability of features, we only impose that the
-\texttt{age} cannot be decreased.
+dropout regularization. Using the Gravitational generator
+\autocite{altmeyer2023endogenous} we will generate counterfactuals for
+ten randomly chosen individuals that would be denied credit based on our
+pre-trained model. Concerning the mutability of features, we only impose
+that the \texttt{age} cannot be decreased.
 
@@
 contains 60,000 training samples of handwritten digits in the form of
-28x28 pixel grey-scale images (LeCun 1998). Each image is associated
-with a label indicating the digit (0-9) that the image represents. The
-data makes for an interesting case study of CE because humans have a
-good idea of what plausible counterfactuals of digits look like. For
-example, if you were asked to pick up an eraser and turn the digit in
-the left panel of Figure~\ref{fig-mnist} into a four (4) you would know
-exactly what to do: just erase the top part.
+28x28 pixel grey-scale images \autocite{lecun1998mnist}. Each image is
+associated with a label indicating the digit (0-9) that the image
+represents. The data makes for an interesting case study of CE because
+humans have a good idea of what plausible counterfactuals of digits look
+like. For example, if you were asked to pick up an eraser and turn the
+digit in the left panel of Figure~\ref{fig-mnist} into a four (4) you
+would know exactly what to do: just erase the top part.
 
@@
 supervised models built in \texttt{MLJ.jl}, an extensive Machine
-Learning framework for Julia (Blaom et al. 2020). This may also involve
-adding support for regression models as well as additional
+Learning framework for Julia \autocite{blaom2020mlj}. This may also
+involve adding support for regression models as well as additional
 non-differentiable models. In terms of counterfactual generators, there
 is a list of recent methodologies that we would like to implement
-including MINT (Karimi, Schölkopf, and Valera 2021), ROAR (Upadhyay,
-Joshi, and Lakkaraju 2021) and FACE (Poyiadzi et al. 2020).
+including MINT \autocite{karimi2021algorithmic}, ROAR
+\autocite{upadhyay2021robust} and FACE \autocite{poyiadzi2020face}.
 
@@
 explanations including \emph{Adult}, \emph{Give Me Some Credit} and
-\emph{German Credit} (Karimi, Barthe, et al. 2020). There is scope for
+\emph{German Credit} \autocite{karimi2020survey}. There is scope for
 adding data sources that have so far not been explored much in this
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-altmeyer2023endogenous}{}}%
-Altmeyer, Patrick, Giovan Angela, Aleksander Buszydlik, Karol Dobiczek,
-Arie van Deursen, and Cynthia Liem. 2023. {``Endogenous {Macrodynamics}
-in {Algorithmic} {Recourse}.''} In \emph{First {IEEE} {Conference} on
-{Secure} and {Trustworthy} {Machine} {Learning}}.
-\url{https://doi.org/10.1109/satml54575.2023.00036}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-antoran2020getting}{}}%
-Antorán, Javier, Umang Bhatt, Tameem Adel, Adrian Weller, and José
-Miguel Hernández-Lobato. 2020. {``Getting a Clue: {A} Method for
-Explaining Uncertainty Estimates.''}
-\url{https://arxiv.org/abs/2006.06848}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-arrieta2020explainable}{}}%
-Arrieta, Alejandro Barredo, Natalia Diaz-Rodriguez, Javier Del Ser,
-Adrien Bennetot, Siham Tabik, Alberto Barbado, Salvador Garcia, et al.
-2020. {``Explainable {Artificial Intelligence} ({XAI}): {Concepts},
-Taxonomies, Opportunities and Challenges Toward Responsible {AI}.''}
-\emph{Information Fusion} 58: 82--115.
-\url{https://doi.org/10.1016/j.inffus.2019.12.012}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-becker1996adult}{}}%
-Barry Becker, Ronny Kohavi. 1996. {``Adult.''} UCI Machine Learning
-Repository. \url{https://doi.org/10.24432/C5XW20}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-blaom2020mlj}{}}%
-Blaom, Anthony D., Franz Kiraly, Thibaut Lienart, Yiannis Simillides,
-Diego Arenas, and Sebastian J. Vollmer. 2020. {``{MLJ}: {A Julia}
-Package for Composable Machine Learning.''} \emph{Journal of Open Source
-Software} 5 (55): 2704. \url{https://doi.org/10.21105/joss.02704}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-borch2022machine}{}}%
-Borch, Christian. 2022. {``Machine Learning, Knowledge Risk, and
-Principal-Agent Problems in Automated Trading.''} \emph{Technology in
-Society}, 101852. \url{https://doi.org/10.1016/j.techsoc.2021.101852}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-buolamwini2018gender}{}}%
-Buolamwini, Joy, and Timnit Gebru. 2018. {``Gender Shades:
-{Intersectional} Accuracy Disparities in Commercial Gender
-Classification.''} In \emph{Conference on Fairness, Accountability and
-Transparency}, 77--91. {PMLR}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-dandl2023counterfactuals}{}}%
-Dandl, Susanne, Andreas Hofheinz, Martin Binder, Bernd Bischl, and
-Giuseppe Casalicchio. 2023. {``Counterfactuals: {An} {R} {Package} for
-{Counterfactual} {Explanation} {Methods}.''} arXiv.
-\url{http://arxiv.org/abs/2304.06569}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-delaney2021uncertainty}{}}%
-Delaney, Eoin, Derek Greene, and Mark T. Keane. 2021. {``Uncertainty
-{Estimation} and {Out}-of-{Distribution} {Detection} for
-{Counterfactual} {Explanations}: {Pitfalls} and {Solutions}.''} arXiv.
-\url{http://arxiv.org/abs/2107.09734}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fan2020interpretability}{}}%
-Fan, Fenglei, Jinjun Xiong, and Ge Wang. 2020. {``On Interpretability of
-Artificial Neural Networks.''} \url{https://arxiv.org/abs/2001.02522}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-goodfellow2014explaining}{}}%
-Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
-{``Explaining and Harnessing Adversarial Examples.''}
-\url{https://arxiv.org/abs/1412.6572}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hoffman1994german}{}}%
-Hoffman, Hans. 1994. {``German {Credit Data}.''}
-\url{https://archive.ics.uci.edu/ml/datasets/statlog+(german+credit+data)}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-innes2018flux}{}}%
-Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
-\emph{Journal of Open Source Software} 3 (25): 602.
-\url{https://doi.org/10.21105/joss.00602}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-joshi2019realistic}{}}%
-Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
-Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
-Actionable Explanations in Black-Box Decision Making Systems.''}
-\url{https://arxiv.org/abs/1907.09615}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kaggle2011give}{}}%
-Kaggle. 2011. {``Give Me Some Credit, {Improve} on the State of the Art
-in Credit Scoring by Predicting the Probability That Somebody Will
-Experience Financial Distress in the Next Two Years.''} {Kaggle}.
-\url{https://www.kaggle.com/c/GiveMeSomeCredit}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2020survey}{}}%
-Karimi, Amir-Hossein, Gilles Barthe, Bernhard Schölkopf, and Isabel
-Valera. 2020. {``A Survey of Algorithmic Recourse: Definitions,
-Formulations, Solutions, and Prospects.''}
-\url{https://arxiv.org/abs/2010.04050}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2021algorithmic}{}}%
-Karimi, Amir-Hossein, Bernhard Schölkopf, and Isabel Valera. 2021.
-{``Algorithmic Recourse: From Counterfactual Explanations to
-Interventions.''} In \emph{Proceedings of the 2021 {ACM Conference} on
-{Fairness}, {Accountability}, and {Transparency}}, 353--62.
-
-\leavevmode\vadjust pre{\hypertarget{ref-karimi2020algorithmic}{}}%
-Karimi, Amir-Hossein, Julius Von Kügelgen, Bernhard Schölkopf, and
-Isabel Valera. 2020. {``Algorithmic Recourse Under Imperfect Causal
-Knowledge: A Probabilistic Approach.''}
-\url{https://arxiv.org/abs/2006.06831}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kaur2020interpreting}{}}%
-Kaur, Harmanpreet, Harsha Nori, Samuel Jenkins, Rich Caruana, Hanna
-Wallach, and Jennifer Wortman Vaughan. 2020. {``Interpreting
-Interpretability: Understanding Data Scientists' Use of Interpretability
-Tools for Machine Learning.''} In \emph{Proceedings of the 2020 {CHI}
-Conference on Human Factors in Computing Systems}, 1--14.
-\url{https://doi.org/10.1145/3313831.3376219}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-krizhevsky2009learning}{}}%
-Krizhevsky, A. 2009. {``Learning {Multiple} {Layers} of {Features} from
-{Tiny} {Images}.''} In.
-\url{https://www.semanticscholar.org/paper/Learning-Multiple-Layers-of-Features-from-Tiny-Krizhevsky/5d90f06bb70a0a3dced62413346235c02b1aa086}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lakshminarayanan2016simple}{}}%
-Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2016.
-{``Simple and Scalable Predictive Uncertainty Estimation Using Deep
-Ensembles.''} \url{https://arxiv.org/abs/1612.01474}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-laugel2017inversea}{}}%
-Laugel, Thibault, Marie-Jeanne Lesot, Christophe Marsala, Xavier Renard,
-and Marcin Detyniecki. 2017. {``Inverse {Classification} for
-{Comparison}-Based {Interpretability} in {Machine} {Learning}.''} arXiv.
-\url{https://doi.org/10.48550/arXiv.1712.08443}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lecun1998mnist}{}}%
-LeCun, Yann. 1998. {``The {MNIST} Database of Handwritten Digits.''}
-
-\leavevmode\vadjust pre{\hypertarget{ref-miller2019explanation}{}}%
-Miller, Tim. 2019. {``Explanation in Artificial Intelligence: {Insights}
-from the Social Sciences.''} \emph{Artificial Intelligence} 267: 1--38.
-\url{https://doi.org/10.1016/j.artint.2018.07.007}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-molnar2020interpretable}{}}%
-Molnar, Christoph. 2020. \emph{Interpretable Machine Learning}. {Lulu.
-com}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mothilal2020explaining}{}}%
-Mothilal, Ramaravind K, Amit Sharma, and Chenhao Tan. 2020.
-{``Explaining Machine Learning Classifiers Through Diverse
-Counterfactual Explanations.''} In \emph{Proceedings of the 2020
-{Conference} on {Fairness}, {Accountability}, and {Transparency}},
-607--17. \url{https://doi.org/10.1145/3351095.3372850}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-oneil2016weapons}{}}%
-O'Neil, Cathy. 2016. \emph{Weapons of Math Destruction: {How} Big Data
-Increases Inequality and Threatens Democracy}. {Crown}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pace1997sparse}{}}%
-Pace, R Kelley, and Ronald Barry. 1997. {``Sparse Spatial
-Autoregressions.''} \emph{Statistics \& Probability Letters} 33 (3):
-291--97. \url{https://doi.org/10.1016/s0167-7152(96)00140-x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2021carla}{}}%
-Pawelczyk, Martin, Sascha Bielawski, Johannes van den Heuvel, Tobias
-Richter, and Gjergji Kasneci. 2021. {``Carla: A Python Library to
-Benchmark Algorithmic Recourse and Counterfactual Explanation
-Algorithms.''} \url{https://arxiv.org/abs/2108.00783}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pawelczyk2022probabilistically}{}}%
-Pawelczyk, Martin, Teresa Datta, Johannes van-den-Heuvel, Gjergji
-Kasneci, and Himabindu Lakkaraju. 2022. {``Probabilistically {Robust}
-{Recourse}: {Navigating} the {Trade}-Offs Between {Costs} and
-{Robustness} in {Algorithmic} {Recourse}.''} \emph{arXiv Preprint
-arXiv:2203.06768}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-poyiadzi2020face}{}}%
-Poyiadzi, Rafael, Kacper Sokol, Raul Santos-Rodriguez, Tijl De Bie, and
-Peter Flach. 2020. {``{FACE}: {Feasible} and Actionable Counterfactual
-Explanations.''} In \emph{Proceedings of the {AAAI}/{ACM Conference} on
-{AI}, {Ethics}, and {Society}}, 344--50.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rudin2019stop}{}}%
-Rudin, Cynthia. 2019. {``Stop Explaining Black Box Machine Learning
-Models for High Stakes Decisions and Use Interpretable Models
-Instead.''} \emph{Nature Machine Intelligence} 1 (5): 206--15.
-\url{https://doi.org/10.1038/s42256-019-0048-x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schut2021generating}{}}%
-Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
-Sacaleanu, Yarin Gal, et al. 2021. {``Generating {Interpretable
-Counterfactual Explanations By Implicit Minimisation} of {Epistemic} and
-{Aleatoric Uncertainties}.''} In \emph{International {Conference} on
-{Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-slack2021counterfactual}{}}%
-Slack, Dylan, Anna Hilgard, Himabindu Lakkaraju, and Sameer Singh. 2021.
-{``Counterfactual Explanations Can Be Manipulated.''} \emph{Advances in
-Neural Information Processing Systems} 34.
-
-\leavevmode\vadjust pre{\hypertarget{ref-slack2020fooling}{}}%
-Slack, Dylan, Sophie Hilgard, Emily Jia, Sameer Singh, and Himabindu
-Lakkaraju. 2020. {``Fooling Lime and Shap: {Adversarial} Attacks on Post
-Hoc Explanation Methods.''} In \emph{Proceedings of the {AAAI}/{ACM
-Conference} on {AI}, {Ethics}, and {Society}}, 180--86.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tolomei2017interpretable}{}}%
-Tolomei, Gabriele, Fabrizio Silvestri, Andrew Haines, and Mounia Lalmas.
-2017. {``Interpretable {Predictions} of {Tree}-Based {Ensembles} via
-{Actionable} {Feature} {Tweaking}.''} In \emph{Proceedings of the 23rd
-{ACM} {SIGKDD} {International} {Conference} on {Knowledge} {Discovery}
-and {Data} {Mining}}, 465--74.
-\url{https://doi.org/10.1145/3097983.3098039}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-upadhyay2021robust}{}}%
-Upadhyay, Sohini, Shalmali Joshi, and Himabindu Lakkaraju. 2021.
-{``Towards {Robust} and {Reliable Algorithmic Recourse}.''}
-\url{https://arxiv.org/abs/2102.13620}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ustun2019actionable}{}}%
-Ustun, Berk, Alexander Spangher, and Yang Liu. 2019. {``Actionable
-Recourse in Linear Classification.''} In \emph{Proceedings of the
-{Conference} on {Fairness}, {Accountability}, and {Transparency}},
-10--19. \url{https://doi.org/10.1145/3287560.3287566}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-varshney2022trustworthy}{}}%
-Varshney, Kush R. 2022. \emph{Trustworthy {Machine Learning}}.
-{Chappaqua, NY, USA}: {Independently Published}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-verma2020counterfactual}{}}%
-Verma, Sahil, John Dickerson, and Keegan Hines. 2020. {``Counterfactual
-Explanations for Machine Learning: {A} Review.''}
-\url{https://arxiv.org/abs/2010.10596}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wachter2017counterfactual}{}}%
-Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
-{``Counterfactual Explanations Without Opening the Black Box:
-{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
-841. \url{https://doi.org/10.2139/ssrn.3063289}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xiao2017fashion}{}}%
-Xiao, Han, Kashif Rasul, and Roland Vollgraf. 2017. {``Fashion-{MNIST}:
-A {Novel} {Image} {Dataset} for {Benchmarking} {Machine} {Learning}
-{Algorithms}.''} arXiv. \url{https://doi.org/10.48550/arXiv.1708.07747}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yeh2009comparisons}{}}%
-Yeh, I-Cheng, and Che-hui Lien. 2009. {``The Comparisons of Data Mining
-Techniques for the Predictive Accuracy of Probability of Default of
-Credit Card Clients.''} \emph{Expert Systems with Applications} 36 (2):
-2473--80. \url{https://doi.org/10.1016/j.eswa.2007.12.020}.
-
-\end{CSLReferences}
+\printbibliography[title=References]
 

### 2024-01-07 examples/CounterfactualExplanations.jl/paper.tex
link and citecolor sorted

@@
   colorlinks=true,
-  linkcolor={blue},
+  linkcolor={black},
   filecolor={Maroon},
-  citecolor={Blue},
+  citecolor={black},
   urlcolor={Blue},
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 

### 2024-01-07 examples/CounterfactualExplanations.jl/paper.tex
cant figure out how to get border colors for links

--- 2023-11-17 examples/CounterfactualExplanations.jl/paper.tex
@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
@@
 Intelligence, Counterfactual Explanations, Algorithmic Recourse},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
+  hidelinks,
   pdfcreator={LaTeX via pandoc}}
@@
 
-We present \texttt{CounterfactualExplanations.jl}: a package for
-generating Counterfactual Explanations (CE) and Algorithmic Recourse
-(AR) for black-box models in Julia. CE explain how inputs into a model
-need to change to yield specific model predictions. Explanations that
-involve realistic and actionable changes can be used to provide AR: a
-set of proposed actions for individuals to change an undesirable outcome
-for the better. In this article, we discuss the usefulness of CE for
-Explainable Artificial Intelligence and demonstrate the functionality of
-our package. The package is straightforward to use and designed with a
-focus on customization and extensibility. We envision it to one day be
-the go-to place for explaining arbitrary predictive models in Julia
-through a diverse suite of counterfactual generators.
+We present
+\href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}}:
+a package for generating Counterfactual Explanations (CE) and
+Algorithmic Recourse (AR) for black-box models in Julia. CE explain how
+inputs into a model need to change to yield specific model predictions.
+Explanations that involve realistic and actionable changes can be used
+to provide AR: a set of proposed actions for individuals to change an
+undesirable outcome for the better. In this article, we discuss the
+usefulness of CE for Explainable Artificial Intelligence and demonstrate
+the functionality of our package. The package is straightforward to use
+and designed with a focus on customization and extensibility. We
+envision it to one day be the go-to place for explaining arbitrary
+predictive models in Julia through a diverse suite of counterfactual
+generators.
 \end{abstract}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 

### 2024-01-07 examples/CounterfactualExplanations.jl/paper.qmd

@@
 abstract: |
-    We present `CounterfactualExplanations.jl`: a package for generating Counterfactual Explanations (CE) and Algorithmic Recourse (AR) for black-box models in Julia.
+    We present [`CounterfactualExplanations.jl`](https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl): a package for generating Counterfactual Explanations (CE) and Algorithmic Recourse (AR) for black-box models in Julia.
 CE explain how inputs into a model need to change to yield specific model predictions.

### 2024-01-07 examples/CounterfactualExplanations.jl/paper.tex

@@
 % Remove whitespace after paragraphs
-\setlength{\parskip}{0pt}
-
-\bibliographystyle{juliacon}
+\setlength{\parskip}{-1pt}
 \usepackage{orcidlink}
@@
 \fi
-\usepackage[]{biblatex}
-\addbibresource{ref.bib}
+\usepackage[]{natbib}
+\bibliographystyle{juliacon.bst}
 \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 problems typically associated with these models: they tend to be
-unstable \autocite{goodfellow2014explaining}, encode existing biases
-\autocite{buolamwini2018gender} and learn representations that are
+unstable \citep{goodfellow2014explaining}, encode existing biases
+\citep{buolamwini2018gender} and learn representations that are
 surprising or even counter-intuitive from a human perspective
-\autocite{buolamwini2018gender}. Nonetheless, they often form the basis
-for data-driven decision-making systems in real-world applications.
+\citep{buolamwini2018gender}. Nonetheless, they often form the basis for
+data-driven decision-making systems in real-world applications.
 
@@
 stakeholders---that fail to understand the behaviour of their
-agent---i.e.~the black-box system \autocite{borch2022machine}. The group
-of principals may include programmers, product managers and other
+agent---i.e.~the black-box system \citep{borch2022machine}. The group of
+principals may include programmers, product managers and other
 decision-makers who develop and operate the system as well as those
@@
 ``You cannot appeal to (algorithms). They do not listen. Nor do they
-bend.'' \autocite{oneil2016weapons}
+bend.'' \citep{oneil2016weapons}
 \end{quote}
@@
 et al.~(2020) provide a broad overview of XAI
-\autocite{arrieta2020explainable}; Fan et al.~(2020) focus on
+\citep{arrieta2020explainable}; Fan et al.~(2020) focus on
 explainability in the context of deep learning
-\autocite{fan2020interpretability}; and finally, Karimi et al.~(2020)
-\autocite{karimi2020survey} and Verma et al.~(2020)
-\textcite{verma2020counterfactual} offer detailed reviews of the
-literature on Counterfactual Explanations and Algorithmic Recourse (see
-also \textcite{molnar2020interpretable} and
-\textcite{varshney2022trustworthy}). Miller (2019) explicitly discusses
-the concept of explainability from the perspective of a social scientist
-\autocite{miller2019explanation}.
+\citep{fan2020interpretability}; and finally, Karimi et al.~(2020)
+\citep{karimi2020survey} and Verma et al.~(2020)
+\citet{verma2020counterfactual} offer detailed reviews of the literature
+on Counterfactual Explanations and Algorithmic Recourse (see also
+\citet{molnar2020interpretable} and \citet{varshney2022trustworthy}).
+Miller (2019) explicitly discusses the concept of explainability from
+the perspective of a social scientist \citep{miller2019explanation}.
 
@@
 often used interchangeably, but this can lead to confusion. We find the
-distinction made in \textcite{rudin2019stop} useful: Interpretable AI
+distinction made in \citet{rudin2019stop} useful: Interpretable AI
 involves models that are inherently interpretable and transparent such
@@
 models altogether and instead focus solely on interpretable AI
-\textcite{rudin2019stop}. While we agree that initial efforts should
-always be geared towards interpretable models, avoiding black boxes
-altogether would entail missed opportunities and anyway is probably not
-very realistic at this point. For that reason, we expect the need for
-XAI to persist in the medium term. Explainable AI can further be broadly
+\citet{rudin2019stop}. While we agree that initial efforts should always
+be geared towards interpretable models, avoiding black boxes altogether
+would entail missed opportunities and anyway is probably not very
+realistic at this point. For that reason, we expect the need for XAI to
+persist in the medium term. Explainable AI can further be broadly
 divided into \textbf{global} and \textbf{local} explainability: the
@@
 while the latter involves explanations for individual predictions
-\autocite{molnar2020interpretable}. Tools for global explainability
-include partial dependence plots (PDP), which involve the computation of
+\citep{molnar2020interpretable}. Tools for global explainability include
+partial dependence plots (PDP), which involve the computation of
 marginal effects through Monte Carlo, and global surrogates. A surrogate
@@
 put too much trust in explanations produced by LIME and SHAP
-\autocite{kaur2020interpreting}. Another recent work has shown that both
+\citep{kaur2020interpreting}. Another recent work has shown that both
 methods can be easily fooled: they depend on random input perturbations,
 a property that can be abused by adverse agents to essentially whitewash
-strongly biased black-box models \autocite{slack2020fooling}. In related
+strongly biased black-box models \citep{slack2020fooling}. In related
 work, the same authors find that while gradient-based Counterfactual
 Explanations can also be manipulated, there is a straightforward way to
-protect against this in practice \autocite{slack2021counterfactual}. In
-the context of quality assessment, it is also worth noting
-that---contrary to surrogate explainers---CE always achieve full
-fidelity by construction: counterfactuals are searched with respect to
-the black-box classifier, not some proxy for it. That being said, CE
-should also be used with care and research around them is still in its
-early stages.
+protect against this in practice \citep{slack2021counterfactual}. In the
+context of quality assessment, it is also worth noting that---contrary
+to surrogate explainers---CE always achieve full fidelity by
+construction: counterfactuals are searched with respect to the black-box
+classifier, not some proxy for it. That being said, CE should also be
+used with care and research around them is still in its early stages.
 
@@
 change the model output to a desired value or label
-\autocite{molnar2020interpretable}. Typically the underlying methodology
-is presented in the context of binary classification:
+\citep{molnar2020interpretable}. Typically the underlying methodology is
+presented in the context of binary classification:
 \(M: \mathcal{X} \mapsto \mathcal{Y}\) where
@@
 that the ideas presented here also carry over to multi-class problems
-and regression \autocite{molnar2020interpretable}.
+and regression \citep{molnar2020interpretable}.
 
 The counterfactual search objective originally proposed by
-\textcite{wachter2017counterfactual} is as follows
+\citet{wachter2017counterfactual} is as follows
 
@@
 Numerous extensions to this simple approach have been developed since CE
-were first proposed in 2017 (see \textcite{verma2020counterfactual} and
-\textcite{karimi2020survey} for surveys). The various approaches largely
+were first proposed in 2017 (see \citet{verma2020counterfactual} and
+\citet{karimi2020survey} for surveys). The various approaches largely
 differ in that they use different flavours of search objective defined
@@
 between factual and counterfactual features should be small
-\autocite{wachter2017counterfactual}; \textbf{actionability} --- the
+\citep{wachter2017counterfactual}; \textbf{actionability} --- the
 proposed recourse should be actionable
-\autocite{ustun2019actionable,poyiadzi2020face}; \textbf{plausibility}
---- the counterfactual explanation should be plausible to a human
-\autocite{joshi2019realistic,schut2021generating}; \textbf{sparsity} ---
+\citep{ustun2019actionable, poyiadzi2020face}; \textbf{plausibility} ---
+the counterfactual explanation should be plausible to a human
+\citep{joshi2019realistic, schut2021generating}; \textbf{sparsity} ---
 the counterfactual explanation should involve as few individual feature
-changes as possible \autocite{schut2021generating}; \textbf{robustness}
---- the counterfactual explanation should be robust to domain and model
-shifts \autocite{upadhyay2021robust}; \textbf{diversity} --- ideally
+changes as possible \citep{schut2021generating}; \textbf{robustness} ---
+the counterfactual explanation should be robust to domain and model
+shifts \citep{upadhyay2021robust}; \textbf{diversity} --- ideally
 multiple diverse counterfactuals should be provided
-\autocite{mothilal2020explaining}; and \textbf{causality} ---
+\citep{mothilal2020explaining}; and \textbf{causality} ---
 counterfactuals should respect the structural causal model underlying
 the data generating process
-\autocite{karimi2020algorithmic,karimi2021algorithmic}.
+\citep{karimi2020algorithmic, karimi2021algorithmic}.
 
@@
 \href{https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest}{CARLA}
-\autocite{pawelczyk2021carla}.\footnote{While we were writing this
-  paper, the \texttt{R} package \texttt{counterfactuals} was released
-  \autocite{dandl2023counterfactuals}. The developers seem to also
-  envision a unifying framework, but the project appears to still be in
-  its early stages.} In addition to that, there exists open-source code
-for some specific approaches to CE that have been proposed in recent
-years. The approach-specific implementations that we have been able to
-find are generally well-documented, but exclusively in Python. For
-example, a PyTorch implementation of a greedy generator for Bayesian
-models proposed in \textcite{schut2021generating} has been released. As
-another example, the popular
-\href{https://github.com/interpretml}{InterpretML} library includes an
-implementation of a diverse counterfactual generator
-\autocite{mothilal2020explaining}.
+\citep{pawelczyk2021carla}.\footnote{While we were writing this paper,
+  the \texttt{R} package \texttt{counterfactuals} was released
+  \citep{dandl2023counterfactuals}. The developers seem to also envision
+  a unifying framework, but the project appears to still be in its early
+  stages.} In addition to that, there exists open-source code for some
+specific approaches to CE that have been proposed in recent years. The
+approach-specific implementations that we have been able to find are
+generally well-documented, but exclusively in Python. For example, a
+PyTorch implementation of a greedy generator for Bayesian models
+proposed in \citet{schut2021generating} has been released. As another
+example, the popular \href{https://github.com/interpretml}{InterpretML}
+library includes an implementation of a diverse counterfactual generator
+\citep{mothilal2020explaining}.
 
@@
 The package currently offers native support for models built and trained
-in \href{https://fluxml.ai/}{Flux} \autocite{innes2018flux} as well as a
+in \href{https://fluxml.ai/}{Flux} \citep{innes2018flux} as well as a
 small subset of models made available through
 \href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ}
-\autocite{blaom2020mlj}. While in general it is assumed that users
-resort to this package to explain their pre-trained models, we provide a
-simple API call to train the following
+\citep{blaom2020mlj}. While in general it is assumed that users resort
+to this package to explain their pre-trained models, we provide a simple
+API call to train the following
 \href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/}{models}:
@@
 \item
-  Deep Ensemble \textcite{lakshminarayanan2016simple}
+  Deep Ensemble \citet{lakshminarayanan2016simple}
 \item
@@
 existing causal knowledge
-\autocite{joshi2019realistic,karimi2021algorithmic}. While this notion
-is compatible with all of our gradient-based generators, only some
+\citep{joshi2019realistic, karimi2021algorithmic}. While this notion is
+compatible with all of our gradient-based generators, only some
 generators search a latent space by default. Finally, composability
@@
 counterfactuals that are robust to endogenous model shifts
-\autocite{altmeyer2023endogenous}; \emph{CLUE} searches plausible
+\citep{altmeyer2023endogenous}; \emph{CLUE} searches plausible
 counterfactuals in the latent embedding of a generative model by
-explicitly minimising predictive entropy \autocite{antoran2020getting};
+explicitly minimising predictive entropy \citep{antoran2020getting};
 \emph{DiCE} is designed to generate multiple, maximally diverse
-counterfactuals \autocite{mothilal2020explaining}; \emph{FeatureTweak}
+counterfactuals \citep{mothilal2020explaining}; \emph{FeatureTweak}
 leverages the internals of decision trees to search counterfactuals on a
 feature-by-feature basis, finding the counterfactual that tweaks the
-features in the least costly way \autocite{tolomei2017interpretable};
+features in the least costly way \citep{tolomei2017interpretable};
 \emph{Gravitational} aims to generate plausible and robust
 counterfactuals by minimising the distance to observed samples in the
-target class \autocite{altmeyer2023endogenous}; \emph{Greedy} aims to
+target class \citep{altmeyer2023endogenous}; \emph{Greedy} aims to
 generate plausible counterfactuals by implicitly minimising predictive
-uncertainty of Bayesian classifiers \autocite{schut2021generating};
+uncertainty of Bayesian classifiers \citep{schut2021generating};
 \emph{GrowingSpheres} is model-agnostic, relying solely on identifying
@@
 increasing the search radius and then moving counterfactuals in that
-direction\autocite{laugel2017inversea}; \emph{PROBE} generates
+direction\citep{laugel2017inversea}; \emph{PROBE} generates
 probabilistically robust counterfactuals
-\autocite{pawelczyk2022probabilistically}; \emph{REVISE} addresses the
-need for plausibility by searching counterfactuals in the latent
-embedding of a Variational Autoencoder (VAE)
-\autocite{joshi2019realistic}; \emph{Wachter} is the baseline approach
-that only penalises the distance to the original sample
-\autocite{wachter2017counterfactual}.
+\citep{pawelczyk2022probabilistically}; \emph{REVISE} addresses the need
+for plausibility by searching counterfactuals in the latent embedding of
+a Variational Autoencoder (VAE) \citep{joshi2019realistic};
+\emph{Wachter} is the baseline approach that only penalises the distance
+to the original sample \citep{wachter2017counterfactual}.
 
@@
 \endlastfoot
-ClaPROAR \autocite{altmeyer2023endogenous} & gradient based & feature &
+ClaPROAR \citep{altmeyer2023endogenous} & gradient based & feature &
 yes \\
-CLUE \autocite{antoran2020getting} & gradient based & latent & yes \\
-DiCE \autocite{mothilal2020explaining} & gradient based & feature &
+CLUE \citep{antoran2020getting} & gradient based & latent & yes \\
+DiCE \citep{mothilal2020explaining} & gradient based & feature & yes \\
+FeatureTweak \citep{tolomei2017interpretable} & tree based & feature &
+no \\
+Gravitational \citep{altmeyer2023endogenous} & gradient based & feature
+& yes \\
+Greedy \citep{schut2021generating} & gradient based & feature & yes \\
+GrowingSpheres \citep{laugel2017inversea} & agnostic & feature & no \\
+PROBE \citep{pawelczyk2022probabilistically} & gradient based & feature
+& no \\
+REVISE \citep{joshi2019realistic} & gradient based & latent & yes \\
+Wachter \citep{wachter2017counterfactual} & gradient based & feature &
 yes \\
-FeatureTweak \autocite{tolomei2017interpretable} & tree based & feature
-& no \\
-Gravitational \autocite{altmeyer2023endogenous} & gradient based &
-feature & yes \\
-Greedy \autocite{schut2021generating} & gradient based & feature &
-yes \\
-GrowingSpheres \autocite{laugel2017inversea} & agnostic & feature &
-no \\
-PROBE \autocite{pawelczyk2022probabilistically} & gradient based &
-feature & no \\
-REVISE \autocite{joshi2019realistic} & gradient based & latent & yes \\
-Wachter \autocite{wachter2017counterfactual} & gradient based & feature
-& yes \\
 \end{longtable}
@@
 \item
-  Adult Census \autocite{becker1996adult}
+  Adult Census \citep{becker1996adult}
 \item
-  California Housing \autocite{pace1997sparse}
+  California Housing \citep{pace1997sparse}
 \item
-  CIFAR10 \autocite{krizhevsky2009learning}
+  CIFAR10 \citep{krizhevsky2009learning}
 \item
-  German Credit \autocite{hoffman1994german}
+  German Credit \citep{hoffman1994german}
 \item
-  Give Me Some Credit \autocite{kaggle2011give}
+  Give Me Some Credit \citep{kaggle2011give}
 \item
-  MNIST \autocite{lecun1998mnist} and Fashion MNIST
-  \autocite{xiao2017fashion}
+  MNIST \citep{lecun1998mnist} and Fashion MNIST \citep{xiao2017fashion}
 \item
-  UCI defaultCredit \autocite{yeh2009comparisons}
+  UCI defaultCredit \citep{yeh2009comparisons}
 \end{itemize}
@@
 \(\mathcal{X}\) and \(\text{diversity}(\cdot)\) is the penalty proposed
-by \textcite{mothilal2020explaining} that induces diverse sets of
+by \citet{mothilal2020explaining} that induces diverse sets of
 counterfactuals. As in Equation~\ref{eq-solution}, \(\ell\) is the loss
@@
 counterfactual from its original value. The API call to compute the
-distance metric defined in \textcite{wachter2017counterfactual}, for
+distance metric defined in \citet{wachter2017counterfactual}, for
 instance, is as simple as \texttt{evaluate(ce;\ measure=distance\_mad)},
@@
 how native support for \href{https://fluxml.ai/}{\texttt{Flux.jl}}
-\autocite{innes2018flux} deep learning models was enabled.\footnote{Flux
+\citep{innes2018flux} deep learning models was enabled.\footnote{Flux
   models are now natively supported by our package and can be
@@
 quantifying the uncertainty around the generated counterfactuals
-\autocite{delaney2021uncertainty}. One idea could be, to use dropout to
+\citep{delaney2021uncertainty}. One idea could be, to use dropout to
 randomly switch features on and off in each iteration. Without dwelling
@@
 The \emph{Give Me Some Credit} dataset is one of the tabular real-world
-datasets that ship with the package \autocite{kaggle2011give}. It can be
+datasets that ship with the package \citep{kaggle2011give}. It can be
 used to train a binary classifier to predict whether a borrower is
@@
 dropout regularization. Using the Gravitational generator
-\autocite{altmeyer2023endogenous} we will generate counterfactuals for
-ten randomly chosen individuals that would be denied credit based on our
+\citep{altmeyer2023endogenous} we will generate counterfactuals for ten
+randomly chosen individuals that would be denied credit based on our
 pre-trained model. Concerning the mutability of features, we only impose
@@
 contains 60,000 training samples of handwritten digits in the form of
-28x28 pixel grey-scale images \autocite{lecun1998mnist}. Each image is
+28x28 pixel grey-scale images \citep{lecun1998mnist}. Each image is
 associated with a label indicating the digit (0-9) that the image
@@
 supervised models built in \texttt{MLJ.jl}, an extensive Machine
-Learning framework for Julia \autocite{blaom2020mlj}. This may also
-involve adding support for regression models as well as additional
+Learning framework for Julia \citep{blaom2020mlj}. This may also involve
+adding support for regression models as well as additional
 non-differentiable models. In terms of counterfactual generators, there
 is a list of recent methodologies that we would like to implement
-including MINT \autocite{karimi2021algorithmic}, ROAR
-\autocite{upadhyay2021robust} and FACE \autocite{poyiadzi2020face}.
+including MINT \citep{karimi2021algorithmic}, ROAR
+\citep{upadhyay2021robust} and FACE \citep{poyiadzi2020face}.
 
@@
 explanations including \emph{Adult}, \emph{Give Me Some Credit} and
-\emph{German Credit} \autocite{karimi2020survey}. There is scope for
-adding data sources that have so far not been explored much in this
-context including additional image datasets as well as audio, natural
-language and time-series data.
+\emph{German Credit} \citep{karimi2020survey}. There is scope for adding
+data sources that have so far not been explored much in this context
+including additional image datasets as well as audio, natural language
+and time-series data.
 
@@
 
-\printbibliography[title=References]
+\renewcommand\refname{References}
+  \bibliography{ref.bib}
 

### 2024-01-07 examples/CounterfactualExplanations.jl/paper.tex
block-headigns

--- 2024-01-07 examples/CounterfactualExplanations.jl/paper.tex
@@
 \setcounter{secnumdepth}{5}
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
 
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 

### 2024-01-07 examples/CounterfactualExplanations.jl/paper.tex
better way to deal with trailing white space after paras

@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
 \usepackage{xcolor}
@@
 \makeatother
-
-% Remove whitespace after paragraphs
-\setlength{\parskip}{0pt}
 
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, sharp corners, interior hidden, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 

### 2024-01-08 examples/CounterfactualExplanations.jl/paper.tex
going to turn the presentation into a website

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, sharp corners, interior hidden, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, sharp corners, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
 

### 2024-01-08 template.qmd
going to turn the presentation into a website

@@
 
-Support for executable listings is only available in the development version of Quarto (version $\geq$ 1.4.0).
-It is currently not clear how to support custom formatting, for example through the existing `jlcode.sty` environment.
+- Support for executable listings is only available in the latest version of Quarto (version $\geq$ 1.4.0).
+- It is currently not clear how to support custom formatting, for example through the existing `jlcode.sty` environment. 
+- Cross-referencing listings does not seem to work in HTML output.
 
@@
 This is the recommended way to create tables because it is compatible with various output formats.
-Tables can be cross-referenced and the table caption will be rendered as a LaTeX table caption.
+Tables can be cross-referenced as `@tbl-1` which renders as @tbl-1.
 
@@
 
-<!-- | Default | Left | Right | Center |
+| Default | Left | Right | Center |
 |---------|:-----|------:|:------:|
@@
 
-: Demonstration of pipe table syntax {#tbl-1} -->
+: Demonstration of pipe table syntax {#tbl-1}
 
@@
 The table caption can be added as a paragraph.
-The following creates @tbl-1 which can be cross-referenced as `@tbl-1`.
+The following creates @tbl-2 which can be cross-referenced as `@tbl-2`.
 
 ```
-::: {#tbl-1}
+::: {#tbl-2}
 
@@
 
-::: {#tbl-1}
+::: {#tbl-2}
 
@@
 :::
+
+# Citations
+
+This `@altmeyer2023endogenous` renders the citation as @altmeyer2023endogenous. The bibliography is defined in `bibliography.bib`.
+
+# References
